---
title: "Noe Leo Final"
author: "Léo-Paul Garelli and Noé Mansuy"
format: pdf
---
Github project : https://github.com/LEOPAUL55/Noe_Leo_final.git

```{r}
#| message: false
here::i_am("Noe_Leo_Final.Rproj")
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(vroom)
library(data.table)
library(readr)
```

# population 2018

Because the 2018 data for incomes is by IRIS code and not by GEO code, we need to ponderate each mean of the 2018 IRIS codes belonging to the same city by their 2018 population in order to find the true values we're looking for.

```{r}
pop2018_IRIS <- read_delim(here("data","pop2018.csv"),
    na = "s") |>
  select(IRIS,P18_POP)
    
```



# Real estate prices

Source : https://www.data.gouv.fr/datasets/indicateurs-immobiliers-par-commune-et-par-annee-prix-et-volumes-sur-la-periode-2014-2024

This is a collection of various variables on residential housing market at the city level between 2014 and 2024. What is interesting for us is the mean price by square meter of the goods that have been sold each year. This was published by Boris Mericksay


```{r}
ppsm_2016 <- read_delim(here("data","price per square meter","dvf2016.csv"),
    na = "s")
  
ppsm_2017 <- read_delim(here("data","price per square meter","dvf2017.csv"),
    na = "s")

ppsm_2018 <- read_delim(here("data","price per square meter","dvf2018.csv"),
    na = "s")

ppsm_2019 <- read_delim(here("data","price per square meter","dvf2019.csv"),
    na = "s")

ppsm_2020 <- read_delim(here("data","price per square meter","dvf2020.csv"),
    na = "s")

ppsm_2021 <- read_delim(here("data","price per square meter","dvf2021.csv"),
    na = "s")
```
# Poverty

sources : 

- 2016, 2017, 2018 : https://www.data.gouv.fr/datasets/revenus-et-pauvrete-des-menages-aux-niveaux-national-et-local-revenus-localises-sociaux-et-fiscaux/

- 2019 : https://www.insee.fr/fr/statistiques/6036902?sommaire=6036904

- 2020 : https://www.insee.fr/fr/statistiques/6692392?sommaire=6692394

- 2021 : https://www.insee.fr/fr/statistiques/7756729?sommaire=7756859

These are collections of indicators of disposable income in France, at the city level, between 2016 and 2021. These are made by the Insee.



```{r}
pov2016 <- read_delim(here("data","poverty","pov2016.csv"),
                      na = "s") 
pov2017 <- read_delim(here("data","poverty","pov2017.csv"),
                      na = "s")
pov2018 <- read_delim(here("data","poverty","pov2018.csv"),
                      na = "s")
pov2019 <- read_delim(here("data","poverty","pov2019.csv"),
                      na = "s")
pov2020 <- read_delim(here("data","poverty","pov2020.csv"),
                      na = "s")
pov2021 <- read_delim(here("data","poverty","pov2021.csv"),
                      na = "s") 
```
    
```{r}
pov2016_1 <- pov2016 |>
  select(
    MED16,
    TP6016,
    D916,
    D116,
    CODGEO,
    LIBGEO)|>
  rename(
    MED = MED16,
    TP60 = TP6016,
    D1 = D116,
    D9 = D916
  )

pov2017_1 <- pov2017 |>
  select(
    MED17,
    TP6017,
    D917,
    D117,
    CODGEO,
    )|>
  rename(
    MED = MED17,
    TP60 = TP6017,
    D1 = D117,
    D9 = D917
  ) 

pov2018_1 <- pov2018 |>
  select(
    DEC_MED18,
    DEC_TP6018,
    DEC_D918,
    DEC_D118,
    IRIS
  )|>
  rename(
    MED = DEC_MED18,
    TP60 = DEC_TP6018,
    D1 = DEC_D118,
    D9 = DEC_D918
  )

pov2019_1 <- pov2019 |>
  select(
    MED19,
    TP6019,
    D919,
    D119,
    CODGEO)|>
  rename(
    MED = MED19,
    TP60 = TP6019,
    D1 = D119,
    D9 = D919
  )

pov2020_1 <- pov2020 |>
  select(
    MED20,
    TP6020,
    D920,
    D120,
    CODGEO)|>
  rename(
    MED = MED20,
    TP60 = TP6020,
    D1 = D120,
    D9 = D920
  )

pov2021_1 <- pov2021 |>
  select(
    GEO,
    `Niveau de vie médian (en euros)`,
    `Taux de pauvreté (en %) au seuil de 60 % de la médiane du niveau de vie`,
    `1er décile du niveau de vie (en euros)`,
    `9e décile du niveau de vie (en euros)`
   ) |>
  rename(
    CODGEO = GEO,
    MED = `Niveau de vie médian (en euros)`,
    TP60 = `Taux de pauvreté (en %) au seuil de 60 % de la médiane du niveau de vie`,
    D1 = `1er décile du niveau de vie (en euros)`,
    D9 = `9e décile du niveau de vie (en euros)`
  )
```
```{r}
names <- pov2016_1 |>
  select(CODGEO, LIBGEO)
```

```{r}

pov2018_1 <- pov2018_1 |>
  left_join(pop2018_IRIS,by = "IRIS")

cols_to_weight <- c("MED", "TP60", "D9", "D1")

result <- pov2018_1 %>%
  group_by() %>%
  summarize(
    # Apply weighted.mean to each column in cols_to_weight
    across(all_of(cols_to_weight), 
           ~ weighted.mean(., P18_POP, na.rm = TRUE),
           .names = "{.col}_weighted"),
    
    # Add total population
    Total_POP = sum(P18_POP, na.rm = TRUE)
  ) %>%
  ungroup()
```


```{r}
result <- pov2018_1 %>%
  mutate(CODGEO = substr(IRIS, 1, 5)) %>%
  
  group_by(CODGEO) %>%
  
  summarize(
    MED = weighted.mean(MED, P18_POP, na.rm = TRUE),
    TP60 = weighted.mean(TP60, P18_POP, na.rm = TRUE),
    D9 = weighted.mean(D9, P18_POP, na.rm = TRUE),
    D1 = weighted.mean(D1, P18_POP, na.rm = TRUE),
    
    P18_POP = sum(P18_POP, na.rm = TRUE),  # Renommer si besoin

    
    .groups = 'drop'
  ) |>
  select(-P18_POP)
  
```


# Delinquency

Source : https://www.data.gouv.fr/datasets/bases-statistiques-communale-departementale-et-regionale-de-la-delinquance-enregistree-par-la-police-et-la-gendarmerie-nationales/

this is a database of all the crimes and offences (except traffic ofences) at the city level in France, published by French Police Force.

```{r}
crime_raw <- read.csv(here("data_criminality.csv"))
```

```{r}
crime_1 <- read_delim(here( "data_criminality.csv"),
                      delim = ";",
                      locale = locale(decimal_mark = ","),
                      col_types = cols(
                        annee = col_integer(), 
                        nombre = col_integer()))

spec(crime_1)
```
```{r}
crime_2 <- crime_1 |>
  select(-est_diffuse, 
         -taux_pour_mille,
         -insee_pop_millesime,
         -insee_log_millesime,
         -insee_log,
         -insee_log_millesime,
         -complement_info_nombre,
         -complement_info_taux)
```
```{r}
crime_3 <- crime_2 |>
  pivot_wider(
    names_from = indicateur,
    values_from = nombre
  )
```
```{r}
crime_4 <- crime_3 |>
  filter(insee_pop>=5000) |>
  mutate(total_crimes = rowSums(across(`Violences physiques intrafamiliales`:`Escroqueries et fraudes aux moyens de paiement`), na.rm = TRUE)) |>
  rename(
    CODGEO = CODGEO_2025
  )
```
```{r}
crime_2016 <- crime_4 |>
  filter(annee == "2016")
```
```{r}
crime_2017 <- crime_4 |>
  filter(annee == "2017")
```
```{r}
crime_2018 <- crime_4 |>
  filter(annee == "2018")
```
```{r}
crime_2019 <- crime_4 |>
  filter(annee == "2019")
```
```{r}
crime_2020 <- crime_4 |>
  filter(annee == "2020")
```
```{r}
crime_2021 <- crime_4 |>
  filter(annee == "2021")
```



## Merging and grouping crimes per year per per city
```{r}
cpov_2016 <- crime_2016|>
  left_join(pov2016_1, by = "CODGEO")

cols_to_sum <- names(cpov_2016)[5:19]

cpov_2016 <- cpov_2016 %>%
  group_by(CODGEO, annee, insee_pop,MED,TP60,D9,D1,LIBGEO) %>%
  summarise(across(all_of(cols_to_sum),~sum(.x, na.rm = TRUE)))
  

```

```{r}
name <- pov2016_1 |>
  select(CODGEO, LIBGEO)

cpov_2017 <- crime_2017|>
  left_join(name, by = "CODGEO")|>
  left_join(pov2017_1, by = "CODGEO")

cpov_2017 <- cpov_2017 %>%
  group_by(CODGEO, annee, insee_pop,MED,TP60,D9,D1,LIBGEO) %>%
  summarise(across(all_of(cols_to_sum),~sum(.x, na.rm = TRUE)))
```
```{r}
cpov_2018 <- crime_2018|>
  left_join(name, by = "CODGEO")|>
  left_join(result, by = "CODGEO")
cpov_2018 <- cpov_2018 %>%
  group_by(CODGEO, annee, insee_pop,MED,TP60,D9,D1,LIBGEO) %>%
  summarise(across(all_of(cols_to_sum),~sum(.x, na.rm = TRUE)))
```
```{r}
cpov_2019 <- crime_2019|>
  left_join(name, by = "CODGEO")|>
  left_join(pov2019_1, by = "CODGEO")
cpov_2019 <- cpov_2019 %>%
  group_by(CODGEO, annee, insee_pop,MED,TP60,D9,D1,LIBGEO) %>%
  summarise(across(all_of(cols_to_sum),~sum(.x, na.rm = TRUE)))
```
```{r}
cpov_2020 <- crime_2020|>
  left_join(name, by = "CODGEO")|>
  left_join(pov2020_1, by = "CODGEO")
cpov_2020 <- cpov_2020 %>%
  group_by(CODGEO, annee, insee_pop,MED,TP60,D9,D1,LIBGEO) %>%
  summarise(across(all_of(cols_to_sum),~sum(.x, na.rm = TRUE)))
```
```{r}
cpov_2021 <- crime_2021|>
  left_join(pov2021_1, by = "CODGEO")
cpov_2021 <- cpov_2021 %>%
  group_by(CODGEO, annee, insee_pop,MED,TP60,D9,D1) %>%
  summarise(across(all_of(cols_to_sum),~sum(.x, na.rm = TRUE)))
```

## Selecting cities present in the 2018 data 
```{r}
cities_of_interest <- unique(result$CODGEO)

cpov_2016 <- cpov_2016 |>
  filter(CODGEO %in% cities_of_interest)
cpov_2017 <- cpov_2017 |>
  filter(CODGEO %in% cities_of_interest)
cpov_2018 <- cpov_2018 |>
  filter(CODGEO %in% cities_of_interest)
cpov_2019 <- cpov_2019 |>
  filter(CODGEO %in% cities_of_interest)
cpov_2020 <- cpov_2020 |>
  filter(CODGEO %in% cities_of_interest)
cpov_2021 <- cpov_2021 |>
  filter(CODGEO %in% cities_of_interest)
```
we now have data on criminality and poverty for each year betweem 2016 and 2021

Now we have to clean and add to the rest the real estate prices data :

```{r}
ppsm_2016 <- ppsm_2016 |> 
  select(
    INSEE_COM,
    Prixm2Moyen
  ) |>
  rename(CODGEO = INSEE_COM)

ppsm_2017 <- ppsm_2017 |> 
  select(
    INSEE_COM,
    Prixm2Moyen
  ) |>
  rename(CODGEO = INSEE_COM)

ppsm_2018 <- ppsm_2018 |> 
  select(
    INSEE_COM,
    Prixm2Moyen
  ) |>
  rename(CODGEO = INSEE_COM)

ppsm_2019 <- ppsm_2019 |> 
  select(
    INSEE_COM,
    Prixm2Moyen
  ) |>
  rename(CODGEO = INSEE_COM)

ppsm_2020 <- ppsm_2020 |> 
  select(
    INSEE_COM,
    Prixm2Moyen
  ) |>
  rename(CODGEO = INSEE_COM)

ppsm_2021 <- ppsm_2021 |> 
  select(
    INSEE_COM,
    Prixm2Moyen
  ) |>
  rename(CODGEO = INSEE_COM)
```
```{r}
data_2016 <- cpov_2016 |>
  left_join(ppsm_2016, by ="CODGEO" )

data_2017 <- cpov_2017 |>
  left_join(ppsm_2017, by ="CODGEO" )

data_2018 <- cpov_2018 |>
  left_join(ppsm_2018, by ="CODGEO" )

data_2019 <- cpov_2019 |>
  left_join(ppsm_2019, by ="CODGEO" )

data_2020 <- cpov_2020 |>
  left_join(ppsm_2020, by ="CODGEO" )

data_2021 <- cpov_2021 |>
  left_join(ppsm_2021, by ="CODGEO" )
```

