---
title: "Noe Leo Final"
author: "Léo-Paul Garelli and Noé Mansuy"
format: html
editor: 
  markdown: 
    wrap: 72
---

Github project : https://github.com/LEOPAUL55/Noe_Leo_final.git

# 1. Research question

Theorized during the 1960 decade, gentrification is an urban phenomenon defined as the appropriation of working-class neighborhoods by the upper class. More formally, gentrification materializes by a rise in the real estate market price, as well as a rise in the average standard linving of a quarter. Indeed, the rehabilitation of houses by upper classes leads to a rise in the real estate price market of the quarter. Disadvantaged classes are forced to leave to worse located areas and are replaced by a wealthier population in their initial neighborhood. This phenomenon creates a conflict within social classes that affects the society in different ways, by the education, the inequalities, the health or the criminality.

This project aims to answer the following question: Does the gentrification process in certain French cities correlates with a decrease in recorded crime in these same municipalities, and is there a displacement effect towards non-gentrified cities? The purpose of the study is to estimate the effect of gentrification on criminality at the scale of France on the short and middle run. The goal is to have a better understanding of this phenomenon to implement adequate public policies. To do so, the study is based on the public data of 899 cities in France between 2016 and 2021.

First, the study try to build a gentrification index that would be able to measure the degree of gentrification of each city on the period, based on the process definition. The idea is to have a better understanding of the span and intensity of the phenomenon in France and to separate cities by degree of gentrification. Then, the main goal is to estimate the impact of gentrification on the evolution of criminality by cities. The gentrification index will enable to estimate the direct and lagged effect of gentrification on the criminality's evolution for each city. Indeed, it will measure the effect of gentrification on the global and by type crime rate's evolution.

# 2. Datasets descriptions

```{r}
#| message: false
here::i_am("Noe_Leo_Final.Rproj")
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(vroom)
library(data.table)
library(stringr)
library(plm)
library(psych)
library(mgcv)
library(viridis)
library(FSA)

# Load pre-filtered datasets
poverty <- readRDS(here("data", "cleaned", "poverty.rds"))
crime <- readRDS(here("data", "cleaned", "crime.rds"))
ppsm <- readRDS(here("data", "cleaned", "ppsm.rds"))
city_names <- readRDS(here("data", "cleaned", "city_names.rds"))

# Recreate data_final
data_final <- poverty |>
  left_join(crime, by = c("CODGEO", "year")) |>
  left_join(ppsm, by = c("CODGEO", "year")) |>
  left_join(city_names, by = "CODGEO") |>
  mutate(
    crime_rate = (total_crimes / insee_pop) * 100
  )

# Verify
cat("✓ Data loaded successfully\n")
cat("  - Cities:", n_distinct(data_final$CODGEO), "\n")
cat("  - Years:", n_distinct(data_final$year), "\n")
cat("  - Observations:", nrow(data_final), "\n")
```

The first step of the project is to ensure the reliability of our findings by selecting adequate and accurate data. We need to create two data sets, with respectively a "level" and an "evolution" data set. The "evolution" one will be created thanks to the "level" one. To implement the method of study, essential informations are needed for describing the economic and criminality situation in every cities on the period. More precisely, the build of the gentrification index is based on economic informations such as the the real estate price market (square meter price), the median income, the poverty rate (defined as 60% of the median income) as well as the first and ninth deciles, for each city per year. For the criminality, the study rely on the number of committed crimes for each city per year and type. Thanks to the informations about the level of population, it is possible to compute the crime rate, global and per type.

## a. Data sources

1.  **Real estate price - Datagouv / Chambre des notaires de Paris** 

The informations about the real estate price per city are available on the France state's site. They are given in the files called "Demande de valeur foncière" for every year, that is from 2016 to 2021. This is a collection of various variables on residential housing market at the city level between 2014 and 2024. What is interesting for us is the mean price by square meter of the goods that have been sold each year. This was published by Boris Mericksay. Unfortunatlely, the informations about Paris at the arrondissement level are missing in those files. These important informations can be found on the Chambre des notaires de Paris' site that gives the trimestrial average square meter price of selling. For simplicity, we only focus on the selling price of appartements in the city from 2016 to 2021 (trimester 4). The available file is a PDF that can be easily convert into a CSV file.

2.  **Average standard of living - Insee** 

The files "poverty" are collections of indicators of disposable income in France, at the city level, between 2016 and 2021. We will only focus on the poverty rate, the median standard of living, the first and the ninth deciles. Poverty data are provided as separate yearly files with heterogeneous structures. Variable names differ across years, and the 2018 data are reported at the IRIS level rather than the commune level. To handle this, we implement a loading strategy that reads each file separately while restricting imports to the variables of interest (years 2016:2021). Poverty indicators are harmonized across years. Non‑numeric values are explicitly converted to missing values during loading, but it won't affect our analysis because this only concerns small cities. The result is a consistent panel of indicators comparable across years.

3.  **Criminality - Datagouv / Ministère de l'intérieur** 

The CSV file "criminality" is a database of all the crimes and offences (except traffic offences) at the city level in France, published by French Police Force, since 2016. These offences were identified following a complaint filed by a victim, a report, a testimony, a flagrant offence, a denunciation, but also on the initiative of the security forces. The total number of crimes per year per city is computed by doing the sum of each type of crimes for every year. Thanks to the informations about the number of inhabitants per city given in the past files, it is is possible to compute the global crime rate and per type. We decided to exclude the crime "drug use (AFD)" and "sexual violence" due to the difference between the Police's data and the real situation, that would bias the analysis.

## b. Informations about the data sets

```{r, echo = FALSE}
info <- function(df) {
  data.frame(file = deparse(substitute(df)), Observations = nrow(df), Variables = ncol(df))
}
```

```{r, echo = FALSE}
info_ppsm <- info(ppsm)
info_poverty <- info(poverty)
info_crime <- info(crime) 
bind_rows(info_ppsm, info_poverty, info_crime) |> knitr::kable()
```

## c. Description of variables

Below is the description of variables that are used in the project.

**CODGEO**: Unique code that corresponds to one city. The first two numbers correspond to a department number.

**LIBGEO**: Name of the city according to the CODGEO.

**year**: year in which the data was collected.

**Prixm2Moyen**: Average price per square meter of the real estate goods sold for the year.

**MED**: Median income for this year.

**TP60**: Poverty rate (set at less than 60% of the median income).

**D1**: 1st income decile (income below which there are 10% of the population).

**D9**: 9th income decile (income below which there are 90% of the population).

**Violences physiques intrafamiliales**: Number of domestic physical violence reported for the year.

**Violences physiques hors cadre familial**: Number of non domestic physical violence reported for the year.

**Vols avec armes**: Number of armed robberies reported for the year.

**Vols sans arme**: Number of non armed violent robberies reported for the year.

**Vols sans violence contre des personnes**: Number of non violent robberies against persons reported for the year.

**Cambriolages de logement**: Number of home invasions reported for the year.

**Vols de véhicules**: Number of vehicule theft reported for the year.

**Vols dans les véhicules**: Number of theft from vehicules reported for the year.

**Vols d'accessoires sur véhicules**: Number of theft of vehicle accessories reported for the year.

**Destructions et dégradations volontaires**: Number of willful damage and vandalism reported for the year.

**Usage de stupéfiants**: Number of reported drug use for the year.

**Trafic de stupéfiants**: Number of drug dealing cases reported for the year.

**Escroqueries et fraudes aux moyens de paiement**: Number of fraud and payment method fraud reported for the year.

**total_crimes**: Total number of crimes per cities reported for this year.

**crime_rate**: Total number of crimes per cities divided by the population number of the year.

**insee_pop**: The city's population.


# Descriptive analysis of the crime rate evolution from 2016 to 2021 - comparison between violent and non violent crimes

To better understand whether gentrification affects interpersonal violence differently from property‑related crimes, we aggregate the 13 crime types into two broad categories:

Violent crimes: Domestic physical violence, non‑domestic physical violence, armed robberies, and violent robberies without weapons.

Non‑violent crimes: Non‑violent thefts, vehicle‑related crimes, burglaries, vandalism, fraud, and drug‑related offences.

While non violent and total crimes decrease following roughly the same path, violent crimes increased from 2016-2021.

```{r}
crime_wide <- data_final |>
  mutate(
    violent_total = rowSums(
      across(c(`Violences physiques intrafamiliales`,
               `Violences physiques hors cadre familial`,
               `Vols avec armes`,
               `Vols violents sans arme`)),
      na.rm = TRUE
    ),
    nonviolent_total = rowSums(
      across(c(`Vols sans violence contre des personnes`,
               `Vols de véhicule`,
               `Vols dans les véhicules`,
               `Vols d'accessoires sur véhicules`,
               `Cambriolages de logement`,
               `Destructions et dégradations volontaires`,
               `Escroqueries et fraudes aux moyens de paiement`,
               `Usage de stupéfiants`,
               `Trafic de stupéfiants`)),
      na.rm = TRUE
    ),
    total_excl_sexual = total_crimes,
    violent_rate = (violent_total / insee_pop) * 100,
    nonviolent_rate = (nonviolent_total / insee_pop) * 100,
    total_rate_excl = (total_excl_sexual / insee_pop) * 100
  ) |>
  group_by(year) |>
  summarise(
    total = median(total_rate_excl, na.rm = TRUE),
    violent = median(violent_rate, na.rm = TRUE),
    nonviolent = median(nonviolent_rate, na.rm = TRUE),
    .groups = "drop"
  )

# Scale factor for secondary axis
scale_factor <- max(crime_wide$nonviolent) / max(crime_wide$violent)

ggplot(crime_wide, aes(x = year)) +
  geom_line(aes(y = nonviolent, color = "Non-Violent"), size = 1.5) +
  geom_point(aes(y = nonviolent, color = "Non-Violent"), size = 4) +
  geom_line(aes(y = total, color = "Total"), size = 1.5) +
  geom_point(aes(y = total, color = "Total"), size = 4) +
  geom_line(aes(y = violent * scale_factor, color = "Violent"), size = 1.5) +
  geom_point(aes(y = violent * scale_factor, color = "Violent"), size = 4) +
  scale_y_continuous(
    name = "Non-Violent & Total (per 100 inh.)",
    sec.axis = sec_axis(~./scale_factor, name = "Violent Crime (per 100 inh.)")
  ) +
  scale_x_continuous(breaks = 2016:2021) +
  scale_color_manual(values = c("Total" = "#2c3e50", 
                                "Violent" = "#e74c3c", 
                                "Non-Violent" = "#3498db")) +
  labs(
    x = "Year",
    color = NULL,
    title = "Crime Rate Evolution in French Cities (2016-2021)",
    subtitle = "Note: Violent crime uses secondary axis (right) | Excluding sexual violence"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.position = "bottom",
    axis.title.y.right = element_text(color = "#e74c3c"),
    axis.text.y.right = element_text(color = "#e74c3c")
  )
```

# Building a gentrification index between first year and last year :

After obtaining clean data sets, we can now analyse it through several aspects. The first step is to build a gentrification index, that is a mathematical formula which enables to compute and compare the degree of gentrification per cities thanks to the available information in the data set (economic description, real estate price). Second, it will be possible to compute descriptive statistics about gentrification to have a better understanding of the process in France through the period. More precisely, to what extent are french cities affected by gentrification and where is the process the most intense on the map. In the same idea, descriptive statistics about crime in France will be computed to also understand this phenomenon and its evolution. Third, we will try to isolate with the data the causal link between the degree of gentrification and the criminality evolution in the cities using an econometric method.

Recalling the definition, gentrification is the process by which working class quarter are appropriated by upper classes, which exclude the initial population and replace it by a wealthier one. In the data, such a process can be observed using the data sets "ppsm" (real estate price) and "poverty" (economic information). Indeed, a gentrified city can be identified through a rise in the real estate price (Prix2Moyen) which will increase the city's standard living and rise the median income (MED) as well as the first and last decile (D1 and D9) and reduce the rate of poor people in the population (fall of TP60).

Using these data, we can create a gentrification index that is implementable to any city using the following formula:

**Gentrification** = (**price_var** * 0.25) + (**MED_var** * 0.20) + (**D9_var** * 0.10) + (**D1_var** * 0.35) + (**TP60_var** * 0.10)

Gentrification is not about the level of the variables, but their variation. Thus we need to compute the variation in percentage of each variable to implement the formula. Moreover, we justify the weight given to each variable thanks to the literature. According to the initial definition (Ruth Glass, 1964) combined with the work of Neil Smith and the Rent-Gap Theory (1979), gentrification mainly materialized by the displacement of the working class as well as the rise in the real estate price, that justifies the heavy weight on the first decile (D1_var) and the square meter price (price_var). To make the index meaningful, we standardize it using the z-score standardization and center it around 0 (if the index is equal to 0, the city is as gentrify as the national mean). Finally, we exclude the city that have a median income greater than 60% of the national one, considering they are already composed of upper classes, so it cannot be gentrified.


Identifying areas that could gentrify :

```{r}
#Taking the earliest year of data for each demographic area (CODGEO)
baseline <- data_final |>
  group_by(CODGEO) |>
  filter(year == min(year)) |>
  #capturing the median income as the starting point
  ungroup() |>
  select(CODGEO, baseline_med = MED)

#Making sure only areas below the "60% of median income" threshold are considered "gentrifiable" (by definition, only lower income areas can gentrify)
percentile_cutoff <- quantile(baseline$baseline_med, 0.60, na.rm = TRUE)
```

Calculating changes over time :

```{r}
gentrifiable_areas <- baseline |>
  filter(baseline_med < percentile_cutoff) |>
  pull(CODGEO)

# Calculating percentage changes from first to last year
evolution <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>  
  group_by(CODGEO) |>
  summarise(
    price_var   = ((last(Prixm2Moyen) / first(Prixm2Moyen)) - 1) * 100,
    med_inc_var = (last(MED) / first(MED) - 1) * 100,
    d9_var      = (last(D9) / first(D9) - 1) * 100,
    d1_var      = (last(D1) / first(D1) - 1) * 100,
    pov_var     = last(TP60) - first(TP60),
    .groups = "drop"
  ) |>
  left_join(city_names, by = "CODGEO")

# Flipping signs for interpretation : making increases in low incomes and decrease in poverty show as positive values: this aligns indicators so higher values = more gentrification
evolution <- evolution |>
  mutate(
    pov_var = -pov_var
  )
```

Standardizing the variables :

```{r}
# Converting all the variables to z-score so it makes the different metrics comparable despite different scales 
evolution_std <- evolution |>
  mutate(across(
    c(price_var, med_inc_var, d9_var, d1_var, pov_var),
    scale
  ))
```

Creating a weighted gentrification index

```{r}
evolution_std <- evolution_std |>
  mutate(
    gent_index = 
      (price_var * 0.25) +   # Emphasizes housing price
      (med_inc_var * 0.2) +  #Median income
      (d9_var * 0.1) +       # Top income
      (d1_var * 0.35) +      #Bottom income
      (pov_var * 0.1)        # Poverty reduction
  )


# Sorting areas by gentrification intensity
evolution_std <- evolution_std |>
  arrange(desc(gent_index))
evolution_std_A <- evolution_std

# Displaying most gentrified areas
evolution_std_A |>
  head(10) |>
  select(CODGEO, LIBGEO, gent_index)

# Checking the validity of the index using Cronbach's alpha 
library(psych)
psych::alpha(evolution_std |> select(price_var, med_inc_var, d9_var, d1_var, pov_var), check.keys = TRUE)

```

-\> The Cronbach's alpha test tells us that the indicator we used is
valid and reliable, and that all indicators contribute meaningfully

# Cross sectional analysis

After constructing the gentrification index for all eligible French cities (those with baseline median income below the 60th percentile of the national distribution), we proceed to examine the relationship between gentrification intensity and changes in crime rates over the 2016–2021 period. The analysis is cross-sectional: for each city, we compute the percentage change in crime rate between the first and last year of observation, and correlate this change with the standardized gentrification index.

To facilitate interpretation, we categorize cities into three groups based on their gentrification index:

Low gentrification: bottom 25% of the index distribution

Medium gentrification: middle 50%

High gentrification: top 25%

These categories allow us to compare how crime evolved differently across areas experiencing varying degrees of socioeconomic transformation.
```{r}
#| include: false

evolution_std_A_groups <- evolution_std_A |>
  mutate(
    gent_category = case_when(
      gent_index > quantile(gent_index, 0.75, na.rm = TRUE) ~ "High",
      gent_index > quantile(gent_index, 0.25, na.rm = TRUE) ~ "Medium",
      TRUE ~ "Low"
    ),
    gent_category = factor(gent_category, levels = c("Low", "Medium", "High"))
  )

table(evolution_std_A_groups$gent_category)

crime_rate_change_continuous <- data_final |>
  filter(year %in% c(2016, 2021)) |>
  select(CODGEO, year, total_crimes, insee_pop) |>
  pivot_wider(
    names_from = year,
    values_from = c(total_crimes, insee_pop),
    names_sep = "_"
  ) |>
  mutate(
    crime_rate_2016 = total_crimes_2016 / insee_pop_2016,
    crime_rate_2021 = total_crimes_2021 / insee_pop_2021,
    crime_rate_change = (crime_rate_2021 - crime_rate_2016) / crime_rate_2016
  ) |>
  left_join(
    evolution_std_A_groups |> select(CODGEO, gent_index, gent_category),
    by = "CODGEO"
  ) |>
  filter(!is.na(gent_category))

```

## Visual exploration: Scatter plot

The scatter plot below displays the relationship between the gentrification index (horizontal axis) and the percentage change in crime rate (vertical axis) for all 899 cities in our sample.

```{r}
library(viridis)

ggplot(crime_rate_change_continuous,
       aes(x = gent_index, y = crime_rate_change)) +
  geom_point(aes(color = gent_index), 
             alpha = 0.6, size = 2.5) +
  scale_color_viridis_c(option = "plasma", name = "Gentrification\nIntensity") +
  geom_smooth(method = "lm", color = "#FF6B6B", size = 1.2, 
              fill = "#FF6B6B", alpha = 0.15) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.3, color = "gray40") +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.3, color = "gray40") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_cartesian(ylim = c(-0.5, 0.5)) +
  labs(
    x = "Gentrification Index (standardized)",
    y = "Crime Rate Change (2016-2021)",
    title = "Small Link Between Gentrification and Crime Changes",
    subtitle = "Analysis of initially lower-income areas (bottom 60% baseline income)",
    caption = "Each point represents one geographic area | Red line shows linear trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0),
    plot.subtitle = element_text(size = 12, color = "gray40", hjust = 0),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )
```
Visually, a slight positive trend emerges: cities with higher gentrification indices tend to experience smaller reductions (or even increases) in crime rates compared to less gentrified cities. The cloud of points is dispersed, suggesting considerable heterogeneity. However, running a pearson correlation test shows us that there is no linear relationship between gentrification intensity (continuous index) and crime rate change : 

```{r}
# Tests wether there's a linear relationship between gentrification intensity and crime change 
cor.test(
  crime_rate_change_continuous$gent_index,
  crime_rate_change_continuous$crime_rate_change,
  use = "complete.obs"
)
```




## Comparing crime rate changes across gentrification categories

To better quantify these differences, we compute the mean and median crime rate changes for each gentrification category, along with 95% confidence intervals.

```{r}
#| include: false

evolution_std_A_groups <- evolution_std_A |>
  mutate(
    gent_category = case_when(
      gent_index > quantile(gent_index, 0.75, na.rm = TRUE) ~ "High",
      gent_index > quantile(gent_index, 0.25, na.rm = TRUE) ~ "Medium",
      TRUE ~ "Low"
    ),
    gent_category = factor(gent_category, levels = c("Low", "Medium", "High"))
  )

```

### Plotting the crime rate change among three arbitrary gentrification levels :

```{r}
# Calculate confidence intervals
ci_stats <- crime_rate_change_continuous |>
  group_by(gent_category) |>
  summarise(
    mean = mean(crime_rate_change, na.rm = TRUE),
    ci_lower = mean - 1.96 * sd(crime_rate_change, na.rm = TRUE) / sqrt(n()),
    ci_upper = mean + 1.96 * sd(crime_rate_change, na.rm = TRUE) / sqrt(n()),
    median = median(crime_rate_change, na.rm = TRUE)
  )

ggplot(ci_stats |> filter(gent_category %in% c("Low", "High", "Medium")),
       aes(x = gent_category, y = mean, color = gent_category)) +
  # CI lines
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0, size = 2, alpha = 0.4) +
  # Mean points
  geom_point(size = 8) +
  # Median points
  geom_point(aes(y = median), shape = 1, size = 8, stroke = 1.5) +
  scale_color_manual(values = c("Low" = "#3498db", "Medium" = "#f39c12", "High" = "#e74c3c")) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.3) +
  coord_cartesian(ylim = c(-0.1, 0.1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    x = NULL,
    y = "Crime Rate Change (2016–2021)",
    title = "Mean Crime Rate Changes Across Gentrification Levels",
    subtitle = "Solid dots = mean | Hollow circles = median | Lines = 95% CI",
    
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    axis.text.x = element_text(size = 14, face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```
All three categories experienced a reduction in crime rates between 2016 
and 2021, consistent with national crime trends in France during this period. 
While the magnitude of reduction appears to vary visually across gentrification 
levels, an ANOVA test reveals these differences are not statistically significant:

Tests whether crime rate changes differ significantly across the three gentrification levels :
```{r}
# Tests whether crime rate changes differ significantly across the three gentrification levels
anova_result <- aov(crime_rate_change ~ gent_category, data = crime_rate_change_continuous)
summary(anova_result)

```
While total crime rates show only a  diminution for all gentrification levels, different crime categories may respond differently to socioeconomic changes. We therefore disaggregate crime into 13 specific types and examine how each evolved across gentrification intensities between 2016 and 2021.

For each city, we compute the population‑adjusted change in crime rates for each of the 13 crime categories, thus accounting for both changes in crime counts and changes in population size, providing a cleaner measure of crime‑rate evolution.
```{r}

# Reshaping crime data per type
crime_change_type_continuous <- data_final |>
  pivot_longer(
    cols = c(
      `Violences physiques intrafamiliales`,
      `Violences physiques hors cadre familial`,
      `Vols avec armes`,
      `Vols violents sans arme`,
      `Vols sans violence contre des personnes`,
      `Vols de véhicule`,
      `Vols dans les véhicules`,
      `Vols d'accessoires sur véhicules`,
      `Cambriolages de logement`,
      `Destructions et dégradations volontaires`,
      `Escroqueries et fraudes aux moyens de paiement`,
      `Usage de stupéfiants`,
      `Trafic de stupéfiants`
    ),
    names_to = "crime_type",
    values_to = "crime_count"
  ) |>
  filter(year %in% c(2016, 2021)) |>
  select(CODGEO, year, crime_type, crime_count, insee_pop) |>
  
  # Creating year-over-year comparison
  pivot_wider(
    names_from = year,
    values_from = c(crime_count, insee_pop),
    names_sep = "_"
  ) |>
  # Calculating poopulation-adjusted change
  mutate(
    pop_adj_rate_change = if_else(
      crime_count_2016 > 0 & insee_pop_2016 > 0,
      ((crime_count_2021 / crime_count_2016) /
       (insee_pop_2021 / insee_pop_2016)) - 1,
      NA_real_
    )
  ) |>
  # Merging with gentrification levels
  left_join(
    evolution_std_A_groups |> select(CODGEO, gent_index, gent_category),
    by = "CODGEO"
  ) |>
  filter(!is.na(gent_category))

# Summary by crime type and gentrification level
crime_change_type_summary <- crime_change_type_continuous |>
  filter(is.finite(pop_adj_rate_change)) |>
  group_by(crime_type, gent_category) |>
  summarise(
    median_change = median(pop_adj_rate_change, na.rm = TRUE),
    n_cities = n(),
    .groups = "drop"
  )

# Plot
ggplot(crime_change_type_summary,
       aes(x = crime_type, y = median_change, fill = gent_category)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("Low" = "steelblue", "Medium" = "orange", "High" = "red")) +
  labs(
    x = "Type of Crime",
    y = "Population-Adjusted Crime Change (2016–2021)",
    fill = "Gentrification\nIntensity",
    title = "Crime Change by Type and Gentrification Intensity"
  ) +
  theme_minimal()
```
For each individual crime type, the direction of change (increase or decrease) is consistent across all three gentrification categories. However, the magnitude of change varies considerably depending on the gentrification intensity for certain crime types. 

Kruskal-Wallis tests were conducted for each of the 13 crime categories 
to assess whether population-adjusted crime rate changes differ 
significantly across gentrification levels.

Only one crime type showed a statistically significant difference: 
**domestic physical violence** (Violences physiques intrafamiliales), 
with χ²(2) = 19.09, p < 0.001. This indicates that gentrification 
intensity significantly predicts how domestic violence evolved (positively)
between 2016 and 2021.

```{r}
# Run Kruskal-Wallis for each crime type
kruskal_by_crime_type <- crime_change_type_continuous |>
  filter(is.finite(pop_adj_rate_change)) |>
  group_by(crime_type) |>
  summarise(
    chi_squared = kruskal.test(pop_adj_rate_change ~ gent_category)$statistic,
    p_value = kruskal.test(pop_adj_rate_change ~ gent_category)$p.value,
    n = n(),
    .groups = "drop"
  ) |>
  mutate(
    significant = ifelse(p_value < 0.05, "Yes", "No")
  ) |>
  arrange(p_value)

# Display results
knitr::kable(kruskal_by_crime_type, 
             digits = 4,
             col.names = c("Crime Type", "Chi-squared", "P-value", "N", "Significant"))
```



## Violent vs Non-violent breakdown

Using the violent and non-violent crime categories defined in the 
descriptive analysis section, we now examine whether gentrification 
affects these two broad categories differently.

```{r}

# Creating crime categories 
crime_change_type_cat <- crime_change_type_continuous |>
  mutate(
    crime_category = case_when(
      crime_type %in% c(
       
        "Violences physiques intrafamiliales",
        "Violences physiques hors cadre familial",
        "Vols avec armes",
        "Vols violents sans arme"
      ) ~ "Violent crimes",
      crime_type %in% c(
        "Vols sans violence contre des personnes",
        "Vols de véhicule",
        "Vols dans les véhicules",
        "Vols d'accessoires sur véhicules",
        "Cambriolages de logement",
        "Destructions et dégradations volontaires",
        "Escroqueries et fraudes aux moyens de paiement",
        "Usage de stupéfiants",
        "Trafic de stupéfiants"
      ) ~ "Non violent crimes",
      TRUE ~ NA_character_
    )
  )
```


```{r}

# Calculating the median population adjusted change for each combination of crime category and gentrification level
crime_cat_summary <- crime_change_type_cat |>
  filter(is.finite(pop_adj_rate_change), !is.na(crime_category)) |>
  group_by(crime_category, gent_category) |>
  summarise(
    median_change = median(pop_adj_rate_change, na.rm = TRUE),
    n_obs = n(),
    .groups = "drop"
  )
```


```{r}
# Plot
ggplot(crime_cat_summary,
       aes(x = crime_category, y = median_change, fill = gent_category)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Low" = "steelblue", "Medium" = "orange", "High" = "red")) +
  labs(
    x = "Crime Category",
    y = "Population-Adjusted Crime Rate Change (2016–2021)",
    fill = "Gentrification\nIntensity",
    title = "Violent vs Non-Violent Crime Rate Change\nby Gentrification Intensity"
  ) +
  theme_minimal()

# Table
knitr::kable(
  crime_cat_summary,
  digits = 3,
  col.names = c("Crime Category", "Gentrification Level", "Median Change in %", "N Observations")
)
```
The results reveal a clear gradient pattern: higher gentrification is associated with smaller crime changes in both directions.

Violent crimes increased everywhere, but the increase was much smaller in high-gentrification cities (+5%) compared to low-gentrification cities (+24%).

Non-violent crimes decreased most in low-gentrification cities (-17%), moderately in medium-gentrification cities, and least in high-gentrification cities.

This suggests that gentrification has a stabilizing effect on crime rates, dampening both the rise in violent crime and the decline in non-violent crime. Gentrifying areas experience more moderate changes compared to non-gentrifying areas.

To verify whether these differences are statistically significant, we conduct pairwise Wilcoxon tests with Bonferroni correction:

```{r}
# Prepare data for violent crimes
violent_data <- crime_change_type_cat |>
  filter(crime_category == "Violent crimes", is.finite(pop_adj_rate_change)) |>
  group_by(CODGEO, gent_category) |>
  summarise(median_change = median(pop_adj_rate_change), .groups = "drop")

# Prepare data for non-violent crimes
nonviolent_data <- crime_change_type_cat |>
  filter(crime_category == "Non violent crimes", is.finite(pop_adj_rate_change)) |>
  group_by(CODGEO, gent_category) |>
  summarise(median_change = median(pop_adj_rate_change), .groups = "drop")

# Pairwise Wilcoxon tests
violent_pairwise <- pairwise.wilcox.test(
  violent_data$median_change,
  violent_data$gent_category,
  p.adjust.method = "bonferroni"
)

nonviolent_pairwise <- pairwise.wilcox.test(
  nonviolent_data$median_change,
  nonviolent_data$gent_category,
  p.adjust.method = "bonferroni"
)

# Create combined table
tibble(
  Comparison = c("Low vs Medium", "Low vs High", "Medium vs High"),
  `Violent crimes (p-value)` = c(
    violent_pairwise$p.value["Medium", "Low"],
    violent_pairwise$p.value["High", "Low"],
    violent_pairwise$p.value["High", "Medium"]
  ),
  `Non-violent crimes (p-value)` = c(
    nonviolent_pairwise$p.value["Medium", "Low"],
    nonviolent_pairwise$p.value["High", "Low"],
    nonviolent_pairwise$p.value["High", "Medium"]
  )
) |>
  knitr::kable(digits = 4, caption = "Pairwise Wilcoxon Tests (Bonferroni corrected)")
```
For violent crimes, high-gentrification cities differ significantly from both low-gentrification (p < 0.001) and medium-gentrification cities (p < 0.001), confirming that the dampening effect of gentrification on violent crime is statistically robust.

For non-violent crimes, no pairwise comparisons reach statistical significance (all p > 0.05), suggesting that the observed differences may be due to chance.


# Panel Analysis

The cross-sectional analysis examined the relationship between gentrification and crime changes over the entire 2016-2021 period. However, this approach cannot capture the temporal dynamics of how gentrification affects crime over time. Does gentrification in one year affect crime the following year? Do effects accumulate or reverse over multiple years?

To address these questions, we construct a year-over-year gentrification index that measures annual changes in gentrification intensity for each city. This allows us to test whether lagged gentrification predicts subsequent crime changes using panel data methods.

## Year over year gentrification index

Instead of measuring gentrification as the total change from 2016 to 2021, we now compute annual growth rates for each component of the gentrification index. This creates a time-varying measure that captures the intensity of gentrification in each city-year.

```{r}
gent_yearly <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    # Year-over-year growth rates (log differences)
    price_g = log(Prixm2Moyen) - log(dplyr::lag(Prixm2Moyen)),
    med_g   = log(MED) - log(dplyr::lag(MED)),
    d9_g    = log(D9) - log(dplyr::lag(D9)),
    d1_g    = -(log(D1) - log(dplyr::lag(D1))),
    pov_g   = -(TP60 - dplyr::lag(TP60))
  ) |>
  ungroup() |>
  filter(!is.na(price_g)) |>
  mutate(
    # Standardizing within the dataset
    price_g_s = as.numeric(scale(price_g)),
    med_g_s   = as.numeric(scale(med_g)),
    d9_g_s    = as.numeric(scale(d9_g)),
    d1_g_s    = as.numeric(scale(d1_g)),
    pov_g_s   = as.numeric(scale(pov_g)),
    
    # Weighted index (SAME WEIGHTS as first-to-last year index)
    gent_index = (price_g_s * 0.25) +
                 (med_g_s * 0.20) +
                 (d9_g_s * 0.10) +
                 (d1_g_s * 0.35) +
                 (pov_g_s * 0.10)
  )
# Calculate year-over-year crime rate change
crime_yearly <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    crime_rate = total_crimes / insee_pop,
    crime_change = log(crime_rate) - log(dplyr::lag(crime_rate))
  ) |>
  ungroup() |>
  select(CODGEO, year, crime_rate, crime_change, total_crimes, insee_pop)

# Create the panel by joining gentrification index with crime data
crime_panel <- gent_yearly |>
  select(CODGEO, year, gent_index) |>
  left_join(crime_yearly, by = c("CODGEO", "year")) |>
  filter(!is.na(crime_change), !is.na(gent_index))

```

```{r}
cat("Panel observations:", nrow(crime_panel), "\n")
cat("Cities:", n_distinct(crime_panel$CODGEO), "\n")
cat("Years:", unique(crime_panel$year), "\n")
summary(crime_panel$gent_index)
```
The resulting panel contains 2,688 city-year observations across 538 cities from 2017 to 2021. The gentrification index is centered around zero (mean ≈ 0) with values ranging from -2.29 to +2.20, indicating substantial variation in annual gentrification intensity across cities and years.


## Lagged effects

### On total crimes

To test whether gentrification in one year affects crime the following year, we estimate a fixed effects panel model. This approach controls for both time-invariant city characteristics (city fixed effects) and common yearly shocks (year fixed effects).

```{r}
library(plm)
panel_data <- pdata.frame(crime_panel, index = c("CODGEO", "year"))
model <- plm(crime_change ~ lag(gent_index), 
             data = panel_data, 
             model = "within",  # Fixed effects
             effect = "twoways") # Both area and year FE
summary(model)
```
The results show a small but statistically significant negative effect (β = -0.015, p = 0.03). Cities that experienced more gentrification in one year tended to have slightly smaller increases (or larger decreases) in total crime the following year. However, the effect size is modest: the model explains only 0.3% of the variance in crime changes (R² = 0.003).

### On violent and non violent crimes

Given that our cross-sectional analysis found significant effects specifically for violent crimes, we now examine whether this pattern holds in the panel setting. We test whether gentrification in year t predicts changes in violent and non-violent crime in year t+1.

```{r}
# Creating violent crimes panel
violent_crimes_panel <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  mutate(
    violent_total = rowSums(
      across(c(`Violences physiques intrafamiliales`,
               `Violences physiques hors cadre familial`,
               `Vols avec armes`,
               `Vols violents sans arme`)),
      na.rm = TRUE
    ),
    violent_rate = violent_total / insee_pop
  ) |>
  left_join(
    gent_yearly |> select(CODGEO, year, gent_index),
    by = c("CODGEO", "year")
  ) |>
  group_by(CODGEO) |>
  mutate(
    violent_change = log(violent_rate + 0.001) - log(dplyr::lag(violent_rate) + 0.001),
    gent_index_lag1 = dplyr::lag(gent_index, 1)
  ) |>
  ungroup()

# Creating non-violent crimes panel
nonviolent_crimes_panel <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  mutate(
    nonviolent_total = rowSums(
      across(c(`Vols sans violence contre des personnes`,
               `Vols de véhicule`,
               `Vols dans les véhicules`,
               `Vols d'accessoires sur véhicules`,
               `Cambriolages de logement`,
               `Destructions et dégradations volontaires`,
               `Escroqueries et fraudes aux moyens de paiement`,
               `Usage de stupéfiants`,
               `Trafic de stupéfiants`)),
      na.rm = TRUE
    ),
    nonviolent_rate = nonviolent_total / insee_pop
  ) |>
  left_join(
    gent_yearly |> select(CODGEO, year, gent_index),
    by = c("CODGEO", "year")
  ) |>
  group_by(CODGEO) |>
  mutate(
    nonviolent_change = log(nonviolent_rate + 0.001) - log(dplyr::lag(nonviolent_rate) + 0.001),
    gent_index_lag1 = dplyr::lag(gent_index, 1)
  ) |>
  ungroup()

# Filtering for analysis
violent_for_analysis <- violent_crimes_panel |>
  filter(!is.na(violent_change) & !is.na(gent_index_lag1))

nonviolent_for_analysis <- nonviolent_crimes_panel |>
  filter(!is.na(nonviolent_change) & !is.na(gent_index_lag1))

# Correlation tests
cor_violent <- cor.test(
  violent_for_analysis$gent_index_lag1,
  violent_for_analysis$violent_change,
  use = "complete.obs"
)

cor_nonviolent <- cor.test(
  nonviolent_for_analysis$gent_index_lag1,
  nonviolent_for_analysis$nonviolent_change,
  use = "complete.obs"
)

# Summary table
tibble(
  `Crime Type` = c("Violent crimes", "Non-violent crimes"),
  Correlation = c(cor_violent$estimate, cor_nonviolent$estimate),
  `P-value` = c(cor_violent$p.value, cor_nonviolent$p.value),
  N = c(nrow(violent_for_analysis), nrow(nonviolent_for_analysis))
) |>
  knitr::kable(digits = 4, caption = "Lagged Gentrification Effect on Crime (1-year lag)")
```
Violent crimes show a significant negative correlation (r = -0.102, p < 0.001): cities that experienced more gentrification in one year tended to have smaller increases (or larger decreases) in violent crime the following year.

Non-violent crimes show a stronger and also significant negative correlation (r = -0.178, p < 0.001): gentrification is associated with subsequent reductions in non-violent crime as well.

Interestingly, the lagged effect is stronger for non-violent crimes than for violent crimes in the panel setting, which contrasts with our cross-sectional findings. This suggests that while gentrification's immediate (same-period) effect may differ by crime type, its delayed effect tends to reduce both categories of crime.

## Multiple lags analysis

```{r}
# Creating multiple lags for violent crimes
violent_with_lags <- violent_crimes_panel |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    gent_index_lag2 = dplyr::lag(gent_index, 2),
    gent_index_lag3 = dplyr::lag(gent_index, 3)
  ) |>
  ungroup()

# Creating multiple lags for non-violent crimes
nonviolent_with_lags <- nonviolent_crimes_panel |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    gent_index_lag2 = dplyr::lag(gent_index, 2),
    gent_index_lag3 = dplyr::lag(gent_index, 3)
  ) |>
  ungroup()

# Creating multiple lags for total crime
total_with_lags <- crime_panel |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    gent_index_lag2 = dplyr::lag(gent_index, 2),
    gent_index_lag3 = dplyr::lag(gent_index, 3)
  ) |>
  ungroup()

# Function to compute correlations for each lag
compute_lag_cors <- function(data, change_var) {
  lag1_data <- data |> filter(!is.na({{change_var}}) & !is.na(gent_index_lag1))
  lag2_data <- data |> filter(!is.na({{change_var}}) & !is.na(gent_index_lag2))
  lag3_data <- data |> filter(!is.na({{change_var}}) & !is.na(gent_index_lag3))
  
  cor_lag1 <- cor.test(lag1_data$gent_index_lag1, pull(lag1_data, {{change_var}}))
  cor_lag2 <- cor.test(lag2_data$gent_index_lag2, pull(lag2_data, {{change_var}}))
  cor_lag3 <- cor.test(lag3_data$gent_index_lag3, pull(lag3_data, {{change_var}}))
  
  tibble(
    Lag = c(1, 2, 3),
    Correlation = c(cor_lag1$estimate, cor_lag2$estimate, cor_lag3$estimate),
    P_value = c(cor_lag1$p.value, cor_lag2$p.value, cor_lag3$p.value),
    N = c(nrow(lag1_data), nrow(lag2_data), nrow(lag3_data))
  )
}

# Compute for each crime type
total_cors <- compute_lag_cors(total_with_lags, crime_change) |> mutate(Crime_Type = "Total")
violent_cors <- compute_lag_cors(violent_with_lags, violent_change) |> mutate(Crime_Type = "Violent")
nonviolent_cors <- compute_lag_cors(nonviolent_with_lags, nonviolent_change) |> mutate(Crime_Type = "Non-violent")

# Combine and reshape to wide format
all_lag_results <- bind_rows(total_cors, violent_cors, nonviolent_cors) |>
  mutate(
    P_value = ifelse(P_value < 0.001, "< 0.001", round(P_value, 4)),
    Result = paste0(round(Correlation, 3), " (", P_value, ")")
  ) |>
  select(Crime_Type, Lag, Result) |>
  pivot_wider(
    names_from = Lag,
    values_from = Result,
    names_prefix = "Lag "
  )

knitr::kable(all_lag_results, 
             col.names = c("Crime Type", "1 year", "2 years", "3 years"),
             caption = "Gentrification Effects on Crime at Different Lags (Correlation, P-value)")
```

This oscillating pattern suggests that gentrification's effect on crime is not permanent. The initial crime reduction is partially reversed in the short-medium term, possibly reflecting displacement effects, adjustment dynamics, or statistical mean reversion. Notably, total and non-violent crime follow nearly identical patterns, while violent crime shows the same directional effects but with smaller magnitudes.

To ensure robustness, we estimate fixed effects panel models controlling for year effects and mean reversion in crime rates. Year fixed effects are particularly important given that 2020 saw an unusual decrease in crime likely due to COVID-19 lockdowns, which could otherwise confound our results. The models include gentrification at 1-year and 2-year lags; we do not extend to 3 lags as our 6-year panel (2016-2021) would leave too few observations for reliable estimation.
```{r}
# Prepare violent crimes data
panel_violent_robust <- violent_with_lags |>
  group_by(CODGEO) |>
  mutate(violent_change_lag1 = dplyr::lag(violent_change, 1)) |>
  ungroup() |>
  filter(!is.na(violent_change) & 
         !is.na(gent_index_lag1) & 
         !is.na(gent_index_lag2) & 
         !is.na(violent_change_lag1)) |>
  pdata.frame(index = c("CODGEO", "year"))

# Prepare non-violent crimes data
nonviolent_with_lags <- nonviolent_crimes_panel |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    gent_index_lag2 = dplyr::lag(gent_index, 2),
    nonviolent_change_lag1 = dplyr::lag(nonviolent_change, 1)
  ) |>
  ungroup()

panel_nonviolent_robust <- nonviolent_with_lags |>
  filter(!is.na(nonviolent_change) & 
         !is.na(gent_index_lag1) & 
         !is.na(gent_index_lag2) & 
         !is.na(nonviolent_change_lag1)) |>
  pdata.frame(index = c("CODGEO", "year"))

# Prepare total crime data
total_with_lags <- crime_panel |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    gent_index_lag2 = dplyr::lag(gent_index, 2),
    crime_change_lag1 = dplyr::lag(crime_change, 1)
  ) |>
  ungroup()

panel_total_robust <- total_with_lags |>
  filter(!is.na(crime_change) & 
         !is.na(gent_index_lag1) & 
         !is.na(gent_index_lag2) & 
         !is.na(crime_change_lag1)) |>
  pdata.frame(index = c("CODGEO", "year"))

# Model for violent crimes
model_violent <- plm(
  violent_change ~ gent_index_lag1 + gent_index_lag2 + violent_change_lag1,
  data = panel_violent_robust,
  model = "within",
  effect = "twoways"
)

# Model for non-violent crimes
model_nonviolent <- plm(
  nonviolent_change ~ gent_index_lag1 + gent_index_lag2 + nonviolent_change_lag1,
  data = panel_nonviolent_robust,
  model = "within",
  effect = "twoways"
)

# Model for total crime
model_total <- plm(
  crime_change ~ gent_index_lag1 + gent_index_lag2 + crime_change_lag1,
  data = panel_total_robust,
  model = "within",
  effect = "twoways"
)

# Extract coefficients and p-values
extract_results <- function(model, crime_type) {
  coefs <- summary(model)$coefficients
  tibble(
    `Crime Type` = crime_type,
    Variable = c("Gentrification (Lag 1)", "Gentrification (Lag 2)", "Mean reversion"),
    Estimate = coefs[1:3, "Estimate"],
    `P-value` = coefs[1:3, "Pr(>|t|)"]
  )
}

# Combine results
results_table <- bind_rows(
  extract_results(model_total, "Total"),
  extract_results(model_violent, "Violent"),
  extract_results(model_nonviolent, "Non-violent")
) |>
  mutate(
    `P-value` = ifelse(`P-value` < 0.001, "< 0.001", round(`P-value`, 3)),
    Significant = case_when(
      as.numeric(ifelse(`P-value` == "< 0.001", 0, `P-value`)) < 0.05 ~ "Yes",
      as.numeric(ifelse(`P-value` == "< 0.001", 0, `P-value`)) < 0.10 ~ "Borderline",
      TRUE ~ "No"
    )
  )

knitr::kable(results_table, 
             digits = 3,
             caption = "Fixed Effects Panel Models with Year Controls and Mean Reversion")
```
The results confirm that gentrification has a modest but significant crime-reducing effect at the 1-year lag, particularly for total crime (β = -0.019, p = 0.022) and violent crime (β = -0.027, p = 0.020). The effect on non-violent crime is borderline significant (p = 0.052).

Interestingly, only violent crime shows a potential cumulative effect at the 2-year lag (β = -0.022, p = 0.054), suggesting that gentrification's impact on interpersonal violence may persist longer than its effect on property crime.

All models show strong mean reversion (β ≈ -0.50), indicating that crime rates naturally oscillate over time. After controlling for this, gentrification explains a small but statistically significant portion of crime changes.

# Robustness checks

## Granger Causality

But maybe the causality is in the other direction ? does gentrification cause violent crime change or does violent crime causes gentrification ? Let's use Granger Causality
```{r}
library(plm)

# Preparing balanced panel
granger_data <- violent_crimes_panel |>
  filter(!is.na(violent_change) & !is.na(gent_index) & !is.na(gent_index_lag1)) |>
  group_by(CODGEO) |>
  filter(n() >= 3) |>  # Need at least 3 observations per area
  ungroup()

granger_panel <- pdata.frame(granger_data, index = c("CODGEO", "year"))

# Model 1: Crime change predicted by lagged gentrification
model_gent_causes_crime <- plm(
  violent_change ~ gent_index_lag1,
  data = granger_panel,
  model = "within",
  effect = "individual"
)

# Model 2: Reverse - does crime cause gentrification?
granger_data2 <- granger_data |>
  group_by(CODGEO) |>
  mutate(violent_change_lag1 = dplyr::lag(violent_change, 1)) |>
  ungroup() |>
  filter(!is.na(violent_change_lag1))

granger_panel2 <- pdata.frame(granger_data2, index = c("CODGEO", "year"))

model_crime_causes_gent <- plm(
  gent_index ~ violent_change_lag1,
  data = granger_panel2,
  model = "within",
  effect = "individual"
)

print("=== Does Gentrification Cause Crime? ===")
summary(model_gent_causes_crime)

print("=== Does Crime Cause Gentrification? ===")
summary(model_crime_causes_gent)
```
-> Granger causality tests reveal bidirectional relationships: gentrification 
significantly reduces subsequent violent crime (β = -0.036, p < 0.001), while 
violent crime reductions also predict increased gentrification (β = -0.137, 
p = 0.041). This suggests a self-reinforcing cycle where safer neighborhoods 
attract investment that further reduces crime.

Visualization 1: Reverse causality
```{r}
ggplot(granger_data2 |> filter(!is.na(violent_change_lag1) & !is.na(gent_index)), 
       aes(x = violent_change_lag1, y = gent_index)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Previous Year's Violent Crime Change",
    y = "This Year's Gentrification Index",
    title = "Crime Drives Gentrification Patterns",
    subtitle = "Areas with declining crime attract more gentrification (r = -0.19, p < 0.001)"
  ) +
  theme_minimal()
```

## Heterogeneity by city size

Comparing the effects of gentrification on  richer vs poorer areas :
```{r}
# Separating areas by income
interaction_data <- violent_crimes_panel |>
  left_join(baseline |> select(CODGEO, baseline_med), by = "CODGEO") |>
  mutate(
    income_tercile = ntile(baseline_med, 3),
    income_group = case_when(
      income_tercile == 1 ~ "Low Income",
      income_tercile == 2 ~ "Medium Income",
      income_tercile == 3 ~ "High Income"
    )
  ) |>
  filter(!is.na(violent_change) & !is.na(gent_index_lag1))

# Test by income group
interaction_results <- interaction_data |>
  group_by(income_group) |>
  summarise(
    correlation = cor(gent_index_lag1, violent_change, use = "complete.obs"),
    p_value = cor.test(gent_index_lag1, violent_change)$p.value,
    n = n()
  )

print(interaction_results)

# Regression with interaction
interaction_model <- lm(violent_change ~ gent_index_lag1 * income_group, 
                        data = interaction_data)
summary(interaction_model)
```
-> There is NO statistically significant difference in how gentrification affects violent crime across income groups.



Visualization 1: Interaction effect
```{r}
ggplot(interaction_data |> filter(!is.na(income_group)), 
       aes(x = gent_index_lag1, y = violent_change, color = income_group)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("Low Income" = "#e74c3c", 
                                 "Medium Income" = "#f39c12", 
                                 "High Income" = "#27ae60")) +
  labs(
    x = "Previous Year's Gentrification Index",
    y = "Violent Crime Change",
    color = "Initial Income Level",
    title = "Gentrification's Effect on Crime Varies by Initial Income",
    subtitle = "Gentrification REDUCES crime in low-income areas, but INCREASES it in high-income areas"
  ) +
  theme_minimal()
```
## Spatial spillover effects

Do neighboring areas' gentrification affect local crime?
```{r}
#Identifying department from CODGEO
spatial_data <- violent_crimes_panel |>
  mutate(department = substr(CODGEO, 1, 2)) |>
  group_by(department, year) |>
  mutate(
    # Average gentrification in the same department (excluding self)
    n_in_dept = n(),
    sum_gent_dept = sum(gent_index, na.rm = TRUE),
    neighbor_gent = (sum_gent_dept - gent_index) / (n_in_dept - 1)
  ) |>
  ungroup() |>
  group_by(CODGEO) |>
  mutate(
    neighbor_gent_lag1 = dplyr::lag(neighbor_gent, 1)
  ) |>
  ungroup() |>
  filter(!is.na(violent_change) & !is.na(neighbor_gent_lag1))

# Does neighboring gentrification affect local crime?
cor_neighbor <- cor.test(spatial_data$neighbor_gent_lag1, spatial_data$violent_change)
print("=== Neighboring Gentrification Effect ===")
print(paste("Correlation:", round(cor_neighbor$estimate, 4)))
print(paste("P-value:", round(cor_neighbor$p.value, 4)))

# Combined model
spatial_model <- lm(violent_change ~ gent_index_lag1 + neighbor_gent_lag1, 
                    data = spatial_data)
summary(spatial_model)
```

```{r}
# First, create total crime change variable
total_crime_panel <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  mutate(
    total_crime_rate = total_crimes / insee_pop
  ) |>
  left_join(
    gent_yearly |> select(CODGEO, year, gent_index),
    by = c("CODGEO", "year")
  ) |>
  group_by(CODGEO) |>
  mutate(
    total_crime_change = log(total_crime_rate + 0.0001) - 
                         log(dplyr::lag(total_crime_rate) + 0.0001),
    gent_index_lag1 = dplyr::lag(gent_index, 1)
  ) |>
  ungroup()

# Now run the same spatial analysis for total crimes
spatial_total <- total_crime_panel |>
  mutate(department = substr(CODGEO, 1, 2)) |>
  group_by(department, year) |>
  mutate(
    n_in_dept = n(),
    sum_gent_dept = sum(gent_index, na.rm = TRUE),
    neighbor_gent = (sum_gent_dept - gent_index) / (n_in_dept - 1)
  ) |>
  ungroup() |>
  group_by(CODGEO) |>
  mutate(neighbor_gent_lag1 = dplyr::lag(neighbor_gent, 1)) |>
  ungroup() |>
  filter(!is.na(total_crime_change) & !is.na(neighbor_gent_lag1))

# Correlation
cor_total_neighbor <- cor.test(spatial_total$neighbor_gent_lag1, 
                               spatial_total$total_crime_change)
print("=== TOTAL CRIMES: Neighboring Gentrification Effect ===")
print(paste("Correlation:", round(cor_total_neighbor$estimate, 4)))
print(paste("P-value:", round(cor_total_neighbor$p.value, 4)))

# Combined model
spatial_model_total <- lm(total_crime_change ~ gent_index_lag1 + neighbor_gent_lag1, 
                          data = spatial_total)
summary(spatial_model_total)
```
```{r}
# First calculate non-violent crimes
nonviolent_panel <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  mutate(
    nonviolent_total = rowSums(
      across(c(
        `Vols sans violence contre des personnes`,
        `Vols de véhicule`,
        `Vols dans les véhicules`,
        `Vols d'accessoires sur véhicules`,
        `Cambriolages de logement`,
        `Destructions et dégradations volontaires`,
        `Escroqueries et fraudes aux moyens de paiement`,
        `Usage de stupéfiants`,
        `Trafic de stupéfiants`
      )),
      na.rm = TRUE
    ),
    nonviolent_rate = nonviolent_total / insee_pop
  ) |>
  left_join(
    gent_yearly |> select(CODGEO, year, gent_index),
    by = c("CODGEO", "year")
  ) |>
  group_by(CODGEO) |>
  mutate(
    nonviolent_change = log(nonviolent_rate + 0.0001) - 
                       log(dplyr::lag(nonviolent_rate) + 0.0001),
    gent_index_lag1 = dplyr::lag(gent_index, 1)
  ) |>
  ungroup()

# Spatial analysis for non-violent crimes
spatial_nonviolent <- nonviolent_panel |>
  mutate(department = substr(CODGEO, 1, 2)) |>
  group_by(department, year) |>
  mutate(
    n_in_dept = n(),
    sum_gent_dept = sum(gent_index, na.rm = TRUE),
    neighbor_gent = (sum_gent_dept - gent_index) / (n_in_dept - 1)
  ) |>
  ungroup() |>
  group_by(CODGEO) |>
  mutate(neighbor_gent_lag1 = dplyr::lag(neighbor_gent, 1)) |>
  ungroup() |>
  filter(!is.na(nonviolent_change) & !is.na(neighbor_gent_lag1))

# Correlation
cor_nonviolent_neighbor <- cor.test(spatial_nonviolent$neighbor_gent_lag1, 
                                    spatial_nonviolent$nonviolent_change)
print("=== NON-VIOLENT CRIMES: Neighboring Gentrification Effect ===")
print(paste("Correlation:", round(cor_nonviolent_neighbor$estimate, 4)))
print(paste("P-value:", round(cor_nonviolent_neighbor$p.value, 4)))

# Combined model
spatial_model_nonviolent <- lm(nonviolent_change ~ gent_index_lag1 + neighbor_gent_lag1, 
                               data = spatial_nonviolent)
summary(spatial_model_nonviolent)
```


-> Spillover effect for total crimes, not violent crimes and violent crimes : Cities surrounded by slowly-gentrifying (or non-gentrifying) neighbors tend to experience MORE crime growth.

## Heterogeneity by city size

Do we observe different effects depending on the size of the city ? (urban/rural comparison based on population)
```{r}
# Creating urban/rural indicator based on population
pop_data <- data_final |>
  group_by(CODGEO) |>
  summarise(mean_pop = mean(insee_pop, na.rm = TRUE))

urban_rural_data <- violent_crimes_panel |>
  left_join(pop_data, by = "CODGEO") |>
  mutate(
    area_type = case_when(
      mean_pop > 50000 ~ "Large City",
      mean_pop > 10000 ~ "Medium City",
      TRUE ~ "Small City"
    )
  ) |>
  filter(!is.na(violent_change) & !is.na(gent_index_lag1))

# Testing by area type
urban_rural_results <- urban_rural_data |>
  group_by(area_type) |>
  summarise(
    correlation = cor(gent_index_lag1, violent_change, use = "complete.obs"),
    p_value = cor.test(gent_index_lag1, violent_change)$p.value,
    n = n()
  )

print(urban_rural_results)

# Visualization
ggplot(urban_rural_data, aes(x = gent_index_lag1, y = violent_change, color = area_type)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~area_type) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Previous Year's Gentrification Index",
    y = "Violent Crime Change",
    title = "Gentrification-Crime Relationship by City Size"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
In larger and medium cities there is a significantly negative correlation between the lagged gentrification and violent crime change rate, but not in small cities.

# Conclusion
## Summary of key findings
## Limitations

The main limitation of the analyse is the lack of data both in terms of year and city. Whole areas are missing, like Alsace or Moselle, as well as important cities at the arrondissement level (Marseille, Lyon). Moreover, using the Police data to study criminality can be biased. It can reflect the police activity or the number of police officers rather than the true number of crimes and offences. 

## Policy implications

## References

Ruth Glass (1964): https://www.tandfonline.com/doi/full/10.1080/03058034.2024.2425593 Neil Smith (1979): https://miguelangelmartinez.net/IMG/pdf/1979_smith_gentrification_japa.pdf