---
title: "Noe Leo Final"
author: "Léo-Paul Garelli and Noé Mansuy"
format: html
---
Github project : https://github.com/LEOPAUL55/Noe_Leo_final.git

```{r}
#| message: false
here::i_am("Noe_Leo_Final.Rproj")
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(vroom)
library(data.table)
library(readr)
```

# Data Loading and Cleaning
## population 2018

Source : https://www.insee.fr/fr/statistiques/5650720

This is the 2018 population by IRIS code. Made by the INSEE.

Because the 2018 income data is available by IRIS code rather than by GEO code, we need to weight each IRIS-level mean by its 2018 population in order to compute the correct city-level values. Since the 2018 population data at the IRIS level is available for only 994 of the cities included in the other years, we will have to restrict our analysis to these 994 cities. 

```{r}
pop2018_IRIS <- read_delim(here("data","pop2018.csv"),
    na = "s") |>
  select(IRIS,P18_POP)
    
```



## Real estate prices

Source : https://www.data.gouv.fr/datasets/indicateurs-immobiliers-par-commune-et-par-annee-prix-et-volumes-sur-la-periode-2014-2024

This is a collection of various variables on residential housing market at the city level between 2014 and 2024. What is interesting for us is the mean price by square meter of the goods that have been sold each year. This was published by Boris Mericksay


```{r}
ppsm_2016 <- read_delim(here("data","price per square meter","dvf2016.csv"),
    na = "s")
  
ppsm_2017 <- read_delim(here("data","price per square meter","dvf2017.csv"),
    na = "s")

ppsm_2018 <- read_delim(here("data","price per square meter","dvf2018.csv"),
    na = "s")

ppsm_2019 <- read_delim(here("data","price per square meter","dvf2019.csv"),
    na = "s")

ppsm_2020 <- read_delim(here("data","price per square meter","dvf2020.csv"),
    na = "s")

ppsm_2021 <- read_delim(here("data","price per square meter","dvf2021.csv"),
    na = "s")
```
## Poverty

sources : 

- 2016, 2017, 2018 : https://www.data.gouv.fr/datasets/revenus-et-pauvrete-des-menages-aux-niveaux-national-et-local-revenus-localises-sociaux-et-fiscaux/

- 2019 : https://www.insee.fr/fr/statistiques/6036902?sommaire=6036904

- 2020 : https://www.insee.fr/fr/statistiques/6692392?sommaire=6692394

- 2021 : https://www.insee.fr/fr/statistiques/7756729?sommaire=7756859

These are collections of indicators of disposable income in France, at the city level, between 2016 and 2021. These are made by the Insee.



```{r}
pov2016 <- read_delim(here("data","poverty","pov2016.csv"),
                      na = "s") 
pov2017 <- read_delim(here("data","poverty","pov2017.csv"),
                      na = "s")
pov2018 <- read_delim(here("data","poverty","pov2018.csv"),
                      na = "s")
pov2019 <- read_delim(here("data","poverty","pov2019.csv"),
                      na = "s")
pov2020 <- read_delim(here("data","poverty","pov2020.csv"),
                      na = "s")
pov2021 <- read_delim(here("data","poverty","pov2021.csv"),
                      na = "s") 
```
    
```{r}
pov2016_1 <- pov2016 |>
  select(
    MED16,
    TP6016,
    D916,
    D116,
    CODGEO,
    LIBGEO)|>
  rename(
    MED = MED16,
    TP60 = TP6016,
    D1 = D116,
    D9 = D916
  )

pov2017_1 <- pov2017 |>
  select(
    MED17,
    TP6017,
    D917,
    D117,
    CODGEO,
    )|>
  rename(
    MED = MED17,
    TP60 = TP6017,
    D1 = D117,
    D9 = D917
  ) 

pov2018_1 <- pov2018 |>
  select(
    DEC_MED18,
    DEC_TP6018,
    DEC_D918,
    DEC_D118,
    IRIS
  )|>
  rename(
    MED = DEC_MED18,
    TP60 = DEC_TP6018,
    D1 = DEC_D118,
    D9 = DEC_D918
  )

pov2019_1 <- pov2019 |>
  select(
    MED19,
    TP6019,
    D919,
    D119,
    CODGEO)|>
  rename(
    MED = MED19,
    TP60 = TP6019,
    D1 = D119,
    D9 = D919
  )

pov2020_1 <- pov2020 |>
  select(
    MED20,
    TP6020,
    D920,
    D120,
    CODGEO)|>
  rename(
    MED = MED20,
    TP60 = TP6020,
    D1 = D120,
    D9 = D920
  )

pov2021_1 <- pov2021 |>
  select(
    GEO,
    `Niveau de vie médian (en euros)`,
    `Taux de pauvreté (en %) au seuil de 60 % de la médiane du niveau de vie`,
    `1er décile du niveau de vie (en euros)`,
    `9e décile du niveau de vie (en euros)`
   ) |>
  rename(
    CODGEO = GEO,
    MED = `Niveau de vie médian (en euros)`,
    TP60 = `Taux de pauvreté (en %) au seuil de 60 % de la médiane du niveau de vie`,
    D1 = `1er décile du niveau de vie (en euros)`,
    D9 = `9e décile du niveau de vie (en euros)`
  )
```
```{r}
names <- pov2016_1 |>
  select(CODGEO, LIBGEO)
```

Computing poverty rates for 2018 :

```{r}
pov2018_1 <- pov2018_1 |>
  left_join(pop2018_IRIS,by = "IRIS")

cols_to_weight <- c("MED", "TP60", "D9", "D1")

result <- pov2018_1 %>%
  group_by() %>%
  summarize(
    # Apply weighted.mean to each column in cols_to_weight
    across(all_of(cols_to_weight), 
           ~ weighted.mean(., P18_POP, na.rm = TRUE),
           .names = "{.col}_weighted"),
    
    # Add total population
    Total_POP = sum(P18_POP, na.rm = TRUE)
  ) %>%
  ungroup()
```


```{r}
result <- pov2018_1 %>%
  mutate(CODGEO = substr(IRIS, 1, 5)) %>%
  
  group_by(CODGEO) %>%
  
  summarize(
    MED = weighted.mean(MED, P18_POP, na.rm = TRUE),
    TP60 = weighted.mean(TP60, P18_POP, na.rm = TRUE),
    D9 = weighted.mean(D9, P18_POP, na.rm = TRUE),
    D1 = weighted.mean(D1, P18_POP, na.rm = TRUE),
    
    P18_POP = sum(P18_POP, na.rm = TRUE), 

    
    .groups = 'drop'
  ) |>
  select(-P18_POP)
  
```


## Delinquency

Source : https://www.data.gouv.fr/datasets/bases-statistiques-communale-departementale-et-regionale-de-la-delinquance-enregistree-par-la-police-et-la-gendarmerie-nationales/

this is a database of all the crimes and offences (except traffic ofences) at the city level in France, published by French Police Force.

```{r}
crime_raw <- read.csv(here("data_criminality.csv"))
```

```{r}
crime_1 <- read_delim(here( "data_criminality.csv"),
                      delim = ";",
                      locale = locale(decimal_mark = ","),
                      col_types = cols(
                        annee = col_integer(), 
                        nombre = col_integer()))

spec(crime_1)
```
```{r}
crime_2 <- crime_1 |>
  select(-est_diffuse, 
         -taux_pour_mille,
         -insee_pop_millesime,
         -insee_log_millesime,
         -insee_log,
         -insee_log_millesime,
         -complement_info_nombre,
         -complement_info_taux)
```
```{r}
crime_3 <- crime_2 |>
  pivot_wider(
    names_from = indicateur,
    values_from = nombre
  )
```
```{r}
crime_4 <- crime_3 |>
  filter(insee_pop>=5000) |>
  mutate(total_crimes = rowSums(across(`Violences physiques intrafamiliales`:`Escroqueries et fraudes aux moyens de paiement`), na.rm = TRUE)) |>
  rename(
    CODGEO = CODGEO_2025
  )
```
```{r}
crime_2016 <- crime_4 |>
  filter(annee == "2016")
```
```{r}
crime_2017 <- crime_4 |>
  filter(annee == "2017")
```
```{r}
crime_2018 <- crime_4 |>
  filter(annee == "2018")
```
```{r}
crime_2019 <- crime_4 |>
  filter(annee == "2019")
```
```{r}
crime_2020 <- crime_4 |>
  filter(annee == "2020")
```
```{r}
crime_2021 <- crime_4 |>
  filter(annee == "2021")
```



## Merging and grouping crimes per year per per city
```{r}
cpov_2016 <- crime_2016|>
  left_join(pov2016_1, by = "CODGEO")

cols_to_sum <- names(cpov_2016)[5:19]

cpov_2016 <- cpov_2016 %>%
  group_by(CODGEO, annee, insee_pop,MED,TP60,D9,D1,LIBGEO) %>%
  summarise(across(all_of(cols_to_sum),~sum(.x, na.rm = TRUE)))
  

```

```{r}
name <- pov2016_1 |>
  select(CODGEO, LIBGEO)

cpov_2017 <- crime_2017|>
  left_join(name, by = "CODGEO")|>
  left_join(pov2017_1, by = "CODGEO")

cpov_2017 <- cpov_2017 %>%
  group_by(CODGEO, annee, insee_pop,MED,TP60,D9,D1,LIBGEO) %>%
  summarise(across(all_of(cols_to_sum),~sum(.x, na.rm = TRUE)))
```
```{r}
cpov_2018 <- crime_2018|>
  left_join(name, by = "CODGEO")|>
  left_join(result, by = "CODGEO")
cpov_2018 <- cpov_2018 %>%
  group_by(CODGEO, annee, insee_pop,MED,TP60,D9,D1,LIBGEO) %>%
  summarise(across(all_of(cols_to_sum),~sum(.x, na.rm = TRUE)))
```
```{r}
cpov_2019 <- crime_2019|>
  left_join(name, by = "CODGEO")|>
  left_join(pov2019_1, by = "CODGEO")
cpov_2019 <- cpov_2019 %>%
  group_by(CODGEO, annee, insee_pop,MED,TP60,D9,D1,LIBGEO) %>%
  summarise(across(all_of(cols_to_sum),~sum(.x, na.rm = TRUE)))
```
```{r}
cpov_2020 <- crime_2020|>
  left_join(name, by = "CODGEO")|>
  left_join(pov2020_1, by = "CODGEO")
cpov_2020 <- cpov_2020 %>%
  group_by(CODGEO, annee, insee_pop,MED,TP60,D9,D1,LIBGEO) %>%
  summarise(across(all_of(cols_to_sum),~sum(.x, na.rm = TRUE)))
```
```{r}
cpov_2021 <- crime_2021|>
  left_join(name, by = "CODGEO")|>
  left_join(pov2021_1, by = "CODGEO")
cpov_2021 <- cpov_2021 %>%
  group_by(CODGEO, annee, insee_pop,MED,TP60,D9,D1,LIBGEO) %>%
  summarise(across(all_of(cols_to_sum),~sum(.x, na.rm = TRUE)))
```

## Selecting cities present in the 2018 data 
```{r}
cities_of_interest <- unique(result$CODGEO)

cpov_2016 <- cpov_2016 |>
  filter(CODGEO %in% cities_of_interest)
cpov_2017 <- cpov_2017 |>
  filter(CODGEO %in% cities_of_interest)
cpov_2018 <- cpov_2018 |>
  filter(CODGEO %in% cities_of_interest)
cpov_2019 <- cpov_2019 |>
  filter(CODGEO %in% cities_of_interest)
cpov_2020 <- cpov_2020 |>
  filter(CODGEO %in% cities_of_interest)
cpov_2021 <- cpov_2021 |>
  filter(CODGEO %in% cities_of_interest)
```
we now have data on criminality and poverty for each year betweem 2016 and 2021

Now we have to clean and add to the rest the real estate prices data :

```{r}
ppsm_2016 <- ppsm_2016 |> 
  select(
    INSEE_COM,
    Prixm2Moyen
  ) |>
  rename(CODGEO = INSEE_COM)

ppsm_2017 <- ppsm_2017 |> 
  select(
    INSEE_COM,
    Prixm2Moyen
  ) |>
  rename(CODGEO = INSEE_COM)

ppsm_2018 <- ppsm_2018 |> 
  select(
    INSEE_COM,
    Prixm2Moyen
  ) |>
  rename(CODGEO = INSEE_COM)

ppsm_2019 <- ppsm_2019 |> 
  select(
    INSEE_COM,
    Prixm2Moyen
  ) |>
  rename(CODGEO = INSEE_COM)

ppsm_2020 <- ppsm_2020 |> 
  select(
    INSEE_COM,
    Prixm2Moyen
  ) |>
  rename(CODGEO = INSEE_COM)

ppsm_2021 <- ppsm_2021 |> 
  select(
    INSEE_COM,
    Prixm2Moyen
  ) |>
  rename(CODGEO = INSEE_COM)
```
```{r}
data_2016 <- cpov_2016 |>
  left_join(ppsm_2016, by ="CODGEO" )

data_2017 <- cpov_2017 |>
  left_join(ppsm_2017, by ="CODGEO" )

data_2018 <- cpov_2018 |>
  left_join(ppsm_2018, by ="CODGEO" )

data_2019 <- cpov_2019 |>
  left_join(ppsm_2019, by ="CODGEO" )

data_2020 <- cpov_2020 |>
  left_join(ppsm_2020, by ="CODGEO" )

data_2021 <- cpov_2021 |>
  left_join(ppsm_2021, by ="CODGEO" )
```

-> We have our clean data for all years between 2016 and 2021

# Cleaned data tables : summary and most important variables

First, we computed the 2018 poverty rates by city, weighting the poverty rates of each geocode by the population of the corresponding IRIS codes. We then kept only the variables of interest for each dataset (poverty rates, crime statistics, and housing prices) and merged them. Now, for each year (2016–2021), we have a dataframe that includes the number of crimes for each crime category (summing crimes corresponding to different units of account when they refer anyways all to a unic crime) for each city with complete poverty rate observations across all years. The dataset also includes the mean price per square meter of real estate transactions in each city for the corresponding year.

```{r}
descriptions <- c(
  "Unique code that corresponds to one city. The first two numbers correspond to a department",
  "The concerned year",
  "The city's population",
  "Median income for this year",
  "Poverty rate (set at less than 60% of the median income)",
  "9th income decile",
  "1st income decile",
  "The city's name",
  "Number of domestic physical violence reported for the year",
  "Number of non domestic physical violence reported for the year",
  "Number of sexual violence reported for the year",
  "Number of armed robberies reported for the year",
  "Number of non armed violent robberies reported for the year",
  "Number of non violent robberies against persons reported for the year",
  "Number of home invasions reported for the year",
  "Number of vehicle theft reported for the year",
  "Number of theft from vehicles reported for the year",
  "Number of theft of vehicle accessories reported for the year",
  "Number of willful damage and vandalism reported for the year",
  "Number of reported drug use for the year",
  "Number of illegal drug use sanctioned by fixed fine reported for the year",
  "Number of drug dealing cases reported for the year",
  "Number of fraud and payment method fraud reported for the year",
  "Price per square meter of the real estate goods sold for the year"
)

summary <- data.table(
  Variable = names(data_2016),
  Type = sapply(data_2016, class),
  Description = descriptions
)

knitr::kable(summary)

dataset_info <- data.table(
  Info = c("Number of rows", "Number of columns"),
  Value = c(nrow(data_2016), ncol(data_2016)) 
)

knitr::kable(dataset_info)

```
(Similar for each year)

# Research question : 

Does the acceleration of the gentrification process in certain French cities with more than 5000 inhabitants (2016-2021) correlate with a decrease in recorded crime in these same municipalities, and is there a displacement effect towards non-gentrified cities?

The first goal will be to make a gentrification index per city with a few variables: square meter price growth, median wage growth, poverty rate growth.

Then we will select the cities with the highest gentrification index (say top 20%) and compare it to the rest of the cities : do we observe a declination in terms of delinquency between the top 20% and the others ? If yes, do we observe a spatial displacement of crime to non-gentrifying areas ?

