---
title: "Noe Leo Final"
author: "Léo-Paul Garelli and Noé Mansuy"
format: html
---

Github project : https://github.com/LEOPAUL55/Noe_Leo_final.git

```{r}
#| message: false
here::i_am("Noe_Leo_Final.Rproj")
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(vroom)
library(data.table)
library(readr)
library(stringr)
```

# Data Loading and Cleaning

## population 2018

Source : https://www.insee.fr/fr/statistiques/5650720

This is the 2018 population by IRIS code. Made by the INSEE.

Because the 2018 income data is available by IRIS code rather than by GEO code, we need to weight each IRIS-level mean by its 2018 population in order to compute the correct city-level values. Since the 2018 population data at the IRIS level is available for only 994 of the cities included in the other years, we will have to restrict our analysis to these 994 cities.

```{r}
pop2018_IRIS <- read_delim(here("data","pop2018.csv"),
    na = "s") |>
  select(IRIS,P18_POP)
    
```

## Real estate prices

Source : https://www.data.gouv.fr/datasets/indicateurs-immobiliers-par-commune-et-par-annee-prix-et-volumes-sur-la-periode-2014-2024

This is a collection of various variables on residential housing market at the city level between 2014 and 2024. What is interesting for us is the mean price by square meter of the goods that have been sold each year. This was published by Boris Mericksay

Real‑estate price data are provided as separate yearly CSV files. To improve efficiency and avoid redundant code, we use vroom’s multi‑file loading functionality to read all files simultaneously. The year is extracted from the file name, and city identifiers are harmonized to allow consistent merging with other datasets.

```{r}
ppsm_files <- list.files(
  here("data", "price per square meter"),
  pattern = "\\.csv$",
  full.names = TRUE
)

ppsm <- vroom(
  ppsm_files,
  delim = ",",
  col_select = c(INSEE_COM, Prixm2Moyen),
  id = "source"
) |>
  mutate(
    year = str_extract(source, "\\d{4}") |> as.integer(),
    CODGEO = INSEE_COM
  ) |>
  select(CODGEO, year, Prixm2Moyen)
```

## Poverty

sources :

-   2016, 2017, 2018 : https://www.data.gouv.fr/datasets/revenus-et-pauvrete-des-menages-aux-niveaux-national-et-local-revenus-localises-sociaux-et-fiscaux/

-   2019 : https://www.insee.fr/fr/statistiques/6036902?sommaire=6036904

-   2020 : https://www.insee.fr/fr/statistiques/6692392?sommaire=6692394

-   2021 : https://www.insee.fr/fr/statistiques/7756729?sommaire=7756859

These are collections of indicators of disposable income in France, at the city level, between 2016 and 2021. These are made by the Insee.

We will only focus on the poverty rate, the median standard of living, the first and the ninth deciles.

Poverty data are provided as separate yearly files with heterogeneous structures. Variable names differ across years, and the 2018 data are reported at the IRIS level rather than the commune level. To handle this, we implement a loading strategy that reads each file separately while restricting imports to the variables of interest (years 2016:2021). Poverty indicators are harmonized across years. Non‑numeric values are explicitly converted to missing values during loading, but it won't affect our analysis because this only concerns small cities. The result is a consistent panel of indicators comparable across years.



```{r}
poverty_files <- list.files(
  here("data", "poverty"),
  pattern = "\\.csv$",
  full.names = TRUE
)

load_poverty <- function(file) {

  year <- as.integer(stringr::str_extract(file, "\\d{4}"))

  if (year %in% c(2016, 2017, 2019, 2020)) {

    out <- vroom(
      file,
      delim = ";",
      col_select = any_of(c(
        "CODGEO",
        "MED16", "MED17", "MED19", "MED20",
        "TP6016", "TP6017", "TP6019", "TP6020",
        "D116", "D117", "D119", "D120",
        "D916", "D917", "D919", "D920"
      ))
    ) |>
      transmute(
        CODGEO,
        MED  = Reduce(coalesce, pick(any_of(c("MED16", "MED17", "MED19", "MED20")))),
        TP60 = Reduce(coalesce, pick(any_of(c("TP6016", "TP6017", "TP6019", "TP6020")))),
        D1   = Reduce(coalesce, pick(any_of(c("D116", "D117", "D119", "D120")))),
        D9   = Reduce(coalesce, pick(any_of(c("D916", "D917", "D919", "D920")))),
        year
      )

  } else if (year == 2018) {

    out <- vroom(
      file,
      delim = ";",
      col_select = any_of(c(
        "IRIS",
        "DEC_MED18",
        "DEC_TP6018",
        "DEC_D118",
        "DEC_D918"
      ))
    ) |>
      transmute(
        IRIS,
        MED  = DEC_MED18,
        TP60 = DEC_TP6018,
        D1   = DEC_D118,
        D9   = DEC_D918,
        year
      )

  } else if (year == 2021) {

    out <- vroom(
      file,
      delim = ";",
      col_select = any_of(c(
        "GEO",
        "Niveau de vie médian (en euros)",
        "Taux de pauvreté (en %) au seuil de 60 % de la médiane du niveau de vie",
        "1er décile du niveau de vie (en euros)",
        "9e décile du niveau de vie (en euros)"
      ))
    ) |>
      transmute(
        CODGEO = GEO,
        MED  = `Niveau de vie médian (en euros)`,
        TP60 = `Taux de pauvreté (en %) au seuil de 60 % de la médiane du niveau de vie`,
        D1   = `1er décile du niveau de vie (en euros)`,
        D9   = `9e décile du niveau de vie (en euros)`,
        year
      )
  }

  
  out |>
    mutate(across(
      c(MED, TP60, D1, D9),
      ~ suppressWarnings(as.numeric(.x))
    ))
}
poverty_raw <- bind_rows(
  lapply(poverty_files, load_poverty)
)



```
Aggregating 2018 IRIS-level poverty data to the city level and combining weighted 2018 data with commune-level data for other years :

```{r}
poverty_2018_city <- poverty_raw |>
  filter(year == 2018) |>
  left_join(pop2018_IRIS, by = "IRIS") |>
  mutate(CODGEO = substr(IRIS, 1, 5)) |>
  group_by(CODGEO) |>
  summarise(
    MED = weighted.mean(MED, P18_POP, na.rm = TRUE),
    TP60 = weighted.mean(TP60, P18_POP, na.rm = TRUE),
    D1 = weighted.mean(D1, P18_POP, na.rm = TRUE),
    D9 = weighted.mean(D9, P18_POP, na.rm = TRUE),
    year = 2018,
    .groups = "drop"
  )

poverty_city <- bind_rows(
  poverty_raw |> filter(year != 2018),
  poverty_2018_city
)
```

## Delinquency

Source: https://www.data.gouv.fr/datasets/bases-statistiques-communale-departementale-et-regionale-de-la-delinquance-enregistree-par-la-police-et-la-gendarmerie-nationales

Here is the loading of the database for the delinquency. This is a database of all the crimes and offences (except traffic offences) at the city level in France, published by French Police Force, since 2016. 

The csv format of the file creates problems of fromat that must be solved (types of column, delimiotation...).
```{r}
crime <- vroom(
  here("data_criminality.csv"),
  delim = ";",
  locale = locale(decimal_mark = ","),
  col_select = c(
    annee,
    CODGEO_2025,
    indicateur,
    nombre,
    insee_pop
  ),
  col_types = cols(
    annee = col_integer(),
    nombre = col_integer(),
    insee_pop = col_integer()
  )
)
```

The type of crimes must be reshape to long format (in rows) to wide format (in columns).
```{r}
crime_wide <- crime |>
  pivot_wider(
    names_from = indicateur,
    values_from = nombre,
    values_fill = 0
  ) |>
  rename(CODGEO = CODGEO_2025)
```

The total number of crimes per year per city is computed by doing the sum of each type of crimes.
```{r}
crime_final <- crime_wide |>
  mutate(
    total_crimes = rowSums(
      across(-c(CODGEO, annee, insee_pop)),
      na.rm = TRUE
    ),
    year = annee
  ) |>
  select(-annee)
```


## Selecting cities present in the 2018 data

The analysis will only focus on the cities where all the informations are available on 2018, because of the IRIS nomenclature.
```{r}
years_of_interest <- 2016:2021

crime_final <- crime_final |>
  filter(year %in% years_of_interest)

ppsm <- ppsm |>
  filter(year %in% years_of_interest)

data_final <- crime_final |>
  left_join(poverty_city, by = c("CODGEO", "year")) |>
  left_join(ppsm, by = c("CODGEO", "year"))
```
```{r}
data_final <- crime_final |>
  left_join(poverty_city, by = c("CODGEO", "year")) |>
  left_join(ppsm, by = c("CODGEO", "year"))
```
```{r}
cities_2018 <- poverty_2018_city |>
  pull(CODGEO) |>
  unique()

data_final <- data_final |>
  filter(CODGEO %in% cities_2018) |>
  select(-IRIS)
```

Adding the cities names: 
```{r}
city_names <- vroom(
  here("data", "poverty", "pov2016.csv"),
  delim = ";",
  col_select = c(CODGEO, LIBGEO)
) |>
  distinct()

data_final <- data_final |>
  left_join(city_names, by = "CODGEO")
```

-\> We obtained a clean dataset for all years between 2016 and 2021

# Cleaned data tables : summary and most important variables

After loading the datasets, we needed to convert the IRIS code into the GEOCODE by weighting the poverty rate of each geocode by the population of the corresponding IRIS code. By doing that, we obtained clean data for 994 cities. We only focus on these cities in the analysis from 2016 to 2021. We also computed the poverty rate for each city.

We only kept the variables of interest by loading the datasets, that are the poverty rate, the first and last decile, the median income, the number of crimes per crime type and the real estate price (average price per square meter).Then we merged them into an only dataset that combines all the data per cities per years. 

```{r}
descriptions <- c(
  "Unique code that corresponds to one city. The first two numbers correspond to a department",
  "The city's population",
  "Number of domestic physical violence reported for the year",
  "Number of non domestic physical violence reported for the year",
  "Number of sexual violence reported for the year",
  "Number of armed robberies reported for the year",
  "Number of non armed violent robberies reported for the year",
  "Number of non violent robberies against persons reported for the year",
  "Number of home invasions reported for the year",
  "Number of vehicle theft reported for the year",
  "Number of theft from vehicles reported for the year",
  "Number of theft of vehicle accessories reported for the year",
  "Number of willful damage and vandalism reported for the year",
  "Number of reported drug use for the year",
  "Number of illegal drug use sanctioned by fixed fine reported for the year",
  "Number of drug dealing cases reported for the year",
  "Number of fraud and payment method fraud reported for the year",
  "Total number of crimes per cities reported for this year",
  "The concerned year",
  "Median income for this year",
  "Poverty rate (set at less than 60% of the median income)",
  "1st income decile",
  "9th income decile",
  "Price per square meter of the real estate goods sold for the year", 
  "The city's name"
)

summary <- data.table(
  Variable = names(data_final),
  Type = sapply(data_final, class),
  Description = descriptions
)

knitr::kable(summary)

dataset_info <- data.table(
  Info = c("Number of rows", "Number of columns"),
  Value = c(nrow(data_final), ncol(data_final)) 
)

knitr::kable(dataset_info)

```

# Research question :

Does the acceleration of the gentrification process in certain French cities with more than 5000 inhabitants (2016-2021) correlate with a decrease in recorded crime in these same municipalities, and is there a displacement effect towards non-gentrified cities?

The first goal will be to make a gentrification index per city with a few variables: square meter price growth, median wage growth, poverty rate growth.

Then we will select the cities with the highest gentrification index (say top 20%) and compare it to the rest of the cities : do we observe a declination in terms of delinquency between the top 20% and the others? If yes, do we observe a spatial displacement of crime to non-gentrifying areas?
