---
title: "The Relationship Between City-Level Gentrification and Crime Rates: A Study of French Cities (2016-2021)"
format: 
  pdf:
    fig-align: center
    fig-pos: 'H'
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = FALSE,        # Hide code
  message = FALSE,     # Hide messages
  warning = FALSE,     # Hide warnings
  results = 'markup'   # Keep output/tables visible
)
```

Please find the Github project here : https://github.com/LEOPAUL55/Noe_Leo_final.git

# Research question

Gentrification, theorized in the 1960s, is an urban process in which higher-income groups move into and transform historically working-class neighborhoods. This shift is typically marked by rising real estate prices and an increase in the city or the neighborhood's average standard of living. As wealthier residents renovate properties and demand grows, housing costs rise, often displacing lower-income residents to less central areas or other cities. This leads to social conflict that affects education, inequality, health, and crime.

This project examines whether gentrification in French cities correlates with a decrease in recorded crime within those municipalities, and whether crime is displaced to non-gentrified areas. Our goal is to assess the short- and medium-term impact of gentrification on crime across France, to inform more effective public policies.

We analyze public data from 897 French cities between 2016 and 2021. First, we construct a gentrification index to measure the intensity and geographic spread of the phenomenon across the period, allowing us to classify cities by their level of gentrification. Second, we use this index to estimate the direct and delayed effects of gentrification on crime trends—both overall and by crime type—in each municipality.

# Datasets descriptions

```{r, echo=FALSE}
#| message: false
here::i_am("Noe_Leo_Final.Rproj")
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(vroom)
library(data.table)
library(stringr)
library(plm)
library(psych)
library(mgcv)
library(viridis)
library(FSA)



# Load pre-filtered datasets
poverty <- readRDS(here("data", "cleaned", "poverty.rds"))
crime <- readRDS(here("data", "cleaned", "crime.rds"))
ppsm <- readRDS(here("data", "cleaned", "ppsm.rds"))
city_names <- readRDS(here("data", "cleaned", "city_names.rds"))

# Recreate data_final
data_final <- poverty |>
  left_join(crime, by = c("CODGEO", "year")) |>
  left_join(ppsm, by = c("CODGEO", "year")) |>
  left_join(city_names, by = "CODGEO") |>
  mutate(
    crime_rate = (total_crimes / insee_pop) * 100
  )

```

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  out.width = "85%",
  fig.width = 7,
  fig.height = 5
)

# Set global ggplot theme for all plots
theme_set(
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 10, color = "gray40", hjust = 0.5),
      plot.caption = element_text(size = 8, color = "gray50", hjust = 0.5),
      plot.title.position = "plot"
    )
)
```


To ensure reliable findings, we begin by building two core datasets. The first contains annual data by city (e.g., real estate prices, median income, poverty rate, and crime counts). The second, the evolution dataset, is derived from the first and will track changes over time.

The gentrification index is constructed using economic indicators per city per year: real estate price per square meter, median income, the poverty rate (below 60% of median income), and the first and ninth income deciles. For crime, we use the annual number of recorded crimes by type. Using population data, we then calculate overall and type-specific crime rates.

## Data sources

1.  **Real estate price - Datagouv / Chambre des notaires de Paris** 

The informations about the real estate price per city are available on the France state's site. They are given in the files called "Demande de valeur foncière" for every year, that is from 2016 to 2021. This is a collection of various variables on residential housing market at the city level between 2014 and 2024. What is interesting for us is the mean price by square meter of the goods that have been sold each year. This was published by Boris Mericksay. Unfortunatlely, the informations about Paris at the arrondissement level are missing in those files. These important informations can be found on the Chambre des notaires de Paris' site that gives the trimestrial average square meter price of selling. For simplicity, we only focus on the selling price of appartements in the city from 2016 to 2021 (trimester 4). The available file is a PDF that can be easily convert into a CSV file.

2.  **Average standard of living - Insee** 

The files "poverty" are collections of indicators of disposable income in France, at the city level, between 2016 and 2021. We will only focus on the poverty rate, the median standard of living, the first and the ninth deciles. Poverty data are provided as separate yearly files with heterogeneous structures. Variable names differ across years, and the 2018 data are reported at the IRIS level rather than the commune level. To handle this, we implement a loading strategy that reads each file separately while restricting imports to the variables of interest (years 2016:2021). Poverty indicators are harmonized across years. Non‑numeric values are explicitly converted to missing values during loading, but it won't affect our analysis because this only concerns small cities. The result is a consistent panel of indicators comparable across years.

3.  **Criminality - Datagouv / Ministère de l'intérieur** 

The CSV file "criminality" is a database of all the crimes and offences (except traffic offences) at the city level in France, published by French Police Force, since 2016. These offences were identified following a complaint filed by a victim, a report, a testimony, a flagrant offence, a denunciation, but also on the initiative of the security forces. The total number of crimes per year per city is computed by doing the sum of each type of crimes for every year. Thanks to the informations about the number of inhabitants per city given in the past files, it is is possible to compute the global crime rate and per type.

```{r, echo = FALSE}
info <- function(df) {
  data.frame(file = deparse(substitute(df)), Observations = nrow(df), Variables = ncol(df))
}
```

```{r, echo = FALSE}
info_ppsm <- info(ppsm)
info_poverty <- info(poverty)
info_crime <- info(crime) 
```

```{r}
#| label: summary-table
#| echo: false

descriptions <- c(

"Unique city code (first 2 digits = department)",
"Median income",
"Poverty rate (< 60% of median income)",
"1st income decile",
"9th income decile",
"Year",
"City population",
"Domestic physical violence",
"Non-domestic physical violence",
"Armed robberies",
"Violent robberies (no weapon)",
"Non-violent robberies",
"Home invasions",
"Vehicle thefts",
"Thefts from vehicles",
"Thefts of vehicle accessories",
"Vandalism",
"Drug use",
"Drug dealing",
"Fraud",
"Total crimes",
"Price per m²",
"City name",
"Crime rate per 100 inhabitants"
)

summary_table <- data.table(
  Variable = names(data_final),
  Type = sapply(data_final, function(x) class(x)[1]),
  Description = descriptions
)

knitr::kable(summary_table, caption = "Variables in the final dataset")
```


# Descriptive analysis of the crime rate evolution from 2016 to 2021 - comparison between violent and non violent crimes

To analyze whether gentrification impacts violent and property crimes differently, the original 13 crime categories are grouped into two broader types:

Violent crimes: domestic and non‑domestic physical violence, armed robbery, and unarmed violent robbery.

Non‑violent crimes: theft, vehicle‑related crime, burglary, vandalism, fraud, and drug‑related offenses.

Between 2016 and 2021, total crime and non‑violent crime follow a similar downward trajectory, while violent crime increased.

Note on exclusions: Sexual crimes and drug use offenses (usage de stupéfiants, AFD) were excluded from this analysis and from the total crime figures. Sexual crimes were removed because reporting trends are heavily influenced by social and cultural factors (e.g., the #MeToo movement), making them unreliable for standard temporal comparison. Drug use offenses were excluded as they were not reported at all during the first years of the period studied

```{r, echo=FALSE}
crime_wide <- data_final |>
  mutate(
    violent_total = rowSums(
      across(c(`Violences physiques intrafamiliales`,
               `Violences physiques hors cadre familial`,
               `Vols avec armes`,
               `Vols violents sans arme`)),
      na.rm = TRUE
    ),
    nonviolent_total = rowSums(
      across(c(`Vols sans violence contre des personnes`,
               `Vols de véhicule`,
               `Vols dans les véhicules`,
               `Vols d'accessoires sur véhicules`,
               `Cambriolages de logement`,
               `Destructions et dégradations volontaires`,
               `Escroqueries et fraudes aux moyens de paiement`,
               `Usage de stupéfiants`,
               `Trafic de stupéfiants`)),
      na.rm = TRUE
    ),
    total_excl_sexual = total_crimes,
    violent_rate = (violent_total / insee_pop) * 100,
    nonviolent_rate = (nonviolent_total / insee_pop) * 100,
    total_rate_excl = (total_excl_sexual / insee_pop) * 100
  ) |>
  group_by(year) |>
  summarise(
    total = median(total_rate_excl, na.rm = TRUE),
    violent = median(violent_rate, na.rm = TRUE),
    nonviolent = median(nonviolent_rate, na.rm = TRUE),
    .groups = "drop"
  )

# Scale factor for secondary axis
scale_factor <- max(crime_wide$nonviolent) / max(crime_wide$violent)

ggplot(crime_wide, aes(x = year)) +
  geom_line(aes(y = nonviolent, color = "Non-Violent"), size = 1.5) +
  geom_point(aes(y = nonviolent, color = "Non-Violent"), size = 4) +
  geom_line(aes(y = total, color = "Total"), size = 1.5) +
  geom_point(aes(y = total, color = "Total"), size = 4) +
  geom_line(aes(y = violent * scale_factor, color = "Violent"), size = 1.5) +
  geom_point(aes(y = violent * scale_factor, color = "Violent"), size = 4) +
  scale_y_continuous(
    name = "Non-Violent & Total (per 100 inh.)",
    sec.axis = sec_axis(~./scale_factor, name = "Violent Crime (per 100 inh.)")
  ) +
  scale_x_continuous(breaks = 2016:2021) +
  scale_color_manual(values = c("Total" = "#2c3e50", 
                                "Violent" = "#e74c3c", 
                                "Non-Violent" = "#3498db")) +
  labs(
    x = "Year",
    color = NULL,
    title = str_wrap("Crime Rate Evolution in French Cities (2016-2021)"),
    subtitle = str_wrap("Note: Violent crime uses secondary axis (right) | Excluding sexual violence")
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.position = "bottom",
    axis.title.y.right = element_text(color = "#e74c3c"),
    axis.text.y.right = element_text(color = "#e74c3c")
  )
```

# Building a gentrification index between first year and last year :

After obtaining clean data sets, we can now analyse it through several aspects. The first step is to build a gentrification index, that is a mathematical formula which enables to compute and compare the degree of gentrification per cities thanks to the available information in the data set (economic description, real estate price). Second, it will be possible to compute descriptive statistics about gentrification to have a better understanding of the process in France through the period. More precisely, to what extent are french cities affected by gentrification and where is the process the most intense on the map. In the same idea, descriptive statistics about crime in France will be computed to also understand this phenomenon and its evolution. Third, we will try to isolate with the data the causal link between the degree of gentrification and the criminality evolution in the cities using an econometric method.

Recalling the definition, gentrification is the process by which working class quarter are appropriated by upper classes, which exclude the initial population and replace it by a wealthier one. In the data, such a process can be observed using the data sets "ppsm" (real estate price) and "poverty" (economic information). Indeed, a gentrified city can be identified through a rise in the real estate price (Prix2Moyen) which will increase the city's standard living and rise the median income (MED) as well as the first and last decile (D1 and D9) and reduce the rate of poor people in the population (fall of TP60).

Using these data, we can create a gentrification index that is implementable to any city using the following formula:

**Gentrification** = (**price_var** * 0.25) + (**MED_var** * 0.20) + (**D9_var** * 0.10) + (**D1_var** * 0.35) + (**TP60_var** * 0.10)

Gentrification is not about the level of the variables, but their variation. Thus we need to compute the variation in percentage of each variable to implement the formula. Moreover, we justify the weight given to each variable thanks to the literature. According to the initial definition (Ruth Glass, 1964) combined with the work of Neil Smith and the Rent-Gap Theory (1979), gentrification mainly materialized by the displacement of the working class as well as the rise in the real estate price, that justifies the heavy weight on the first decile (D1_var) and the square meter price (price_var). To make the index meaningful, we standardize it using the z-score standardization and center it around 0 (if the index is equal to 0, the city is as gentrify as the national mean). Finally, we exclude the city that have a median income greater than 60% of the national one, considering they are already composed of upper classes, so it cannot be gentrified.

```{r, echo=FALSE}
#Taking the earliest year of data for each demographic area (CODGEO)
baseline <- data_final |>
  group_by(CODGEO) |>
  filter(year == min(year)) |>
  #capturing the median income as the starting point
  ungroup() |>
  select(CODGEO, baseline_med = MED)

#Making sure only areas below the "60% of median income" threshold are considered "gentrifiable" (by definition, only lower income areas can gentrify)
percentile_cutoff <- quantile(baseline$baseline_med, 0.60, na.rm = TRUE)
```

```{r, echo=FALSE}
gentrifiable_areas <- baseline |>
  filter(baseline_med < percentile_cutoff) |>
  pull(CODGEO)

# Calculating percentage changes from first to last year
evolution <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>  
  group_by(CODGEO) |>
  summarise(
    price_var   = ((last(Prixm2Moyen) / first(Prixm2Moyen)) - 1) * 100,
    med_inc_var = (last(MED) / first(MED) - 1) * 100,
    d9_var      = (last(D9) / first(D9) - 1) * 100,
    d1_var      = (last(D1) / first(D1) - 1) * 100,
    pov_var     = last(TP60) - first(TP60),
    .groups = "drop"
  ) |>
  left_join(city_names, by = "CODGEO")

# Flipping signs for interpretation : making increases in low incomes and decrease in poverty show as positive values: this aligns indicators so higher values = more gentrification
evolution <- evolution |>
  mutate(
    pov_var = -pov_var
  )
```

```{r, echo=FALSE}
# Converting all the variables to z-score so it makes the different metrics comparable despite different scales 
evolution_std <- evolution |>
  mutate(across(
    c(price_var, med_inc_var, d9_var, d1_var, pov_var),
    scale
  ))
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Creating weighted gentrification index
evolution_std <- evolution_std |>
  mutate(
    gent_index = 
      (price_var * 0.25) +
      (med_inc_var * 0.20) +
      (d9_var * 0.10) +
      (d1_var * 0.35) +
      (pov_var * 0.10)
  ) |>
  arrange(desc(gent_index))

evolution_std_A <- evolution_std

# Display top 10 gentrified cities
evolution_std_A |>
  head(10) |>
  select(CODGEO, LIBGEO, gent_index) |>
  mutate(gent_index = round(as.numeric(gent_index), 3)) |>
  knitr::kable(
    col.names = c("City Code", "City Name", "Gentrification Index"),
    booktabs = TRUE,
    align = c("l", "l", "r"),
    caption = "Top 10 Most Gentrified Cities (2016-2021)"
  )
```


# Cross sectional analysis

After constructing the gentrification index for all eligible French cities (those with baseline median income below the 60th percentile of the national distribution), we proceed to examine the relationship between gentrification intensity and changes in crime rates over the 2016–2021 period. The analysis is cross-sectional: for each city, we compute the percentage change in crime rate between the first and last year of observation, and correlate this change with the standardized gentrification index.

To facilitate interpretation, we categorize cities into three groups based on their gentrification index:

Low gentrification: bottom 25% of the index distribution

Medium gentrification: middle 50%

High gentrification: top 25%

These categories allow us to compare how crime evolved differently across areas experiencing varying degrees of socioeconomic transformation.
```{r, echo=FALSE}
#| include: false

evolution_std_A_groups <- evolution_std_A |>
  mutate(
    gent_category = case_when(
      gent_index > quantile(gent_index, 0.75, na.rm = TRUE) ~ "High",
      gent_index > quantile(gent_index, 0.25, na.rm = TRUE) ~ "Medium",
      TRUE ~ "Low"
    ),
    gent_category = factor(gent_category, levels = c("Low", "Medium", "High"))
  )

table(evolution_std_A_groups$gent_category)

crime_rate_change_continuous <- data_final |>
  filter(year %in% c(2016, 2021)) |>
  select(CODGEO, year, total_crimes, insee_pop) |>
  pivot_wider(
    names_from = year,
    values_from = c(total_crimes, insee_pop),
    names_sep = "_"
  ) |>
  mutate(
    crime_rate_2016 = total_crimes_2016 / insee_pop_2016,
    crime_rate_2021 = total_crimes_2021 / insee_pop_2021,
    crime_rate_change = (crime_rate_2021 - crime_rate_2016) / crime_rate_2016
  ) |>
  left_join(
    evolution_std_A_groups |> select(CODGEO, gent_index, gent_category),
    by = "CODGEO"
  ) |>
  filter(!is.na(gent_category))

```

## Visual exploration: Scatter plot

The scatter plot below displays the relationship between the gentrification index (horizontal axis) and the percentage change in crime rate (vertical axis) for all 897 cities in our sample.

```{r, echo=FALSE}
#| echo: false
#| fig-align: center
#| fig-width: 8
#| fig-height: 6
#| out-width: "100%"

library(viridis)

ggplot(crime_rate_change_continuous,
       aes(x = gent_index, y = crime_rate_change)) +
  geom_point(aes(color = gent_index), 
             alpha = 0.6, size = 2.5) +
  scale_color_viridis_c(option = "plasma", name = "Gentrification\nIntensity") +
  geom_smooth(method = "lm", color = "#FF6B6B", size = 1.2, 
              fill = "#FF6B6B", alpha = 0.15) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.3, color = "gray40") +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.3, color = "gray40") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_cartesian(ylim = c(-0.5, 0.5)) +
  labs(
    x = "Gentrification Index (standardized)",
    y = "Crime Rate Change (2016-2021)",
    title = "Small Link Between Gentrification\nand Crime Changes",
    subtitle = "Analysis of initially lower-income areas (bottom 60%)",
    caption = "Each point = one city | Red line = linear trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, color = "gray40", hjust = 0.5),
    plot.caption = element_text(size = 8, color = "gray50"),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )
```
Visually, a slight positive trend emerges: cities with higher gentrification indices tend to experience smaller reductions (or even increases) in crime rates compared to less gentrified cities. The cloud of points is dispersed, suggesting considerable heterogeneity. To formally test whether this visual pattern reflects a true association, we use a Pearson correlation test—the standard method for assessing linear relationships between two continuous variables. The results show no significant linear relationship between gentrification intensity and crime rate change (r = 0.052, df = 535, p = 0.227).

```{r}
#| echo: false

# Pearson correlation test
cor_result <- cor.test(
  crime_rate_change_continuous$gent_index,
  crime_rate_change_continuous$crime_rate_change,
  use = "complete.obs"
)

# Create clean summary table
tibble(
  Statistic = c("Pearson correlation (r)", "t-statistic", "Degrees of freedom", "P-value", "95% CI (lower)", "95% CI (upper)"),
  Value = c(
    round(cor_result$estimate, 4),
    round(cor_result$statistic, 3),
    cor_result$parameter,
    ifelse(cor_result$p.value < 0.001, "< 0.001", round(cor_result$p.value, 4)),
    round(cor_result$conf.int[1], 4),
    round(cor_result$conf.int[2], 4)
  )
) |>
  knitr::kable(
    booktabs = TRUE,
    align = c("l", "r"),
    caption = "Pearson Correlation: Gentrification Index vs. Crime Rate Change"
  )
```


## Comparing crime rate changes across gentrification levels 
To better quantify these differences, we compute the mean and median crime rate changes for each gentrification category, along with 95% confidence intervals.

```{r, echo=FALSE}
#| include: false

evolution_std_A_groups <- evolution_std_A |>
  mutate(
    gent_category = case_when(
      gent_index > quantile(gent_index, 0.75, na.rm = TRUE) ~ "High",
      gent_index > quantile(gent_index, 0.25, na.rm = TRUE) ~ "Medium",
      TRUE ~ "Low"
    ),
    gent_category = factor(gent_category, levels = c("Low", "Medium", "High"))
  )

```


```{r, echo=FALSE}
#| echo: false
#| fig-align: center
#| fig-width: 7
#| fig-height: 5

# Create gentrification categories
evolution_std_A_groups <- evolution_std_A |>
  mutate(
    gent_category = case_when(
      gent_index > quantile(gent_index, 0.75, na.rm = TRUE) ~ "High",
      gent_index > quantile(gent_index, 0.25, na.rm = TRUE) ~ "Medium",
      TRUE ~ "Low"
    ),
    gent_category = factor(gent_category, levels = c("Low", "Medium", "High"))
  )

# Calculate confidence intervals
ci_stats <- crime_rate_change_continuous |>
  group_by(gent_category) |>
  summarise(
    mean = mean(crime_rate_change, na.rm = TRUE),
    ci_lower = mean - 1.96 * sd(crime_rate_change, na.rm = TRUE) / sqrt(n()),
    ci_upper = mean + 1.96 * sd(crime_rate_change, na.rm = TRUE) / sqrt(n()),
    median = median(crime_rate_change, na.rm = TRUE)
  )

# Define viridis plasma colors for 3 categories
plasma_colors <- viridis::plasma(3, begin = 0.2, end = 0.9)

ggplot(ci_stats |> filter(gent_category %in% c("Low", "Medium", "High")),
       aes(x = gent_category, y = mean, color = gent_category)) +
  # CI lines
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.15, size = 1.5, alpha = 0.6) +
  # Mean points
  geom_point(size = 6) +
  # Median points
  geom_point(aes(y = median), shape = 1, size = 6, stroke = 1.5) +
  scale_color_manual(values = plasma_colors) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5, color = "gray40") +
  coord_cartesian(ylim = c(-0.15, 0.05)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    x = "Gentrification Level",
    y = "Crime Rate Change (2016–2021)",
    title = "Mean Crime Rate Changes\nAcross Gentrification Levels",
    subtitle = "Solid dots = mean | Hollow circles = median | Lines = 95% CI"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, color = "gray40", hjust = 0.5),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 11),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```
All three categories experienced a reduction in crime rates between 2016 and 2021, consistent with national crime trends in France during this period. While the magnitude of reduction appears to vary visually across gentrification levels, we need to formally test whether these differences are statistically significant. 

Since we are comparing the means of a continuous variable (crime rate change) across three independent groups (Low, Medium, High gentrification), we use a one-way ANOVA—the standard parametric test for this setting.
```{r, echo=FALSE}
#| echo: false

# ANOVA test
anova_result <- aov(crime_rate_change ~ gent_category, data = crime_rate_change_continuous)
anova_summary <- summary(anova_result)

# Extract key statistics
f_val <- anova_summary[[1]]$`F value`[1]
p_val <- anova_summary[[1]]$`Pr(>F)`[1]
df1 <- anova_summary[[1]]$Df[1]
df2 <- anova_summary[[1]]$Df[2]

# Create clean summary table
tibble(
  Statistic = c("F-statistic", "Degrees of freedom (groups)", "Degrees of freedom (residuals)", "P-value"),
  Value = c(
    round(f_val, 3),
    df1,
    df2,
    ifelse(p_val < 0.001, "< 0.001", round(p_val, 4))
  )
) |>
  knitr::kable(
    booktabs = TRUE,
    align = c("l", "r"),
    caption = "One-Way ANOVA: Crime Rate Change Across Gentrification Levels"
  )
```
The results indicate no statistically significant difference in crime rate changes across gentrification levels (p > 0.05). Despite the visual variation observed in the previous figure, we cannot reject the null hypothesis that all three groups experienced similar crime rate changes.

## Comparing crime rate change by type among gentrification levels

While total crime rates show only a  diminution for all gentrification levels, different crime categories may respond differently to socioeconomic changes. We therefore disaggregate crime into 13 specific types and examine how each evolved across gentrification intensities between 2016 and 2021.

For each city, we compute the population‑adjusted change in crime rates for each of the 13 crime categories, thus accounting for both changes in crime counts and changes in population size, providing a cleaner measure of crime‑rate evolution.
```{r, echo=FALSE}
#| echo: false
#| fig-align: center
#| fig-width: 9
#| fig-height: 7

# Reshaping crime data per type
crime_change_type_continuous <- data_final |>
  pivot_longer(
    cols = c(
      `Violences physiques intrafamiliales`,
      `Violences physiques hors cadre familial`,
      `Vols avec armes`,
      `Vols violents sans arme`,
      `Vols sans violence contre des personnes`,
      `Vols de véhicule`,
      `Vols dans les véhicules`,
      `Vols d'accessoires sur véhicules`,
      `Cambriolages de logement`,
      `Destructions et dégradations volontaires`,
      `Escroqueries et fraudes aux moyens de paiement`,
      `Usage de stupéfiants`,
      `Trafic de stupéfiants`
    ),
    names_to = "crime_type",
    values_to = "crime_count"
  ) |>
  filter(year %in% c(2016, 2021)) |>
  select(CODGEO, year, crime_type, crime_count, insee_pop) |>
  
  # Creating year-over-year comparison
  pivot_wider(
    names_from = year,
    values_from = c(crime_count, insee_pop),
    names_sep = "_"
  ) |>
  # Calculating population-adjusted change
  mutate(
    pop_adj_rate_change = if_else(
      crime_count_2016 > 0 & insee_pop_2016 > 0,
      ((crime_count_2021 / crime_count_2016) /
       (insee_pop_2021 / insee_pop_2016)) - 1,
      NA_real_
    )
  ) |>
  # Merging with gentrification levels
  left_join(
    evolution_std_A_groups |> select(CODGEO, gent_index, gent_category),
    by = "CODGEO"
  ) |>
  filter(!is.na(gent_category))

# Summary by crime type and gentrification level
crime_change_type_summary <- crime_change_type_continuous |>
  filter(is.finite(pop_adj_rate_change)) |>
  group_by(crime_type, gent_category) |>
  summarise(
    median_change = median(pop_adj_rate_change, na.rm = TRUE),
    n_cities = n(),
    .groups = "drop"
  )

# Define viridis plasma colors (same as before)
plasma_colors <- viridis::plasma(3, begin = 0.2, end = 0.9)

# Plot
ggplot(crime_change_type_summary,
       aes(x = crime_type, y = median_change, fill = gent_category)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = plasma_colors) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    x = NULL,
    y = "Population-Adjusted Crime Change (2016–2021)",
    fill = "Gentrification\nIntensity",
    title = str_wrap("Crime Change by Type and Gentrification Intensity", width = 50)
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.text.y = element_text(size = 9),
    legend.position = "right",
    panel.grid.major.y = element_blank()
  )
```
For each individual crime type, the direction of change (increase or decrease) is consistent across all three gentrification categories. However, the magnitude of change varies considerably depending on the gentrification intensity for certain crime types. 

Kruskal-Wallis tests were conducted for each of the 13 crime categories 
to assess whether population-adjusted crime rate changes differ 
significantly across gentrification levels.

Only one crime type showed a statistically significant difference: 
**domestic physical violence** (Violences physiques intrafamiliales), 
with X² = 19.09, p < 0.001. This indicates that gentrification 
intensity significantly predicts how domestic violence evolved (positively)
between 2016 and 2021.

```{r, echo=FALSE}
# Run Kruskal-Wallis for each crime type
kruskal_by_crime_type <- crime_change_type_continuous |>
  filter(is.finite(pop_adj_rate_change)) |>
  group_by(crime_type) |>
  summarise(
    chi_squared = kruskal.test(pop_adj_rate_change ~ gent_category)$statistic,
    p_value = kruskal.test(pop_adj_rate_change ~ gent_category)$p.value,
    n = n(),
    .groups = "drop"
  ) |>
  mutate(
    significant = ifelse(p_value < 0.05, "Yes", "No")
  ) |>
  arrange(p_value)

# Display results
knitr::kable(kruskal_by_crime_type, 
             digits = 4,
             col.names = c("Crime Type", "Chi-squared", "P-value", "N", "Significant"))
```

## Violent vs Non-violent breakdown

Using the violent and non-violent crime categories defined in the 
descriptive analysis section, we now examine whether gentrification 
affects these two broad categories differently.

```{r, echo=FALSE}

# Creating crime categories 
crime_change_type_cat <- crime_change_type_continuous |>
  mutate(
    crime_category = case_when(
      crime_type %in% c(
       
        "Violences physiques intrafamiliales",
        "Violences physiques hors cadre familial",
        "Vols avec armes",
        "Vols violents sans arme"
      ) ~ "Violent crimes",
      crime_type %in% c(
        "Vols sans violence contre des personnes",
        "Vols de véhicule",
        "Vols dans les véhicules",
        "Vols d'accessoires sur véhicules",
        "Cambriolages de logement",
        "Destructions et dégradations volontaires",
        "Escroqueries et fraudes aux moyens de paiement",
        "Usage de stupéfiants",
        "Trafic de stupéfiants"
      ) ~ "Non violent crimes",
      TRUE ~ NA_character_
    )
  )
```


```{r, echo=FALSE}

# Calculating the median population adjusted change for each combination of crime category and gentrification level
crime_cat_summary <- crime_change_type_cat |>
  filter(is.finite(pop_adj_rate_change), !is.na(crime_category)) |>
  group_by(crime_category, gent_category) |>
  summarise(
    median_change = median(pop_adj_rate_change, na.rm = TRUE),
    n_obs = n(),
    .groups = "drop"
  )
```


```{r, echo=FALSE}
# Plot
#| echo: false
#| fig-align: center
#| fig-width: 7
#| fig-height: 5

# Define viridis plasma colors (same as all other graphs)
plasma_colors <- viridis::plasma(3, begin = 0.2, end = 0.9)

# Plot
ggplot(crime_cat_summary,
       aes(x = crime_category, y = median_change, fill = gent_category)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = plasma_colors) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    x = "Crime Category",
    y = "Population-Adjusted Crime Rate Change (2016–2021)",
    fill = "Gentrification\nIntensity",
    title = "Violent vs Non-Violent Crime Rate Change\nby Gentrification Intensity"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    axis.text.x = element_text(size = 10, face = "bold"),
    legend.position = "right"
  )

# Table
knitr::kable(
  crime_cat_summary,
  digits = 3,
  booktabs = TRUE,
  col.names = c("Crime Category", "Gentrification Level", "Median Change", "N Observations"),
  caption = "Violent vs Non-Violent Crime Rate Change by Gentrification Level"
)
```
The results reveal a clear gradient pattern: higher gentrification is associated with smaller crime changes in both directions.

Violent crimes increased everywhere, but the increase was much smaller in high-gentrification cities (+5%) compared to low-gentrification cities (+24%).

Non-violent crimes decreased most in low-gentrification cities (-18%), moderately in medium-gentrification cities, and least in high-gentrification cities.

This suggests that gentrification has a stabilizing effect on crime rates, dampening both the rise in violent crime and the decline in non-violent crime. Gentrifying areas experience more moderate changes compared to non-gentrifying areas.

To verify whether these differences are statistically significant, we conduct pairwise Wilcoxon tests with Bonferroni correction. This correction adjusts p-values to account for multiple comparisons, reducing the risk of false positives.:

```{r, echo=FALSE}
# Prepare data for violent crimes
violent_data <- crime_change_type_cat |>
  filter(crime_category == "Violent crimes", is.finite(pop_adj_rate_change)) |>
  group_by(CODGEO, gent_category) |>
  summarise(median_change = median(pop_adj_rate_change), .groups = "drop")

# Prepare data for non-violent crimes
nonviolent_data <- crime_change_type_cat |>
  filter(crime_category == "Non violent crimes", is.finite(pop_adj_rate_change)) |>
  group_by(CODGEO, gent_category) |>
  summarise(median_change = median(pop_adj_rate_change), .groups = "drop")

# Pairwise Wilcoxon tests
violent_pairwise <- pairwise.wilcox.test(
  violent_data$median_change,
  violent_data$gent_category,
  p.adjust.method = "bonferroni"
)

nonviolent_pairwise <- pairwise.wilcox.test(
  nonviolent_data$median_change,
  nonviolent_data$gent_category,
  p.adjust.method = "bonferroni"
)

# Create combined table
tibble(
  Comparison = c("Low vs Medium", "Low vs High", "Medium vs High"),
  `Violent crimes (p-value)` = c(
    violent_pairwise$p.value["Medium", "Low"],
    violent_pairwise$p.value["High", "Low"],
    violent_pairwise$p.value["High", "Medium"]
  ),
  `Non-violent crimes (p-value)` = c(
    nonviolent_pairwise$p.value["Medium", "Low"],
    nonviolent_pairwise$p.value["High", "Low"],
    nonviolent_pairwise$p.value["High", "Medium"]
  )
) |>
  knitr::kable(digits = 4, caption = "Pairwise Wilcoxon Tests (Bonferroni corrected)")
```
For violent crimes, high-gentrification cities differ significantly from both low-gentrification (p < 0.001) and medium-gentrification cities (p < 0.001), confirming that the dampening effect of gentrification on violent crime is statistically robust.

For non-violent crimes, no pairwise comparisons reach statistical significance (all p > 0.05), suggesting that the observed differences may be due to chance.

# Panel Analysis

The cross-sectional analysis examined the relationship between gentrification and crime changes over the entire 2016-2021 period. However, this approach cannot capture the temporal dynamics of how gentrification affects crime over time. Does gentrification in one year affect crime the following year? Do effects accumulate or reverse over multiple years?

To address these questions, we construct a year-over-year gentrification index that measures annual changes in gentrification intensity for each city. This allows us to test whether lagged gentrification predicts subsequent crime changes using panel data methods.

## Year over year gentrification index

Instead of measuring gentrification as the total change from 2016 to 2021, we now compute annual growth rates for each component of the gentrification index. This creates a time-varying measure that captures the intensity of gentrification in each city-year.

```{r, echo=FALSE}
gent_yearly <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    # Year-over-year growth rates (log differences)
    price_g = log(Prixm2Moyen) - log(dplyr::lag(Prixm2Moyen)),
    med_g   = log(MED) - log(dplyr::lag(MED)),
    d9_g    = log(D9) - log(dplyr::lag(D9)),
    d1_g    = -(log(D1) - log(dplyr::lag(D1))),
    pov_g   = -(TP60 - dplyr::lag(TP60))
  ) |>
  ungroup() |>
  filter(!is.na(price_g)) |>
  mutate(
    # Standardizing within the dataset
    price_g_s = as.numeric(scale(price_g)),
    med_g_s   = as.numeric(scale(med_g)),
    d9_g_s    = as.numeric(scale(d9_g)),
    d1_g_s    = as.numeric(scale(d1_g)),
    pov_g_s   = as.numeric(scale(pov_g)),
    
    # Weighted index (SAME WEIGHTS as first-to-last year index)
    gent_index = (price_g_s * 0.25) +
                 (med_g_s * 0.20) +
                 (d9_g_s * 0.10) +
                 (d1_g_s * 0.35) +
                 (pov_g_s * 0.10)
  )
# Calculate year-over-year crime rate change
crime_yearly <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    crime_rate = total_crimes / insee_pop,
    crime_change = log(crime_rate) - log(dplyr::lag(crime_rate))
  ) |>
  ungroup() |>
  select(CODGEO, year, crime_rate, crime_change, total_crimes, insee_pop)

# Create the panel by joining gentrification index with crime data
crime_panel <- gent_yearly |>
  select(CODGEO, year, gent_index) |>
  left_join(crime_yearly, by = c("CODGEO", "year")) |>
  filter(!is.na(crime_change), !is.na(gent_index))

```

The resulting panel contains 2,688 city-year observations across 538 cities from 2017 to 2021. The gentrification index is centered around zero (mean ≈ 0) with values ranging from -2.29 to +2.20, indicating substantial variation in annual gentrification intensity across cities and years.


## Lagged effects

### On total crimes

To test whether gentrification in one year affects crime the following year, we estimate a fixed effects panel model. This approach controls for both time-invariant city characteristics (city fixed effects) and common yearly shocks (year fixed effects).

```{r, echo=FALSE}
#| echo: false


# Create panel data
panel_data <- pdata.frame(crime_panel, index = c("CODGEO", "year"))

# Fixed effects model
model <- plm(crime_change ~ lag(gent_index), 
             data = panel_data, 
             model = "within",
             effect = "twoways")

# Extract key statistics
model_summary <- summary(model)
beta_val <- round(model_summary$coefficients[1, "Estimate"], 3)
se_val <- round(model_summary$coefficients[1, "Std. Error"], 3)
t_val <- round(model_summary$coefficients[1, "t-value"], 3)
p_val <- model_summary$coefficients[1, "Pr(>|t|)"]
r_sq <- round(model_summary$r.squared["rsq"], 3)

# Create clean summary table
tibble(
  Statistic = c("Coefficient (β)", "Standard Error", "t-statistic", "P-value", "R-squared"),
  Value = c(
    beta_val,
    se_val,
    t_val,
    ifelse(p_val < 0.001, "< 0.001", round(p_val, 3)),
    r_sq
  )
) |>
  knitr::kable(
    booktabs = TRUE,
    align = c("l", "r"),
    caption = "Fixed Effects Panel Model: Lagged Gentrification on Crime Change"
  )
```
The results show a small but statistically significant negative effect (β = -0.015, p = 0.03). Cities that experienced more gentrification in one year tended to have slightly smaller increases (or larger decreases) in total crime the following year. However, the effect size is modest: the model explains only 0.3% of the variance in crime changes (R² = 0.003).

## On violent and non violent crimes

Given that our cross-sectional analysis found significant effects specifically for violent crimes, we now examine whether this pattern holds in the panel setting. We test whether gentrification in year t predicts changes in violent and non-violent crime in year t+1.

```{r, echo=FALSE}
# Creating violent crimes panel
violent_crimes_panel <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  mutate(
    violent_total = rowSums(
      across(c(`Violences physiques intrafamiliales`,
               `Violences physiques hors cadre familial`,
               `Vols avec armes`,
               `Vols violents sans arme`)),
      na.rm = TRUE
    ),
    violent_rate = violent_total / insee_pop
  ) |>
  left_join(
    gent_yearly |> select(CODGEO, year, gent_index),
    by = c("CODGEO", "year")
  ) |>
  group_by(CODGEO) |>
  mutate(
    violent_change = log(violent_rate + 0.001) - log(dplyr::lag(violent_rate) + 0.001),
    gent_index_lag1 = dplyr::lag(gent_index, 1)
  ) |>
  ungroup()

# Creating non-violent crimes panel
nonviolent_crimes_panel <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  mutate(
    nonviolent_total = rowSums(
      across(c(`Vols sans violence contre des personnes`,
               `Vols de véhicule`,
               `Vols dans les véhicules`,
               `Vols d'accessoires sur véhicules`,
               `Cambriolages de logement`,
               `Destructions et dégradations volontaires`,
               `Escroqueries et fraudes aux moyens de paiement`,
               `Usage de stupéfiants`,
               `Trafic de stupéfiants`)),
      na.rm = TRUE
    ),
    nonviolent_rate = nonviolent_total / insee_pop
  ) |>
  left_join(
    gent_yearly |> select(CODGEO, year, gent_index),
    by = c("CODGEO", "year")
  ) |>
  group_by(CODGEO) |>
  mutate(
    nonviolent_change = log(nonviolent_rate + 0.001) - log(dplyr::lag(nonviolent_rate) + 0.001),
    gent_index_lag1 = dplyr::lag(gent_index, 1)
  ) |>
  ungroup()

# Filtering for analysis
violent_for_analysis <- violent_crimes_panel |>
  filter(!is.na(violent_change) & !is.na(gent_index_lag1))

nonviolent_for_analysis <- nonviolent_crimes_panel |>
  filter(!is.na(nonviolent_change) & !is.na(gent_index_lag1))

# Correlation tests
cor_violent <- cor.test(
  violent_for_analysis$gent_index_lag1,
  violent_for_analysis$violent_change,
  use = "complete.obs"
)

cor_nonviolent <- cor.test(
  nonviolent_for_analysis$gent_index_lag1,
  nonviolent_for_analysis$nonviolent_change,
  use = "complete.obs"
)

# Summary table
tibble(
  `Crime Type` = c("Violent crimes", "Non-violent crimes"),
  Correlation = c(cor_violent$estimate, cor_nonviolent$estimate),
  `P-value` = c(cor_violent$p.value, cor_nonviolent$p.value),
  N = c(nrow(violent_for_analysis), nrow(nonviolent_for_analysis))
) |>
  knitr::kable(digits = 4, caption = "Lagged Gentrification Effect on Crime (1-year lag)")
```
Violent crimes show a significant negative correlation (r = -0.102, p < 0.001): cities that experienced more gentrification in one year tended to have smaller increases (or larger decreases) in violent crime the following year.

Non-violent crimes show a stronger and also significant negative correlation (r = -0.178, p < 0.001): gentrification is associated with subsequent reductions in non-violent crime as well.

Interestingly, the lagged effect is stronger for non-violent crimes than for violent crimes in the panel setting, which contrasts with our cross-sectional findings. This suggests that while gentrification's immediate (same-period) effect may differ by crime type, its delayed effect tends to reduce both categories of crime.

## Multiple lags analysis

```{r, echo=FALSE}
# Creating multiple lags for violent crimes
violent_with_lags <- violent_crimes_panel |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    gent_index_lag2 = dplyr::lag(gent_index, 2),
    gent_index_lag3 = dplyr::lag(gent_index, 3)
  ) |>
  ungroup()

# Creating multiple lags for non-violent crimes
nonviolent_with_lags <- nonviolent_crimes_panel |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    gent_index_lag2 = dplyr::lag(gent_index, 2),
    gent_index_lag3 = dplyr::lag(gent_index, 3)
  ) |>
  ungroup()

# Creating multiple lags for total crime
total_with_lags <- crime_panel |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    gent_index_lag2 = dplyr::lag(gent_index, 2),
    gent_index_lag3 = dplyr::lag(gent_index, 3)
  ) |>
  ungroup()

# Function to compute correlations for each lag
compute_lag_cors <- function(data, change_var) {
  lag1_data <- data |> filter(!is.na({{change_var}}) & !is.na(gent_index_lag1))
  lag2_data <- data |> filter(!is.na({{change_var}}) & !is.na(gent_index_lag2))
  lag3_data <- data |> filter(!is.na({{change_var}}) & !is.na(gent_index_lag3))
  
  cor_lag1 <- cor.test(lag1_data$gent_index_lag1, pull(lag1_data, {{change_var}}))
  cor_lag2 <- cor.test(lag2_data$gent_index_lag2, pull(lag2_data, {{change_var}}))
  cor_lag3 <- cor.test(lag3_data$gent_index_lag3, pull(lag3_data, {{change_var}}))
  
  tibble(
    Lag = c(1, 2, 3),
    Correlation = c(cor_lag1$estimate, cor_lag2$estimate, cor_lag3$estimate),
    P_value = c(cor_lag1$p.value, cor_lag2$p.value, cor_lag3$p.value),
    N = c(nrow(lag1_data), nrow(lag2_data), nrow(lag3_data))
  )
}

# Compute for each crime type
total_cors <- compute_lag_cors(total_with_lags, crime_change) |> mutate(Crime_Type = "Total")
violent_cors <- compute_lag_cors(violent_with_lags, violent_change) |> mutate(Crime_Type = "Violent")
nonviolent_cors <- compute_lag_cors(nonviolent_with_lags, nonviolent_change) |> mutate(Crime_Type = "Non-violent")

# Combine and reshape to wide format
all_lag_results <- bind_rows(total_cors, violent_cors, nonviolent_cors) |>
  mutate(
    P_value = ifelse(P_value < 0.001, "< 0.001", round(P_value, 4)),
    Result = paste0(round(Correlation, 3), " (", P_value, ")")
  ) |>
  select(Crime_Type, Lag, Result) |>
  pivot_wider(
    names_from = Lag,
    values_from = Result,
    names_prefix = "Lag "
  )

knitr::kable(all_lag_results, 
             col.names = c("Crime Type", "1 year", "2 years", "3 years"),
             caption = "Gentrification Effects on Crime at Different Lags (Correlation, P-value)")
```

This oscillating pattern suggests that gentrification's effect on crime is not permanent. The initial crime reduction is partially reversed in the short-medium term, possibly reflecting displacement effects, adjustment dynamics, or statistical mean reversion. Notably, total and non-violent crime follow nearly identical patterns, while violent crime shows the same directional effects but with smaller magnitudes.

To ensure robustness, we estimate fixed effects panel models controlling for year effects and mean reversion in crime rates. Year fixed effects are particularly important given that 2020 saw an unusual decrease in crime likely due to COVID-19 lockdowns, which could otherwise confound our results. The models include gentrification at 1-year and 2-year lags; we do not extend to 3 lags as our 6-year panel (2016-2021) would leave too few observations for reliable estimation.
```{r, echo=FALSE}
# Prepare violent crimes data
panel_violent_robust <- violent_with_lags |>
  group_by(CODGEO) |>
  mutate(violent_change_lag1 = dplyr::lag(violent_change, 1)) |>
  ungroup() |>
  filter(!is.na(violent_change) & 
         !is.na(gent_index_lag1) & 
         !is.na(gent_index_lag2) & 
         !is.na(violent_change_lag1)) |>
  pdata.frame(index = c("CODGEO", "year"))

# Prepare non-violent crimes data
nonviolent_with_lags <- nonviolent_crimes_panel |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    gent_index_lag2 = dplyr::lag(gent_index, 2),
    nonviolent_change_lag1 = dplyr::lag(nonviolent_change, 1)
  ) |>
  ungroup()

panel_nonviolent_robust <- nonviolent_with_lags |>
  filter(!is.na(nonviolent_change) & 
         !is.na(gent_index_lag1) & 
         !is.na(gent_index_lag2) & 
         !is.na(nonviolent_change_lag1)) |>
  pdata.frame(index = c("CODGEO", "year"))

# Prepare total crime data
total_with_lags <- crime_panel |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    gent_index_lag2 = dplyr::lag(gent_index, 2),
    crime_change_lag1 = dplyr::lag(crime_change, 1)
  ) |>
  ungroup()

panel_total_robust <- total_with_lags |>
  filter(!is.na(crime_change) & 
         !is.na(gent_index_lag1) & 
         !is.na(gent_index_lag2) & 
         !is.na(crime_change_lag1)) |>
  pdata.frame(index = c("CODGEO", "year"))

# Model for violent crimes
model_violent <- plm(
  violent_change ~ gent_index_lag1 + gent_index_lag2 + violent_change_lag1,
  data = panel_violent_robust,
  model = "within",
  effect = "twoways"
)

# Model for non-violent crimes
model_nonviolent <- plm(
  nonviolent_change ~ gent_index_lag1 + gent_index_lag2 + nonviolent_change_lag1,
  data = panel_nonviolent_robust,
  model = "within",
  effect = "twoways"
)

# Model for total crime
model_total <- plm(
  crime_change ~ gent_index_lag1 + gent_index_lag2 + crime_change_lag1,
  data = panel_total_robust,
  model = "within",
  effect = "twoways"
)

# Extract coefficients and p-values
extract_results <- function(model, crime_type) {
  coefs <- summary(model)$coefficients
  tibble(
    `Crime Type` = crime_type,
    Variable = c("Gentrification (Lag 1)", "Gentrification (Lag 2)", "Mean reversion"),
    Estimate = coefs[1:3, "Estimate"],
    `P-value` = coefs[1:3, "Pr(>|t|)"]
  )
}

# Combine results
results_table <- bind_rows(
  extract_results(model_total, "Total"),
  extract_results(model_violent, "Violent"),
  extract_results(model_nonviolent, "Non-violent")
) |>
  mutate(
    `P-value` = ifelse(`P-value` < 0.001, "< 0.001", round(`P-value`, 3)),
    Significant = case_when(
      as.numeric(ifelse(`P-value` == "< 0.001", 0, `P-value`)) < 0.05 ~ "Yes",
      as.numeric(ifelse(`P-value` == "< 0.001", 0, `P-value`)) < 0.10 ~ "Borderline",
      TRUE ~ "No"
    )
  )

knitr::kable(results_table, 
             digits = 3,
             caption = "Fixed Effects Panel Models with Year Controls and Mean Reversion")
```
The results confirm that gentrification has a modest but significant crime-reducing effect at the 1-year lag, particularly for total crime (β = -0.019, p = 0.022) and violent crime (β = -0.027, p = 0.020). The effect on non-violent crime is borderline significant (p = 0.052).

Interestingly, only violent crime shows a potential cumulative effect at the 2-year lag (β = -0.022, p = 0.054), suggesting that gentrification's impact on interpersonal violence may persist longer than its effect on property crime.

All models show strong mean reversion (β ≈ -0.50), indicating that crime rates naturally oscillate over time. After controlling for this, gentrification explains a small but statistically significant portion of crime changes.

# Robustness checks

## Granger Causality

Our analysis assumes that gentrification affects crime, but the reverse could also be true: areas with declining crime might attract more investment and gentrification. To test the direction of causality, we estimate two models for each crime type:

Does lagged gentrification predict crime changes?
Does lagged crime change predict gentrification?

```{r, echo=FALSE}
library(plm)

# Function to run Granger causality tests
run_granger_tests <- function(crime_data, crime_change_var, crime_type_name) {
  
  # Prepare data
  granger_data <- crime_data |>
    filter(!is.na({{crime_change_var}}) & !is.na(gent_index) & !is.na(gent_index_lag1)) |>
    group_by(CODGEO) |>
    filter(n() >= 3) |>
    ungroup()
  
  granger_panel <- pdata.frame(granger_data, index = c("CODGEO", "year"))
  
  # Model 1: Does gentrification cause crime change?
  model_gent_causes_crime <- plm(
    as.formula(paste0(deparse(substitute(crime_change_var)), " ~ gent_index_lag1")),
    data = granger_panel,
    model = "within",
    effect = "twoways"
  )
  
  # Prepare data for reverse causality
  granger_data2 <- granger_data |>
    group_by(CODGEO) |>
    mutate(crime_change_lag1 = dplyr::lag({{crime_change_var}}, 1)) |>
    ungroup() |>
    filter(!is.na(crime_change_lag1))
  
  granger_panel2 <- pdata.frame(granger_data2, index = c("CODGEO", "year"))
  
  # Model 2: Does crime cause gentrification?
  model_crime_causes_gent <- plm(
    gent_index ~ crime_change_lag1,
    data = granger_panel2,
    model = "within",
    effect = "twoways"
  )
  
  # Extract results
  tibble(
    `Crime Type` = crime_type_name,
    Direction = c("Gentrification → Crime", "Crime → Gentrification"),
    Estimate = c(
      summary(model_gent_causes_crime)$coefficients[1, "Estimate"],
      summary(model_crime_causes_gent)$coefficients[1, "Estimate"]
    ),
    `P-value` = c(
      summary(model_gent_causes_crime)$coefficients[1, "Pr(>|t|)"],
      summary(model_crime_causes_gent)$coefficients[1, "Pr(>|t|)"]
    )
  )
}

# Prepare violent crimes data with lag
violent_granger <- violent_crimes_panel |>
  group_by(CODGEO) |>
  mutate(gent_index_lag1 = dplyr::lag(gent_index, 1)) |>
  ungroup()

# Prepare non-violent crimes data with lag
nonviolent_granger <- nonviolent_crimes_panel |>
  group_by(CODGEO) |>
  mutate(gent_index_lag1 = dplyr::lag(gent_index, 1)) |>
  ungroup()

# Prepare total crime data with lag
total_granger <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  mutate(total_crime_rate = total_crimes / insee_pop) |>
  left_join(gent_yearly |> select(CODGEO, year, gent_index), by = c("CODGEO", "year")) |>
  group_by(CODGEO) |>
  mutate(
    crime_change = log(total_crime_rate + 0.0001) - log(dplyr::lag(total_crime_rate) + 0.0001),
    gent_index_lag1 = dplyr::lag(gent_index, 1)
  ) |>
  ungroup()

# Run Granger tests for each crime type manually

# TOTAL CRIME
granger_total <- total_granger |>
  filter(!is.na(crime_change) & !is.na(gent_index) & !is.na(gent_index_lag1)) |>
  group_by(CODGEO) |>
  filter(n() >= 3) |>
  ungroup()

granger_panel_total <- pdata.frame(granger_total, index = c("CODGEO", "year"))

model_gent_crime_total <- plm(crime_change ~ gent_index_lag1,
                               data = granger_panel_total,
                               model = "within", effect = "twoways")

granger_total2 <- granger_total |>
  group_by(CODGEO) |>
  mutate(crime_change_lag1 = dplyr::lag(crime_change, 1)) |>
  ungroup() |>
  filter(!is.na(crime_change_lag1))

granger_panel_total2 <- pdata.frame(granger_total2, index = c("CODGEO", "year"))

model_crime_gent_total <- plm(gent_index ~ crime_change_lag1,
                               data = granger_panel_total2,
                               model = "within", effect = "twoways")

# VIOLENT CRIME
granger_violent <- violent_granger |>
  filter(!is.na(violent_change) & !is.na(gent_index) & !is.na(gent_index_lag1)) |>
  group_by(CODGEO) |>
  filter(n() >= 3) |>
  ungroup()

granger_panel_violent <- pdata.frame(granger_violent, index = c("CODGEO", "year"))

model_gent_crime_violent <- plm(violent_change ~ gent_index_lag1,
                                 data = granger_panel_violent,
                                 model = "within", effect = "twoways")

granger_violent2 <- granger_violent |>
  group_by(CODGEO) |>
  mutate(violent_change_lag1 = dplyr::lag(violent_change, 1)) |>
  ungroup() |>
  filter(!is.na(violent_change_lag1))

granger_panel_violent2 <- pdata.frame(granger_violent2, index = c("CODGEO", "year"))

model_crime_gent_violent <- plm(gent_index ~ violent_change_lag1,
                                 data = granger_panel_violent2,
                                 model = "within", effect = "twoways")

# NON-VIOLENT CRIME
granger_nonviolent <- nonviolent_granger |>
  filter(!is.na(nonviolent_change) & !is.na(gent_index) & !is.na(gent_index_lag1)) |>
  group_by(CODGEO) |>
  filter(n() >= 3) |>
  ungroup()

granger_panel_nonviolent <- pdata.frame(granger_nonviolent, index = c("CODGEO", "year"))

model_gent_crime_nonviolent <- plm(nonviolent_change ~ gent_index_lag1,
                                    data = granger_panel_nonviolent,
                                    model = "within", effect = "twoways")

granger_nonviolent2 <- granger_nonviolent |>
  group_by(CODGEO) |>
  mutate(nonviolent_change_lag1 = dplyr::lag(nonviolent_change, 1)) |>
  ungroup() |>
  filter(!is.na(nonviolent_change_lag1))

granger_panel_nonviolent2 <- pdata.frame(granger_nonviolent2, index = c("CODGEO", "year"))

model_crime_gent_nonviolent <- plm(gent_index ~ nonviolent_change_lag1,
                                    data = granger_panel_nonviolent2,
                                    model = "within", effect = "twoways")

# Create summary table
granger_results <- tibble(
  `Crime Type` = c("Total", "Total", "Violent", "Violent", "Non-violent", "Non-violent"),
  Direction = rep(c("Gentrification → Crime", "Crime → Gentrification"), 3),
  Estimate = c(
    summary(model_gent_crime_total)$coefficients[1, "Estimate"],
    summary(model_crime_gent_total)$coefficients[1, "Estimate"],
    summary(model_gent_crime_violent)$coefficients[1, "Estimate"],
    summary(model_crime_gent_violent)$coefficients[1, "Estimate"],
    summary(model_gent_crime_nonviolent)$coefficients[1, "Estimate"],
    summary(model_crime_gent_nonviolent)$coefficients[1, "Estimate"]
  ),
  `P-value` = c(
    summary(model_gent_crime_total)$coefficients[1, "Pr(>|t|)"],
    summary(model_crime_gent_total)$coefficients[1, "Pr(>|t|)"],
    summary(model_gent_crime_violent)$coefficients[1, "Pr(>|t|)"],
    summary(model_crime_gent_violent)$coefficients[1, "Pr(>|t|)"],
    summary(model_gent_crime_nonviolent)$coefficients[1, "Pr(>|t|)"],
    summary(model_crime_gent_nonviolent)$coefficients[1, "Pr(>|t|)"]
  )
) |>
  mutate(`P-value` = ifelse(`P-value` < 0.001, "< 0.001", round(`P-value`, 3)))

knitr::kable(granger_results, 
             digits = 3, 
             caption = "Granger Causality Tests by Crime Type")
```
After controlling for city fixed effects, year fixed effects, and department crime trends, neither own gentrification nor neighbor gentrification significantly predicts crime changes. This suggests that there is no evidence of short term spatial spillovers or crime displacement to neighboring areas once we account for common temporal shocks at the city level. However, this result has significant limitations and should not be interpreted as "gentrification has no spatial spillover effect" (see Conclusion)

## Spatial spillover effects

Does gentrification in neighboring cities affect local crime? To test this, we compute the average gentrification index of other cities within the same department (excluding the city itself) and examine whether this "neighbor gentrification" predicts local crime changes. We also control for department-level crime trends to ensure that spillover effects are not simply capturing regional dynamics.
```{r, echo=FALSE}
library(plm)

# Prepare data for each crime type
total_granger <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  mutate(total_crime_rate = total_crimes / insee_pop) |>
  left_join(gent_yearly |> select(CODGEO, year, gent_index), by = c("CODGEO", "year")) |>
  group_by(CODGEO) |>
  mutate(
    crime_change = log(total_crime_rate + 0.0001) - log(dplyr::lag(total_crime_rate) + 0.0001),
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    crime_change_lag1 = dplyr::lag(crime_change, 1)
  ) |>
  ungroup() |>
  filter(!is.na(crime_change) & !is.na(gent_index) & !is.na(gent_index_lag1)) |>
  group_by(CODGEO) |>
  filter(n() >= 3) |>
  ungroup()

violent_granger <- violent_crimes_panel |>
  group_by(CODGEO) |>
  mutate(
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    violent_change_lag1 = dplyr::lag(violent_change, 1)
  ) |>
  ungroup() |>
  filter(!is.na(violent_change) & !is.na(gent_index) & !is.na(gent_index_lag1)) |>
  group_by(CODGEO) |>
  filter(n() >= 3) |>
  ungroup()

nonviolent_granger <- nonviolent_crimes_panel |>
  group_by(CODGEO) |>
  mutate(
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    nonviolent_change_lag1 = dplyr::lag(nonviolent_change, 1)
  ) |>
  ungroup() |>
  filter(!is.na(nonviolent_change) & !is.na(gent_index) & !is.na(gent_index_lag1)) |>
  group_by(CODGEO) |>
  filter(n() >= 3) |>
  ungroup()

# Function to run Granger tests
run_granger <- function(data, crime_var, crime_lag_var) {
  panel <- pdata.frame(data, index = c("CODGEO", "year"))
  
  # Gentrification → Crime
  model1 <- plm(as.formula(paste(crime_var, "~ gent_index_lag1")),
                data = panel, model = "within", effect = "twoways")
  
  # Crime → Gentrification
  data2 <- data |> filter(!is.na(.data[[crime_lag_var]]))
  panel2 <- pdata.frame(data2, index = c("CODGEO", "year"))
  model2 <- plm(as.formula(paste("gent_index ~", crime_lag_var)),
                data = panel2, model = "within", effect = "twoways")
  
  c(summary(model1)$coefficients[1, "Estimate"],
    summary(model1)$coefficients[1, "Pr(>|t|)"],
    summary(model2)$coefficients[1, "Estimate"],
    summary(model2)$coefficients[1, "Pr(>|t|)"])
}

# Run tests
total_res <- run_granger(total_granger, "crime_change", "crime_change_lag1")
violent_res <- run_granger(violent_granger, "violent_change", "violent_change_lag1")
nonviolent_res <- run_granger(nonviolent_granger, "nonviolent_change", "nonviolent_change_lag1")

# Create results table
granger_results <- tibble(
  `Crime Type` = rep(c("Total", "Violent", "Non-violent"), each = 2),
  Direction = rep(c("Gentrification → Crime", "Crime → Gentrification"), 3),
  Estimate = c(total_res[1], total_res[3], violent_res[1], violent_res[3], nonviolent_res[1], nonviolent_res[3]),
  `P-value` = c(total_res[2], total_res[4], violent_res[2], violent_res[4], nonviolent_res[2], nonviolent_res[4])
) |>
  mutate(`P-value` = ifelse(`P-value` < 0.001, "< 0.001", round(`P-value`, 3)))

knitr::kable(granger_results, digits = 3, caption = "Granger Causality Tests by Crime Type")
```

After controlling for city fixed effects, year fixed effects, and department crime trends, neither own gentrification nor neighbor gentrification significantly predicts crime changes. This suggests that there is no evidence of spatial spillovers or crime displacement to neighboring areas once we account for common temporal shocks at the city level. This result however has huge limitations and shouldn't be interpreted as "gentrification doesn't have any spatial spillover effect" (see conclusion).

# Conclusion

Based on the gentrification index we built, the data analysis reveals several 
statistically significant relationships between gentrification and crime rate 
evolution. 

## Cross-sectionnal findings :

Cities with higher gentrification intensity 
experienced more moderate changes in both violent and non-violent crime rates 
between 2016 and 2021. Specifically, violent crimes increased everywhere during 
this period, but the increase was substantially smaller in highly gentrified 
cities (+5%) compared to low-gentrification cities (+24%). This difference is 
statistically significant (p < 0.001). For non-violent crimes, no significant 
differences emerged across gentrification categories.

## Panel analysis findings :

The year-over-year analysis reveals a modest but 
significant crime-reducing effect of gentrification at the 1-year lag, 
particularly for total crime (β = -0.019, p = 0.022) and violent crime 
(β = -0.027, p = 0.020). Interestingly, the multiple-lag analysis shows an 
oscillating pattern: gentrification initially reduces crime rates, but this 
effect partially reverses at the 2-year lag before returning to a negative 
relationship at 3 years. This suggests that gentrification's impact on crime 
is not permanent and may involve complex adjustment dynamics.

## Causality : 

Granger causality tests suggest a bidirectional relationship: 
gentrification predicts subsequent reductions in crime, while crime reductions 
also predict subsequent gentrification. This mutual reinforcement complicates 
causal interpretation.

## Spatial spillovers :

After controlling for city and year fixed effects, we 
found no evidence of crime displacement to neighboring areas. However, this 
null result should be interpreted cautiously given the coarse spatial 
definition (department-level neighbors) and limited temporal variation.

# Limitations :

The study faces several constraints. First, the 6-year panel 
(2016-2021) limits our ability to detect longer-term effects and includes the 
anomalous COVID-19 period (2020). Second, data coverage excludes important 
areas (Alsace, Moselle) and lacks arrondissement-level detail for major cities 
(Paris, Marseille, Lyon). Third, police-recorded crime data may reflect 
reporting behavior and enforcement intensity rather than true crime levels—a 
concern particularly relevant for domestic violence, the only crime type 
showing significant cross-sectional differences. Finally, the gentrification 
index could be enriched 
with additional indicators such as educational attainment, occupation 
structure, or housing tenure changes.

## Policy implications : 

Despite these limitations, the findings suggest that 
gentrification is associated with a stabilizing effect on crime rates, 
dampening both increases in violent crime and decreases in non-violent crime. 
However, the modest effect sizes (R² < 1%) indicate that gentrification 
explains only a small fraction of crime variation, and policymakers should 
not expect neighborhood upgrading alone to substantially reduce crime.

## References

Ruth Glass (1964): https://www.tandfonline.com/doi/full/10.1080/03058034.2024.2425593 

Neil Smith (1979): https://miguelangelmartinez.net/IMG/pdf/1979_smith_gentrification_japa.pdf