---
title: "Noe Leo Final"
author: "Léo-Paul Garelli and Noé Mansuy"
format: html
editor: 
  markdown: 
    wrap: 72
---

Github project : https://github.com/LEOPAUL55/Noe_Leo_final.git

```{r}
#| message: false
here::i_am("Noe_Leo_Final.Rproj")
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(vroom)
library(data.table)
library(readr)
library(stringr)
```

# Data Loading and Cleaning

## population 2018

Source : https://www.insee.fr/fr/statistiques/5650720

This is the 2018 population by IRIS code. Made by the INSEE.

Because the 2018 income data is available by IRIS code rather than by
GEO code, we need to weight each IRIS-level mean by its 2018 population
in order to compute the correct city-level values. Since the 2018
population data at the IRIS level is available for only 994 of the
cities included in the other years, we will have to restrict our
analysis to these 994 cities.

```{r}
pop2018_IRIS <- read_delim(here("data","pop2018.csv"),
    na = "s") |>
  select(IRIS,P18_POP)
```

## Real estate prices

Source :
https://www.data.gouv.fr/datasets/indicateurs-immobiliers-par-commune-et-par-annee-prix-et-volumes-sur-la-periode-2014-2024

This is a collection of various variables on residential housing market
at the city level between 2014 and 2024. What is interesting for us is
the mean price by square meter of the goods that have been sold each
year. This was published by Boris Mericksay

Real‑estate price data are provided as separate yearly CSV files. To
improve efficiency and avoid redundant code, we use vroom’s multi‑file
loading functionality to read all files simultaneously. The year is
extracted from the file name, and city identifiers are harmonized to
allow consistent merging with other datasets.

```{r}
ppsm <- vroom(
  list.files(
    here("data", "price per square meter"),
    pattern = "\\.csv$",
    full.names = TRUE
  ),
  col_select = c(INSEE_COM, Prixm2Moyen),
  id = "source"
) |>
  transmute(
    CODGEO = as.character(INSEE_COM),
    year   = as.integer(str_extract(source, "\\d{4}")),
    Prixm2Moyen
  )
```

We need to add the information for Paris by arrondissements that are
missing in the dataset above. To do so, we will use the data available
at https://paris.notaires.fr/fr/carte-des-prix?utm_source=chatgpt.com.

For simplicity, we only focus on the selling price of appartements in
the city from 2016 to 2021 (trimester 4). The available file is a PDF
that can be easily convert into a CSV file.

```{r}
ppsm_paris <- vroom(
  here("data", "dvfparis.csv"),
  delim = ",",
  quote = "\"",
  col_types = cols()  # infer types automatically
)|>
  mutate(
    year    = as.integer(str_extract(Trimestre, "\\d{4}")),
    quarter = as.integer(str_extract(Trimestre, "(?<=T)\\d"))
  ) |>
  filter(quarter == 4, year >= 2016, year <= 2021) |>
  pivot_longer(
    cols = -c(Trimestre, quarter, year, Centre),
    names_to = "arrondissement",
    values_to = "Prixm2Moyen"
  ) |>
  transmute(
    CODGEO = as.character(75100 + as.integer(str_remove(arrondissement, "er|e"))),
    year,
    Prixm2Moyen = as.numeric(str_remove_all(Prixm2Moyen, "[ €]"))
  ) |>
  arrange(year, CODGEO)

```

After that, we just need to merge the two datasets to obtain clean data
about the real estate price in France.

```{r}
ppsm <- bind_rows(ppsm, ppsm_paris) |>
  arrange(year, CODGEO)
```

## Poverty

sources :

-   2016, 2017, 2018 :
    https://www.data.gouv.fr/datasets/revenus-et-pauvrete-des-menages-aux-niveaux-national-et-local-revenus-localises-sociaux-et-fiscaux/

-   2019 : https://www.insee.fr/fr/statistiques/6036902?sommaire=6036904

-   2020 : https://www.insee.fr/fr/statistiques/6692392?sommaire=6692394

-   2021 : https://www.insee.fr/fr/statistiques/7756729?sommaire=7756859

These are collections of indicators of disposable income in France, at
the city level, between 2016 and 2021. These are made by the Insee.

We will only focus on the poverty rate, the median standard of living,
the first and the ninth deciles.

Poverty data are provided as separate yearly files with heterogeneous
structures. Variable names differ across years, and the 2018 data are
reported at the IRIS level rather than the commune level. To handle
this, we implement a loading strategy that reads each file separately
while restricting imports to the variables of interest (years
2016:2021). Poverty indicators are harmonized across years. Non‑numeric
values are explicitly converted to missing values during loading, but it
won't affect our analysis because this only concerns small cities. The
result is a consistent panel of indicators comparable across years.

```{r}
poverty_files <- list.files(
  here("data", "poverty"),
  pattern = "\\.csv$",
  full.names = TRUE
)

load_poverty <- function(file) {

 year <- as.integer(str_extract(basename(file), "\\d{4}"))


  if (year %in% c(2016, 2017, 2019, 2020, 2021)) {

    out <- vroom(
      file,
      delim = ";",
      col_select = any_of(c(
        "CODGEO",
        "MED16", "MED17", "MED19", "MED20", "MED21",
        "TP6016", "TP6017", "TP6019", "TP6020","TP6021",
        "D116", "D117", "D119", "D120","D121",
        "D916", "D917", "D919", "D920","D921"
      ))
    ) |>
      transmute(
        CODGEO,
        MED  = Reduce(coalesce, pick(any_of(c("MED16", "MED17", "MED19", "MED20","MED21")))),
        TP60 = Reduce(coalesce, pick(any_of(c("TP6016", "TP6017", "TP6019", "TP6020","TP6021")))),
        D1   = Reduce(coalesce, pick(any_of(c("D116", "D117", "D119", "D120","D121")))),
        D9   = Reduce(coalesce, pick(any_of(c("D916", "D917", "D919", "D920","D921")))),
        year
      )

  } else if (year == 2018) {

    out <- vroom(
      file,
      delim = ";",
      col_select = any_of(c(
        "IRIS",
        "DISP_MED18",
        "DISP_TP6018",
        "DISP_D118",
        "DISP_D918"
      ))
    ) |>
      transmute(
        IRIS,
        MED  = DISP_MED18,
        TP60 = DISP_TP6018,
        D1   = DISP_D118,
        D9   = DISP_D918,
        year
      )

  }
 

  
  out |>
    mutate(across(
  any_of(c("MED", "TP60", "D1", "D9")),
  ~ suppressWarnings(as.numeric(gsub(",", ".", .x)))
))
}

poverty <- bind_rows(
  lapply(poverty_files, load_poverty)
)

## Aggregate 2018 IRIS Data to City Level

poverty_2018 <- poverty |>
  filter(year == 2018) |>
  left_join(pop2018_IRIS, by = "IRIS") |>
  mutate(CODGEO = substr(IRIS, 1, 5)) |>
  group_by(CODGEO) |>
  summarise(
    MED = weighted.mean(MED, P18_POP, na.rm = TRUE),
    TP60 = weighted.mean(TP60, P18_POP, na.rm = TRUE),
    D1 = weighted.mean(D1, P18_POP, na.rm = TRUE),
    D9 = weighted.mean(D9, P18_POP, na.rm = TRUE),
    year = 2018,
    .groups = "drop"
  )

codgeo_2018 <- poverty_2018 |> pull(CODGEO)

poverty <- bind_rows(
  poverty |> filter(year != 2018),
  poverty_2018
) |>
  filter(CODGEO %in% codgeo_2018)

## Crime Data

crime <- vroom(
  here("data_criminality.csv"),
  delim = ";",
  locale = locale(decimal_mark = ","),
  col_select = c(
    annee,
    CODGEO_2025,
    indicateur,
    nombre,
    insee_pop
  ),
  col_types = cols(
    annee = col_integer(),
    nombre = col_integer(),
    insee_pop = col_integer()
  )
) |>
  pivot_wider(
    names_from = indicateur,
    values_from = nombre,
    values_fill = 0
  ) |>
  rename(CODGEO = CODGEO_2025) |>
  mutate(
    total_crimes = rowSums(
      across(-c(CODGEO, annee, insee_pop)),
      na.rm = TRUE
    ),
    year = annee
  ) |>
  select(-annee)

## City Names

city_names <- vroom(
  here("data", "poverty", "pov2016.csv"),
  delim = ";",
  col_select = c(CODGEO, LIBGEO)
) |>
  distinct()

## Create Final Dataset

data_final <- poverty |>
  left_join(crime, by = c("CODGEO", "year")) |>
  left_join(ppsm, by = c("CODGEO", "year")) |>
  select(-IRIS) |>
  left_join(city_names, by = "CODGEO") |>
  distinct(CODGEO, year, .keep_all = TRUE) |>
  group_by(CODGEO) |>
  filter(!is.na(Prixm2Moyen)) |>
  filter(n_distinct(year) == 6) |>
  ungroup() |>
  filter(insee_pop >= MIN_POPULATION) |>
  mutate(
    crime_rate = (total_crimes / insee_pop) * 100,
    covid_period = year >= 2020,
    department = substr(CODGEO, 1, 2)
  )

# Create violent and non-violent crime categories
data_final <- data_final |>
  mutate(
    violent_crimes = rowSums(
      across(c(
        `Violences physiques intrafamiliales`,
        `Violences physiques hors cadre familial`,
        `Vols avec armes`,
        `Vols violents sans arme`
      )),
      na.rm = TRUE
    ),
    nonviolent_crimes = rowSums(
      across(c(
        `Vols sans violence contre des personnes`,
        `Vols de véhicule`,
        `Vols dans les véhicules`,
        `Vols d'accessoires sur véhicules`,
        `Cambriolages de logement`,
        `Destructions et dégradations volontaires`,
        `Escroqueries et fraudes aux moyens de paiement`,
        `Usage de stupéfiants`,
        `Trafic de stupéfiants`
      )),
      na.rm = TRUE
    ),
    violent_rate = (violent_crimes / insee_pop) * 100,
    nonviolent_rate = (nonviolent_crimes / insee_pop) * 100
  )

cat("Final dataset:", n_distinct(data_final$CODGEO), "cities over", 
    n_distinct(data_final$year), "years\n")
```



## Delinquency

Source:
https://www.data.gouv.fr/datasets/bases-statistiques-communale-departementale-et-regionale-de-la-delinquance-enregistree-par-la-police-et-la-gendarmerie-nationales

Here is the loading of the database for the delinquency. This is a
database of all the crimes and offences (except traffic offences) at the
city level in France, published by French Police Force, since 2016.

The CSV format of the file creates problems of fromat that must be
solved (types of column, delimiotation...).

The type of crimes must be reshape to long format (in rows) to wide
format (in columns).

The total number of crimes per year per city is computed by doing the
sum of each type of crimes.

```{r}
crime <- vroom(
  here("data_criminality.csv"),
  delim = ";",
  locale = locale(decimal_mark = ","),
  col_select = c(
    annee,
    CODGEO_2025,
    indicateur,
    nombre,
    insee_pop
  ),
  col_types = cols(
    annee = col_integer(),
    nombre = col_integer(),
    insee_pop = col_integer()
  )
)|>
  pivot_wider(
    names_from = indicateur,
    values_from = nombre,
    values_fill = 0
  ) |>
  rename(CODGEO = CODGEO_2025)|>
  mutate(
    total_crimes = rowSums(
      across(-c(CODGEO, annee, insee_pop)),
      na.rm = TRUE
    ),
    year = annee
  ) |>
  select(-annee)
```

## Selecting cities present in the 2018 data

The analysis will only focus on the cities where all the informations
are available on 2018, because of the IRIS nomenclature.

The period that we will study is from 2016 to 2021. We are adding in the
dataset the cities' name as well as removing the ones for which we don't
have the information of the real estate price in at least one year.

We also compute the crime rate.

```{r}
city_names <- vroom(
  here("data", "poverty", "pov2016.csv"),
  delim = ";",
  col_select = c(CODGEO, LIBGEO)
) |>
  distinct()

data_final <- poverty |>
  left_join(crime, by = c("CODGEO", "year")) |>
  left_join(ppsm, by = c("CODGEO", "year"))|>
  select(-IRIS)|>
  left_join(city_names, by = "CODGEO")|>
  distinct(CODGEO, year, .keep_all = TRUE) |>
  group_by(CODGEO) |>
  filter(!is.na(Prixm2Moyen))|>
  filter(n_distinct(year) == 6) |>
  ungroup() |>
  mutate(
    crime_rate = (total_crimes/insee_pop)*100
  )
```

We obtained a clean dataset for 897 cities between 2016 and 2021.

# Cleaned data tables : summary and most important variables

After loading the datasets, we needed to convert the IRIS code into the
GEOCODE by weighting the poverty rate of each geocode by the population
of the corresponding IRIS code. By doing that, we obtained clean data
for 994 cities. We only focus on these cities in the analysis from 2016
to 2021. We also computed the poverty rate for each city.

We only kept the variables of interest by loading the datasets, that are
the poverty rate, the first and last decile, the median income, the
number of crimes per crime type and the real estate price (average price
per square meter).Then we merged them into an only dataset that combines
all the data per cities per years.

```{r}
descriptions <- c(
  "Unique code that corresponds to one city. The first two numbers correspond to a department",
  "The city's population",
  "Number of domestic physical violence reported for the year",
  "Number of non domestic physical violence reported for the year",
  "Number of sexual violence reported for the year",
  "Number of armed robberies reported for the year",
  "Number of non armed violent robberies reported for the year",
  "Number of non violent robberies against persons reported for the year",
  "Number of home invasions reported for the year",
  "Number of vehicle theft reported for the year",
  "Number of theft from vehicles reported for the year",
  "Number of theft of vehicle accessories reported for the year",
  "Number of willful damage and vandalism reported for the year",
  "Number of reported drug use for the year",
  "Number of illegal drug use sanctioned by fixed fine reported for the year",
  "Number of drug dealing cases reported for the year",
  "Number of fraud and payment method fraud reported for the year",
  "Total number of crimes per cities reported for this year",
  "The concerned year",
  "Median income for this year",
  "Poverty rate (set at less than 60% of the median income)",
  "1st income decile",
  "9th income decile",
  "Price per square meter of the real estate goods sold for the year", 
  "The city's name"
)

summary <- data.table(
  Variable = names(data_final),
  Type = sapply(data_final, class),
  Description = descriptions
)

knitr::kable(summary)

dataset_info <- data.table(
  Info = c("Number of rows", "Number of columns"),
  Value = c(nrow(data_final), ncol(data_final)) 
)

knitr::kable(dataset_info)

```

# Research question :

Does the acceleration of the gentrification process in certain French
cities with more than 5000 inhabitants (2016-2021) correlate with a
decrease in recorded crime in these same municipalities, and is there a
displacement effect towards non-gentrified cities?

The first goal will be to make a gentrification index per city with a
few variables: square meter price growth, median wage growth, poverty
rate growth.

Then we will select the cities with the highest gentrification index
(say top 20%) and compare it to the rest of the cities : do we observe a
declination in terms of delinquency between the top 20% and the others?
If yes, do we observe a spatial displacement of crime to non-gentrifying
areas?

## Real estate price

```{r}
library(geodata)
library(sf)
```

```{r}
city <- st_read(here("data","admin_city.gpkg"),layer = "COMMUNE")|>
  rename(
    CODGEO = code_insee,
    geometry = geometrie
  )

city$CODGEO <- sprintf("%05s", city$CODGEO)
```

```{r}
ppsm_variation <- ppsm |>
  filter(CODGEO %in% data_final$CODGEO)|>
  group_by(CODGEO) |>
  summarise(
    Prixm2_start = Prixm2Moyen[year == min(year)],  
    Prixm2_end   = Prixm2Moyen[year == max(year)],  
    .groups = "drop"
  ) |>
  mutate(
    Prixm2_diff = Prixm2_end - Prixm2_start,                     
    Prixm2_pct  = ((Prixm2_end / Prixm2_start) - 1) * 100          
  )|>
  left_join(city_names, by = "CODGEO")|>
  arrange(desc(Prixm2_pct))

ppsm_variation_map <- city |> 
  left_join(ppsm_variation, by = "CODGEO")
```

# Analysis of the crime rate evolution from 2016 to 2021 - comparison between violent and non violent crimes

```{r}
crime_wide <- data_final |>
  mutate(
    violent_total = rowSums(
      across(c(`Violences physiques intrafamiliales`,
               `Violences physiques hors cadre familial`,
               `Vols avec armes`,
               `Vols violents sans arme`)),
      na.rm = TRUE
    ),
    nonviolent_total = rowSums(
      across(c(`Vols sans violence contre des personnes`,
               `Vols de véhicule`,
               `Vols dans les véhicules`,
               `Vols d'accessoires sur véhicules`,
               `Cambriolages de logement`,
               `Destructions et dégradations volontaires`,
               `Escroqueries et fraudes aux moyens de paiement`,
               `Usage de stupéfiants`,
               `Trafic de stupéfiants`)),
      na.rm = TRUE
    ),
    total_excl_sexual = total_crimes - `Violences sexuelles`,
    violent_rate = (violent_total / insee_pop) * 100,
    nonviolent_rate = (nonviolent_total / insee_pop) * 100,
    total_rate_excl = (total_excl_sexual / insee_pop) * 100
  ) |>
  group_by(year) |>
  summarise(
    total = median(total_rate_excl, na.rm = TRUE),
    violent = median(violent_rate, na.rm = TRUE),
    nonviolent = median(nonviolent_rate, na.rm = TRUE),
    .groups = "drop"
  )

# Scale factor for secondary axis
scale_factor <- max(crime_wide$nonviolent) / max(crime_wide$violent)

ggplot(crime_wide, aes(x = year)) +
  geom_line(aes(y = nonviolent, color = "Non-Violent"), size = 1.5) +
  geom_point(aes(y = nonviolent, color = "Non-Violent"), size = 4) +
  geom_line(aes(y = total, color = "Total"), size = 1.5) +
  geom_point(aes(y = total, color = "Total"), size = 4) +
  geom_line(aes(y = violent * scale_factor, color = "Violent"), size = 1.5) +
  geom_point(aes(y = violent * scale_factor, color = "Violent"), size = 4) +
  scale_y_continuous(
    name = "Non-Violent & Total (per 100 inh.)",
    sec.axis = sec_axis(~./scale_factor, name = "Violent Crime (per 100 inh.)")
  ) +
  scale_x_continuous(breaks = 2016:2021) +
  scale_color_manual(values = c("Total" = "#2c3e50", 
                                "Violent" = "#e74c3c", 
                                "Non-Violent" = "#3498db")) +
  labs(
    x = "Year",
    color = NULL,
    title = "Crime Rate Evolution in French Cities (2016-2021)",
    subtitle = "Note: Violent crime uses secondary axis (right) | Excluding sexual violence"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.position = "bottom",
    axis.title.y.right = element_text(color = "#e74c3c"),
    axis.text.y.right = element_text(color = "#e74c3c")
  )
```


# Building a non binary gentrification index between first year and last year and checking for correlation with change in crime rate.

## Building a non binary gentrification index between first year and last year :

Identifying areas that could gentrify :

```{r}
#Taking the earliest year of data for each demographic area (CODGEO)
baseline <- data_final |>
  group_by(CODGEO) |>
  filter(year == min(year)) |>
  #capturing the median income as the starting point
  ungroup() |>
  select(CODGEO, baseline_med = MED)

#Making sure only areas below the "60% of median income" threshold are considered "gentrifiable" (by definition, only lower income areas can gentrify)
percentile_cutoff <- quantile(baseline$baseline_med, 0.60, na.rm = TRUE)
```

Calculating changes over time :

```{r}
gentrifiable_areas <- baseline |>
  filter(baseline_med < percentile_cutoff) |>
  pull(CODGEO)

# Calculating percentage changes from first to last year
evolution <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>  
  group_by(CODGEO) |>
  summarise(
    price_var   = ((last(Prixm2Moyen) / first(Prixm2Moyen)) - 1) * 100,
    med_inc_var = (last(MED) / first(MED) - 1) * 100,
    d9_var      = (last(D9) / first(D9) - 1) * 100,
    d1_var      = (last(D1) / first(D1) - 1) * 100,
    pov_var     = last(TP60) - first(TP60),
    .groups = "drop"
  ) |>
  left_join(city_names, by = "CODGEO")

# Flipping signs for interpretation : making increases in low incomes and decrease in poverty show as positive values: this aligns indicators so higher values = more gentrification
evolution <- evolution |>
  mutate(
    pov_var = -pov_var
  )
```

Standardizing the variables :

```{r}
# Converting all the variables to z-score so it makes the different metrics comparable despite different scales 
evolution_std <- evolution |>
  mutate(across(
    c(price_var, med_inc_var, d9_var, d1_var, pov_var),
    scale
  ))
```

Creating a weighted gentrification index

```{r}
evolution_std <- evolution_std |>
  mutate(
    gentrification_index = 
      (price_var * 0.25) +        # Emphasize housing price
      (med_inc_var * 0.2) +     # Median income 
      (d9_var * 0.1) +          # Top income 
      (d1_var * 0.35) +          # Bottom income
      (pov_var * 0.1)           # Poverty reduction 
  )


# Sorting areas by gentrification intensity
evolution_std <- evolution_std |>
  arrange(desc(gentrification_index))
evolution_std_A <- evolution_std

# Displaying most gentrified areas
evolution_std_A |>
  head(10) |>
  select(CODGEO, LIBGEO, gentrification_index)

# Checking the validity of the index using Cronbach's alpha 
library(psych)
psych::alpha(evolution_std |> select(price_var, med_inc_var, d9_var, d1_var, pov_var), check.keys = TRUE)

```

-\> The Cronbach's alpha test tells us that the indicator we used is
valid and reliable, and that all indicators contribute meaningfully

## Checking for correlations with crime

Creating gentrification categories :

```{r}
evolution_std_A_groups <- evolution_std_A |>
  mutate(
    gent_category = case_when(
      gentrification_index > quantile(gentrification_index, 0.75, na.rm = TRUE) ~ "High",
      gentrification_index > quantile(gentrification_index, 0.25, na.rm = TRUE) ~ "Medium",
      TRUE ~ "Low"
    ),
    gent_category = factor(gent_category, levels = c("Low", "Medium", "High"))
  )

table(evolution_std_A_groups$gent_category)
```

Low = bottom 25% gentrified cities Medium = middle 50% gentrified cities
high = top 25% gentrified cities (among the "gentrifiable" cities)

Calculating crime rate changes

```{r}
#Restructuring data to have 2016 and 2021 side by side for comparison
crime_rate_change_continuous <- data_final |>
  filter(year %in% c(2016, 2021)) |>
  select(CODGEO, year, total_crimes, insee_pop) |>
  pivot_wider(
    names_from = year,
    values_from = c(total_crimes, insee_pop),
    names_sep = "_"
  ) |>
  mutate(
    # Calculating crime rates per capita for both years
    crime_rate_2016 = total_crimes_2016 / insee_pop_2016,
    crime_rate_2021 = total_crimes_2021 / insee_pop_2021,
    # Computing percentage change in crime rate over the 5-year period
    crime_rate_change = (crime_rate_2021 - crime_rate_2016) / crime_rate_2016
  ) |>
  
  # Merging datasets : combining crime data with gentrification categories - each area has now its crime change and gentrification level 
  
  left_join(
    evolution_std_A_groups |> select(CODGEO, gentrification_index, gent_category),
    by = "CODGEO"
  ) |>
  filter(!is.na(gent_category))

```

Scatter plot of the crime rate change compared to the gentrification
index

```{r}
library(viridis)

ggplot(crime_rate_change_continuous,
       aes(x = gentrification_index, y = crime_rate_change)) +
  geom_point(aes(color = gentrification_index), 
             alpha = 0.6, size = 2.5) +
  scale_color_viridis_c(option = "plasma", name = "Gentrification\nIntensity") +
  geom_smooth(method = "lm", color = "#FF6B6B", size = 1.2, 
              fill = "#FF6B6B", alpha = 0.15) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.3, color = "gray40") +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.3, color = "gray40") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_cartesian(ylim = c(-0.5, 0.5)) +
  labs(
    x = "Gentrification Index (standardized)",
    y = "Crime Rate Change (2016-2021)",
    title = "Small Link Between Gentrification and Crime Changes",
    subtitle = "Analysis of initially lower-income areas (bottom 60% baseline income)",
    caption = "Each point represents one geographic area | Red line shows linear trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0),
    plot.subtitle = element_text(size = 12, color = "gray40", hjust = 0),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )
```
-> Visually, it seems to exist a slight correlation between gentrification intensity and crime rate change between 2016 and 2021 among gentrifiable cities

Plotting the crime rate change among three arbitrary gentrification levels :

```{r}
# Calculate confidence intervals
ci_stats <- crime_rate_change_continuous |>
  group_by(gent_category) |>
  summarise(
    mean = mean(crime_rate_change, na.rm = TRUE),
    ci_lower = mean - 1.96 * sd(crime_rate_change, na.rm = TRUE) / sqrt(n()),
    ci_upper = mean + 1.96 * sd(crime_rate_change, na.rm = TRUE) / sqrt(n()),
    median = median(crime_rate_change, na.rm = TRUE)
  )

ggplot(ci_stats |> filter(gent_category %in% c("Low", "High", "Medium")),
       aes(x = gent_category, y = mean, color = gent_category)) +
  # CI lines
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0, size = 2, alpha = 0.4) +
  # Mean points
  geom_point(size = 8) +
  # Median points
  geom_point(aes(y = median), shape = 1, size = 8, stroke = 1.5) +
  scale_color_manual(values = c("Low" = "#3498db", "Medium" = "#f39c12", "High" = "#e74c3c")) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.3) +
  coord_cartesian(ylim = c(-0.1, 0.1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    x = NULL,
    y = "Crime Rate Change (2016–2021)",
    title = "Mean Crime Rate Changes Across Gentrification Levels",
    subtitle = "Solid dots = mean | Hollow circles = median | Lines = 95% CI",
    
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    axis.text.x = element_text(size = 14, face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```
-> There is a diminution in crime rate between all three gentrification levels

Crime change by type and gentrification intensity

```{r}

# Reshaping crime data per type
crime_change_type_continuous <- data_final |>
  pivot_longer(
    cols = 8:22,
    names_to = "crime_type",
    values_to = "crime_count"
  ) |>
  filter(year %in% c(2016, 2021)) |>
  select(CODGEO, year, crime_type, crime_count, insee_pop) |>
  
  # Creating year-over-year comparison
  pivot_wider(
    names_from = year,
    values_from = c(crime_count, insee_pop),
    names_sep = "_"
  ) |>
  # Calculating poopulation-adjusted change
  mutate(
    pop_adj_rate_change = if_else(
      crime_count_2016 > 0 & insee_pop_2016 > 0,
      ((crime_count_2021 / crime_count_2016) /
       (insee_pop_2021 / insee_pop_2016)) - 1,
      NA_real_
    )
  ) |>
  # Merging with gentrification levels
  left_join(
    evolution_std_A_groups |> select(CODGEO, gentrification_index, gent_category),
    by = "CODGEO"
  ) |>
  filter(!is.na(gent_category))

# Summary by crime type and gentrification level
crime_change_type_summary <- crime_change_type_continuous |>
  filter(is.finite(pop_adj_rate_change)) |>
  group_by(crime_type, gent_category) |>
  summarise(
    median_change = median(pop_adj_rate_change, na.rm = TRUE),
    n_cities = n(),
    .groups = "drop"
  )

# Plot
ggplot(crime_change_type_summary,
       aes(x = crime_type, y = median_change, fill = gent_category)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("Low" = "steelblue", "Medium" = "orange", "High" = "red")) +
  labs(
    x = "Type of Crime",
    y = "Population-Adjusted Crime Change (2016–2021)",
    fill = "Gentrification\nIntensity",
    title = "Crime Change by Type and Gentrification Intensity"
  ) +
  theme_minimal()
```

Creating crime categories (excluding sexual violences) :

```{r}

# Creating crime categories 
crime_change_type_cat <- crime_change_type_continuous |>
  mutate(
    crime_category = case_when(
      crime_type %in% c(
       
        "Violences physiques intrafamiliales",
        "Violences physiques hors cadre familial",
        "Vols avec armes",
        "Vols violents sans arme"
      ) ~ "Violent crimes",
      crime_type %in% c(
        "Vols sans violence contre des personnes",
        "Vols de véhicule",
        "Vols dans les véhicules",
        "Vols d'accessoires sur véhicules",
        "Cambriolages de logement",
        "Destructions et dégradations volontaires",
        "Escroqueries et fraudes aux moyens de paiement",
        "Usage de stupéfiants",
        "Trafic de stupéfiants"
      ) ~ "Non violent crimes",
      TRUE ~ NA_character_
    )
  )
```

Summarizing by category :

```{r}

# Calculating the median population adjusted change for each combination of crime category and gentrification level
crime_cat_summary <- crime_change_type_cat |>
  filter(is.finite(pop_adj_rate_change), !is.na(crime_category)) |>
  group_by(crime_category, gent_category) |>
  summarise(
    median_change = median(pop_adj_rate_change, na.rm = TRUE),
    n_obs = n(),
    .groups = "drop"
  )
```

Violent vs non violent crime change by gentrification intensity :

```{r}
# Plot
ggplot(crime_cat_summary,
       aes(x = crime_category, y = median_change, fill = gent_category)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Low" = "steelblue", "Medium" = "orange", "High" = "red")) +
  labs(
    x = "Crime Category",
    y = "Population-Adjusted Crime Change (2016–2021)",
    fill = "Gentrification\nIntensity",
    title = "Violent vs Non-Violent Crime Change\nby Gentrification Intensity"
  ) +
  theme_minimal()

# Table
knitr::kable(
  crime_cat_summary,
  digits = 3,
  col.names = c("Crime Category", "Gentrification Level", "Mean Change", "N Observations")
)
```

-\> while non violent crimes rate decreases and violent crime rate
increases for all groups, violent crimes increases less as
gentrification intensifies while non violent crimes diminish roughly the same way for all groups

Let's check if the changes are significant :


```{r}
# Tests whether crime rate changes differ significantly across the three gentrification levels

anova_result <- aov(crime_rate_change ~ gent_category, data = crime_rate_change_continuous)
summary(anova_result)

# Tests wether there's a linear relationship between gentrification intensity and crime change 
cor.test(
  crime_rate_change_continuous$gentrification_index,
  crime_rate_change_continuous$crime_rate_change,
  use = "complete.obs"
)
```

-\> nothing is significant

```{r}
# Checking only for violent crimes

violent_data <- crime_change_type_cat |>
  filter(crime_category == "Violent crimes", is.finite(pop_adj_rate_change)) |>
  group_by(CODGEO, gent_category) |>
  summarise(median_change = median(pop_adj_rate_change), .groups = "drop")  # <- median here

summary(aov(median_change ~ gent_category, data = violent_data))
```

-\> not significant

Plotting violent crime growth among gentrification intensities
categories :

```{r}
# Get the medians for the plot
violent_summary <- violent_data |>
  group_by(gent_category) |>
  summarise(
    median_change = median(median_change, na.rm = TRUE),
    # For error bars around median, we can use bootstrapped confidence intervals
    # or use IQR/quantiles instead of SE
    q25 = quantile(median_change, 0.25, na.rm = TRUE),
    q75 = quantile(median_change, 0.75, na.rm = TRUE),
    n = n()
  )

# Plot with median and IQR bars
ggplot(violent_summary, aes(x = gent_category, y = median_change, fill = gent_category)) +
  geom_col(alpha = 0.8) +
  # Use IQR as error bars (shows middle 50% of data)
  geom_errorbar(aes(ymin = q25, ymax = q75), 
                width = 0.2, linewidth = 0.8) +
  scale_fill_manual(values = c("Low" = "steelblue", "Medium" = "orange", "High" = "red")) +
  labs(
    x = "Gentrification Intensity",
    y = "Median Violent Crime Change (Pop-Adjusted)",
    caption = "Error bars show interquartile range (25th-75th percentiles)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Overall test: Are there differences between groups?

```{r}
library(ggsignif)

kruskal_test <- kruskal.test(median_change ~ gent_category, data = violent_data)
print(kruskal_test)

library(FSA)

# Dunn's test with Bonferroni correction
dunn_test <- dunnTest(median_change ~ gent_category, 
                      data = violent_data,
                      method = "bonferroni")
print(dunn_test)

# Or pairwise Wilcoxon tests
pairwise_results <- pairwise.wilcox.test(
  violent_data$median_change,
  violent_data$gent_category,
  p.adjust.method = "bonferroni"
)
print(pairwise_results)

# Epsilon-squared effect size
n_total <- nrow(violent_data)
chi_squared <- 6.9672
df <- 2
epsilon_squared <- (chi_squared - df + 1) / (n_total - df)
print(paste("Effect size (ε²):", round(epsilon_squared, 3)))

# Interpret:
# ε² < 0.06: Small effect
# 0.06 ≤ ε² < 0.14: Medium effect  
# ε² ≥ 0.14: Large effect

# After running pairwise tests, add significance stars
# Example if High vs Low is significant:
ggplot(violent_summary, aes(x = gent_category, y = median_change, fill = gent_category)) +
  geom_col(alpha = 0.8) +
  geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.2, size = 0.8) +
  
  # Add significance annotation (adjust based on your pairwise results)
  geom_signif(
    comparisons = list(c("Low", "High")),
    annotations = "*",
    y_position = max(violent_summary$q75) * 1.1,
    tip_length = 0.01
  ) +
  
  scale_fill_manual(values = c("Low" = "steelblue", "Medium" = "orange", "High" = "red")) +
  labs(
    x = "Gentrification Intensity",
    y = "Median Violent Crime Change",
    subtitle = "Kruskal-Wallis: χ²(2) = 6.97, p = 0.031*",
    caption = "*Significant at p < 0.05 level"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

-\> There is a significant difference between high group and low group

Pairwise wilcox test :

```{r}
pairwise.wilcox.test(violent_data$median_change,
                    violent_data$gent_category,
                    p.adjust.method = "BH")

# Calculate y positions
y_max <- max(violent_summary$median_change + violent_summary$q75, na.rm = TRUE)

  
```

-\> significant for High-Low

# Building a non binary gentrification index (year to year)

Calculating year over year growth rates for gentrifiable areas :

```{r}
gent_yearly <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    # Year-over-year growth rates (log differences)
    price_g = log(Prixm2Moyen) - log(dplyr::lag(Prixm2Moyen)),
    med_g   = log(MED) - log(dplyr::lag(MED)),
    d9_g    = log(D9) - log(dplyr::lag(D9)),
    d1_g    = -(log(D1) - log(dplyr::lag(D1))),
    pov_g   = -(TP60 - dplyr::lag(TP60))
  ) |>
  ungroup() |>
  filter(!is.na(price_g)) |>
  mutate(
    # Standardizing within the dataset
    price_g_s = as.numeric(scale(price_g)),
    med_g_s   = as.numeric(scale(med_g)),
    d9_g_s    = as.numeric(scale(d9_g)),
    d1_g_s    = as.numeric(scale(d1_g)),
    pov_g_s   = as.numeric(scale(pov_g)),
    
    # Weighted index (SAME WEIGHTS as first-to-last year index)
    gent_index = (price_g_s * 0.25) +
                 (med_g_s * 0.20) +
                 (d9_g_s * 0.10) +
                 (d1_g_s * 0.35) +
                 (pov_g_s * 0.10)
  )
# Calculate year-over-year crime rate change
crime_yearly <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    crime_rate = total_crimes / insee_pop,
    crime_change = log(crime_rate) - log(dplyr::lag(crime_rate))
  ) |>
  ungroup() |>
  select(CODGEO, year, crime_rate, crime_change, total_crimes, insee_pop)

# Create the panel by joining gentrification index with crime data
crime_panel <- gent_yearly |>
  select(CODGEO, year, gent_index) |>
  left_join(crime_yearly, by = c("CODGEO", "year")) |>
  filter(!is.na(crime_change), !is.na(gent_index))

```

Testing wether last year's gentrification predicts this year's crime change
```{r}
library(plm)
panel_data <- pdata.frame(crime_panel, index = c("CODGEO", "year"))
model <- plm(crime_change ~ lag(gent_index), 
             data = panel_data, 
             model = "within",  # Fixed effects
             effect = "twoways") # Both area and year FE
summary(model)
```
-> not significant at all

Testing wether last year's gentrification predicts this year's violent crime change
```{r}
# Creating violent crimes panel
violent_crimes_panel <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  mutate(
    violent_total = rowSums(
      across(c(`Violences physiques intrafamiliales`,
               `Violences physiques hors cadre familial`,
               `Vols avec armes`,
               `Vols violents sans arme`)),
      na.rm = TRUE
    ),
    violent_rate = violent_total / insee_pop
  ) |>
  # Joining with gent_yearly
  left_join(
    gent_yearly |> select(CODGEO, year, gent_index),
    by = c("CODGEO", "year")
  ) |>
  
  # Calculating changes and lagged gentrification
  group_by(CODGEO) |>
  mutate(
    violent_change = log(violent_rate + 0.001) - log(dplyr::lag(violent_rate) + 0.001),
    gent_index_lag1 = dplyr::lag(gent_index, 1)
  ) |>
  ungroup()

# Step 2: Check we have data
violent_crimes_panel |>
  summarise(
    total_rows = n(),
    violent_change_valid = sum(!is.na(violent_change)),
    gent_index_valid = sum(!is.na(gent_index)),
    gent_lag1_valid = sum(!is.na(gent_index_lag1)),
    both_valid = sum(!is.na(violent_change) & !is.na(gent_index_lag1))
  )

# Filtering for analysis
violent_for_analysis <- violent_crimes_panel |>
  filter(!is.na(violent_change) & !is.na(gent_index_lag1))

print(paste("Rows for analysis:", nrow(violent_for_analysis)))

# Basic analysis
cor_result <- cor.test(
  violent_for_analysis$gent_index_lag1,
  violent_for_analysis$violent_change,
  use = "complete.obs"
)
print(cor_result)

model <- lm(violent_change ~ gent_index_lag1, data = violent_for_analysis)
summary(model)

```
-> not significant

Testing multiplte lags :
```{r}
# Creating multiple lags
violent_with_lags <- violent_crimes_panel |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    gent_index_lag2 = dplyr::lag(gent_index, 2),
    gent_index_lag3 = dplyr::lag(gent_index, 3)
  ) |>
  ungroup()

# Testing each lag separately
# Lag 1
lag1_data <- violent_with_lags |> filter(!is.na(violent_change) & !is.na(gent_index_lag1))
cor_lag1 <- cor.test(lag1_data$gent_index_lag1, lag1_data$violent_change)
print("=== LAG 1 (1 year delay) ===")
print(paste("Correlation:", round(cor_lag1$estimate, 4)))
print(paste("P-value:", round(cor_lag1$p.value, 4)))
print(paste("N:", nrow(lag1_data)))

# Lag 2
lag2_data <- violent_with_lags |> filter(!is.na(violent_change) & !is.na(gent_index_lag2))
cor_lag2 <- cor.test(lag2_data$gent_index_lag2, lag2_data$violent_change)
print("=== LAG 2 (2 year delay) ===")
print(paste("Correlation:", round(cor_lag2$estimate, 4)))
print(paste("P-value:", round(cor_lag2$p.value, 4)))
print(paste("N:", nrow(lag2_data)))

# Lag 3
lag3_data <- violent_with_lags |> filter(!is.na(violent_change) & !is.na(gent_index_lag3))
cor_lag3 <- cor.test(lag3_data$gent_index_lag3, lag3_data$violent_change)
print("=== LAG 3 (3 year delay) ===")
print(paste("Correlation:", round(cor_lag3$estimate, 4)))
print(paste("P-value:", round(cor_lag3$p.value, 4)))
print(paste("N:", nrow(lag3_data)))
```
-> Lag 2 is significant, not the others

Comparing wether gentrification affects violent vs non violent crime differently :
```{r}
# Create non-violent crimes panel
nonviolent_panel <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  mutate(
    nonviolent_total = rowSums(
      across(c(`Vols sans violence contre des personnes`,
               `Vols de véhicule`,
               `Vols dans les véhicules`,
               `Vols d'accessoires sur véhicules`,
               `Cambriolages de logement`,
               `Destructions et dégradations volontaires`,
               `Escroqueries et fraudes aux moyens de paiement`,
               `Usage de stupéfiants`,
               `Trafic de stupéfiants`)),
      na.rm = TRUE
    ),
    nonviolent_rate = nonviolent_total / insee_pop
  ) |>
  left_join(
    gent_yearly |> select(CODGEO, year, gent_index),
    by = c("CODGEO", "year")
  ) |>
  group_by(CODGEO) |>
  mutate(
    nonviolent_change = log(nonviolent_rate + 0.001) - log(dplyr::lag(nonviolent_rate) + 0.001),
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    gent_index_lag2 = dplyr::lag(gent_index, 2)
  ) |>
  ungroup()

# Test
nv_lag1 <- nonviolent_panel |> filter(!is.na(nonviolent_change) & !is.na(gent_index_lag1))
cor_nv1 <- cor.test(nv_lag1$gent_index_lag1, nv_lag1$nonviolent_change)
print("=== NON-VIOLENT CRIMES - LAG 1 ===")
print(paste("Correlation:", round(cor_nv1$estimate, 4)))
print(paste("P-value:", round(cor_nv1$p.value, 4)))

nv_lag2 <- nonviolent_panel |> filter(!is.na(nonviolent_change) & !is.na(gent_index_lag2))
cor_nv2 <- cor.test(nv_lag2$gent_index_lag2, nv_lag2$nonviolent_change)
print("=== NON-VIOLENT CRIMES - LAG 2 ===")
print(paste("Correlation:", round(cor_nv2$estimate, 4)))
print(paste("P-value:", round(cor_nv2$p.value, 4)))
```

Visualizing all the lag effects of gentrification on violent crime :
```{r}
lag_summary <- data.frame(
  lag = c(1, 2, 3),
  correlation = c(cor_lag1$estimate, cor_lag2$estimate, cor_lag3$estimate),
  p_value = c(cor_lag1$p.value, cor_lag2$p.value, cor_lag3$p.value),
  n = c(nrow(lag1_data), nrow(lag2_data), nrow(lag3_data))
) |>
  mutate(significant = ifelse(p_value < 0.05, "Yes", "No"))

# Plot
ggplot(lag_summary, aes(x = factor(lag), y = correlation, fill = significant)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("No" = "gray", "Yes" = "red")) +
  labs(
    x = "Lag (Years)",
    y = "Correlation with Violent Crime Change",
    title = "Effect of Gentrification on Violent Crime at Different Lags",
    subtitle = "Testing if gentrification effects take time to materialize"
  ) +
  theme_minimal()
```

Including multiple lags in one model (for violent crimes)
```{r}
library(plm)

panel_multi_lag <- violent_with_lags |>
  filter(!is.na(violent_change) & !is.na(gent_index_lag1) & !is.na(gent_index_lag2))

if(nrow(panel_multi_lag) > 100) {
  panel_data <- pdata.frame(panel_multi_lag, index = c("CODGEO", "year"))
  
  model_lags <- plm(
    violent_change ~ gent_index_lag1 + gent_index_lag2 + gent_index_lag3,
    data = panel_data,
    model = "within",
    effect = "individual"
  )
  
  print(summary(model_lags))
}
```

Visualizing the effects of lag 1 and 2 gentrification on non violent crime over time :
```{r}
# Create summary data
lag_effects_nv <- data.frame(
  Lag = c("Lag 1\n(1 year)", "Lag 2\n(2 years)"),
  Correlation = c(0.1411, -0.09),
  P_value = c(0, 0.0003),
  Effect = c("Increase", "Decrease")
)

ggplot(lag_effects_nv, aes(x = Lag, y = Correlation, fill = Effect)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_text(aes(label = paste0("r = ", Correlation, "\np < 0.001")), 
            vjust = ifelse(lag_effects_nv$Correlation > 0, -0.5, 1.5),
            size = 4) +
  scale_fill_manual(values = c("Increase" = "#e74c3c", "Decrease" = "#27ae60")) +
  labs(
    x = "Time After Gentrification",
    y = "Correlation with Non-Violent Crime Change",
    title = "Gentrification Effects on Non-Violent Crime Over Time",
    subtitle = "Short-term increase followed by longer-term decrease"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
Comparison table of the effects of gentrification at lag 1 and 2 on violent and non violent crime
```{r}
# Summary comparison table
comparison <- data.frame(
  Crime_Type = c("Violent", "Violent", "Non-Violent", "Non-Violent"),
  Lag = c("1 year", "2 years", "1 year", "2 years"),
  Correlation = c(cor_lag1$estimate, cor_lag2$estimate, 0.1411, -0.09),
  Significant = c(
    ifelse(cor_lag1$p.value < 0.05, "Yes", "No"),
    ifelse(cor_lag2$p.value < 0.05, "Yes", "No"),
    "Yes", "Yes"
  )
)

knitr::kable(comparison, digits = 3, 
             col.names = c("Crime Type", "Lag", "Correlation", "Significant (p<0.05)"))
```

Testing the effects of lag gentrification by crime type (non violents)
```{r}
# List of crime types to test
crime_types <- names(data_final)[8:22]
crime_types <- crime_types[crime_types != "Violences sexuelles"]


# Function to test each crime type
test_crime_type <- function(crime_col) {
  crime_panel <- data_final |>
    filter(CODGEO %in% gentrifiable_areas) |>
    arrange(CODGEO, year) |>
    mutate(
      crime_rate = .data[[crime_col]] / insee_pop
    ) |>
    left_join(
      gent_yearly |> select(CODGEO, year, gent_index),
      by = c("CODGEO", "year")
    ) |>
    group_by(CODGEO) |>
    mutate(
      crime_change = log(crime_rate + 0.0001) - log(dplyr::lag(crime_rate) + 0.0001),
      gent_index_lag1 = dplyr::lag(gent_index, 1)
    ) |>
    ungroup() |>
    filter(!is.na(crime_change) & !is.na(gent_index_lag1) & is.finite(crime_change))
  
  if(nrow(crime_panel) > 30) {
    cor_result <- cor.test(crime_panel$gent_index_lag1, crime_panel$crime_change)
    return(data.frame(
      crime_type = crime_col,
      correlation = cor_result$estimate,
      p_value = cor_result$p.value,
      n = nrow(crime_panel),
      significant = ifelse(cor_result$p.value < 0.05, "Yes", "No")
    ))
  }
}

# Run for all crime types
crime_results <- bind_rows(lapply(crime_types, test_crime_type))

# Display results
crime_results |>
  arrange(p_value) |>
  knitr::kable(digits = 4, col.names = c("Crime Type", "Correlation", "P-value", "N", "Significant"))

# Visualize
ggplot(crime_results, aes(x = reorder(crime_type, correlation), y = correlation, fill = significant)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("No" = "gray70", "Yes" = "#e74c3c")) +
  labs(
    x = NULL,
    y = "Correlation with Crime Change (Lag 1)",
    title = "Gentrification Effects by Crime Type",
    subtitle = "Which crimes respond most to gentrification?"
  ) +
  theme_minimal()
```
-> Gentrification has a significant positive effect on vols sans violence contre des personnes, destructions et dégradations volontaires, vols de véhicule, violences physique hors cadre familial and a significant negative effect on usage de stupéfiants (AFD).



Comparing the effects of gentrification on  richer vs poorer areas :
```{r}
# Separating areas by income
interaction_data <- violent_crimes_panel |>
  left_join(baseline |> select(CODGEO, baseline_med), by = "CODGEO") |>
  mutate(
    income_tercile = ntile(baseline_med, 3),
    income_group = case_when(
      income_tercile == 1 ~ "Low Income",
      income_tercile == 2 ~ "Medium Income",
      income_tercile == 3 ~ "High Income"
    )
  ) |>
  filter(!is.na(violent_change) & !is.na(gent_index_lag1))

# Test by income group
interaction_results <- interaction_data |>
  group_by(income_group) |>
  summarise(
    correlation = cor(gent_index_lag1, violent_change, use = "complete.obs"),
    p_value = cor.test(gent_index_lag1, violent_change)$p.value,
    n = n()
  )

print(interaction_results)

# Regression with interaction
interaction_model <- lm(violent_change ~ gent_index_lag1 * income_group, 
                        data = interaction_data)
summary(interaction_model)
```
-> Gentrification reduces violent crime in initially low-income areas, but slightly increases it in higher-income areas. The effect of gentrification on crime depends on the neighborhood's initial economic status


But maybe the causality is in the other direction ? does gentrification cause violent crime change or does violent crime causes gentrification ? Let's use Granger Causality
```{r}
library(plm)

# Preparing balanced panel
granger_data <- violent_crimes_panel |>
  filter(!is.na(violent_change) & !is.na(gent_index) & !is.na(gent_index_lag1)) |>
  group_by(CODGEO) |>
  filter(n() >= 3) |>  # Need at least 3 observations per area
  ungroup()

granger_panel <- pdata.frame(granger_data, index = c("CODGEO", "year"))

# Model 1: Crime change predicted by lagged gentrification
model_gent_causes_crime <- plm(
  violent_change ~ gent_index_lag1,
  data = granger_panel,
  model = "within",
  effect = "individual"
)

# Model 2: Reverse - does crime cause gentrification?
granger_data2 <- granger_data |>
  group_by(CODGEO) |>
  mutate(violent_change_lag1 = dplyr::lag(violent_change, 1)) |>
  ungroup() |>
  filter(!is.na(violent_change_lag1))

granger_panel2 <- pdata.frame(granger_data2, index = c("CODGEO", "year"))

model_crime_causes_gent <- plm(
  gent_index ~ violent_change_lag1,
  data = granger_panel2,
  model = "within",
  effect = "individual"
)

print("=== Does Gentrification Cause Crime? ===")
summary(model_gent_causes_crime)

print("=== Does Crime Cause Gentrification? ===")
summary(model_crime_causes_gent)
```
-> Crime changes predict gentrification! Areas that become safer attract more gentrification the following year. This suggests a reverse causality or feedback loop - it's not just that gentrification affects crime, but crime levels also influence where gentrification occurs.

Visualization 1: Interaction effect
```{r}
ggplot(interaction_data |> filter(!is.na(income_group)), 
       aes(x = gent_index_lag1, y = violent_change, color = income_group)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("Low Income" = "#e74c3c", 
                                 "Medium Income" = "#f39c12", 
                                 "High Income" = "#27ae60")) +
  labs(
    x = "Previous Year's Gentrification Index",
    y = "Violent Crime Change",
    color = "Initial Income Level",
    title = "Gentrification's Effect on Crime Varies by Initial Income",
    subtitle = "Gentrification REDUCES crime in low-income areas, but INCREASES it in high-income areas"
  ) +
  theme_minimal()
```

Visualization 2: Reverse causality
```{r}
ggplot(granger_data2 |> filter(!is.na(violent_change_lag1) & !is.na(gent_index)), 
       aes(x = violent_change_lag1, y = gent_index)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Previous Year's Violent Crime Change",
    y = "This Year's Gentrification Index",
    title = "Crime Drives Gentrification Patterns",
    subtitle = "Areas with declining crime attract more gentrification (r = -0.19, p < 0.001)"
  ) +
  theme_minimal()
```
Do neighboring areas' gentrification affect local crime?
```{r}
#Identifying department from CODGEO
spatial_data <- violent_crimes_panel |>
  mutate(department = substr(CODGEO, 1, 2)) |>
  group_by(department, year) |>
  mutate(
    # Average gentrification in the same department (excluding self)
    n_in_dept = n(),
    sum_gent_dept = sum(gent_index, na.rm = TRUE),
    neighbor_gent = (sum_gent_dept - gent_index) / (n_in_dept - 1)
  ) |>
  ungroup() |>
  group_by(CODGEO) |>
  mutate(
    neighbor_gent_lag1 = dplyr::lag(neighbor_gent, 1)
  ) |>
  ungroup() |>
  filter(!is.na(violent_change) & !is.na(neighbor_gent_lag1))

# Does neighboring gentrification affect local crime?
cor_neighbor <- cor.test(spatial_data$neighbor_gent_lag1, spatial_data$violent_change)
print("=== Neighboring Gentrification Effect ===")
print(paste("Correlation:", round(cor_neighbor$estimate, 4)))
print(paste("P-value:", round(cor_neighbor$p.value, 4)))

# Combined model
spatial_model <- lm(violent_change ~ gent_index_lag1 + neighbor_gent_lag1, 
                    data = spatial_data)
summary(spatial_model)
```
-> When surrounding cities gentrify (in this example, cities of the same département), local violent crime increases locally 

Do we observe different effects depending on the size of the city ? (urban/rural comparison based on population)
```{r}
# Creating urban/rural indicator based on population
pop_data <- data_final |>
  group_by(CODGEO) |>
  summarise(mean_pop = mean(insee_pop, na.rm = TRUE))

urban_rural_data <- violent_crimes_panel |>
  left_join(pop_data, by = "CODGEO") |>
  mutate(
    area_type = case_when(
      mean_pop > 50000 ~ "Large City",
      mean_pop > 10000 ~ "Medium City",
      TRUE ~ "Small City"
    )
  ) |>
  filter(!is.na(violent_change) & !is.na(gent_index_lag1))

# Testing by area type
urban_rural_results <- urban_rural_data |>
  group_by(area_type) |>
  summarise(
    correlation = cor(gent_index_lag1, violent_change, use = "complete.obs"),
    p_value = cor.test(gent_index_lag1, violent_change)$p.value,
    n = n()
  )

print(urban_rural_results)

# Visualization
ggplot(urban_rural_data, aes(x = gent_index_lag1, y = violent_change, color = area_type)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~area_type) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Previous Year's Gentrification Index",
    y = "Violent Crime Change",
    title = "Gentrification-Crime Relationship by City Size"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
In larger cities there is a significantly positive correlation between the lagged gentrification and violent crime change rate



