---
title: "Noe Leo Final"
author: "Léo-Paul Garelli and Noé Mansuy"
format: html
editor: 
  markdown: 
    wrap: 72
---

Github project : https://github.com/LEOPAUL55/Noe_Leo_final.git

# 1. Research question

Theorized during the 1960 decade, gentrification is an urban phenomenon defined as the appropriation of working-class neighborhoods by the upper class. More formally, gentrification materializes by a rise in the real estate market price, as well as a rise in the average standard linving of a quarter. Indeed, the rehabilitation of houses by upper classes leads to a rise in the real estate price market of the quarter. Disadvantaged classes are forced to leave to worse located areas and are replaced by a wealthier population in their initial neighborhood. This phenomenon creates a conflict within social classes that affects the society in different ways, by the education, the inequalities, the health or the criminality.

This project aims to answer the following question: Does the gentrification process in certain French cities correlates with a decrease in recorded crime in these same municipalities, and is there a displacement effect towards non-gentrified cities? The purpose of the study is to estimate the effect of gentrification on criminality at the scale of France on the short and middle run. The goal is to have a better understanding of this phenomenon to implement adequate public policies. To do so, the study is based on the public data of 899 cities in France between 2016 and 2021.

First, the study try to build a gentrification index that would be able to measure the degree of gentrification of each city on the period, based on the process definition. The idea is to have a better understanding of the span and intensity of the phenomenon in France and to separate cities by degree of gentrification. Then, the main goal is to estimate the impact of gentrification on the evolution of criminality by cities. The gentrification index will enable to estimate the direct and lagged effect of gentrification on the criminality's evolution for each city. Indeed, it will measure the effect of gentrification on the global and by type crime rate's evolution.

# 2. Datasets descriptions

```{r}
#| message: false
here::i_am("Noe_Leo_Final.Rproj")
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(vroom)
library(data.table)
library(stringr)
library(plm)
library(psych)
library(mgcv)
library(viridis)
library(FSA)

# Load pre-filtered datasets
poverty <- readRDS(here("data", "cleaned", "poverty.rds"))
crime <- readRDS(here("data", "cleaned", "crime.rds"))
ppsm <- readRDS(here("data", "cleaned", "ppsm.rds"))
city_names <- readRDS(here("data", "cleaned", "city_names.rds"))

# Recreate data_final
data_final <- poverty |>
  left_join(crime, by = c("CODGEO", "year")) |>
  left_join(ppsm, by = c("CODGEO", "year")) |>
  left_join(city_names, by = "CODGEO") |>
  mutate(
    crime_rate = (total_crimes / insee_pop) * 100
  )

# Verify
cat("✓ Data loaded successfully\n")
cat("  - Cities:", n_distinct(data_final$CODGEO), "\n")
cat("  - Years:", n_distinct(data_final$year), "\n")
cat("  - Observations:", nrow(data_final), "\n")
```

The first step of the project is to ensure the reliability of our findings by selecting adequate and accurate data. We need to create two data sets, with respectively a "level" and an "evolution" data set. The "evolution" one will be created thanks to the "level" one. To implement the method of study, essential informations are needed for describing the economic and criminality situation in every cities on the period. More precisely, the build of the gentrification index is based on economic informations such as the the real estate price market (square meter price), the median income, the poverty rate (defined as 60% of the median income) as well as the first and ninth deciles, for each city per year. For the criminality, the study rely on the number of committed crimes for each city per year and type. Thanks to the informations about the level of population, it is possible to compute the crime rate, global and per type.

## a. Data sources

1.  **Real estate price - Datagouv / Chambre des notaires de Paris** 

The informations about the real estate price per city are available on the France state's site. They are given in the files called "Demande de valeur foncière" for every year, that is from 2016 to 2021. This is a collection of various variables on residential housing market at the city level between 2014 and 2024. What is interesting for us is the mean price by square meter of the goods that have been sold each year. This was published by Boris Mericksay. Unfortunatlely, the informations about Paris at the arrondissement level are missing in those files. These important informations can be found on the Chambre des notaires de Paris' site that gives the trimestrial average square meter price of selling. For simplicity, we only focus on the selling price of appartements in the city from 2016 to 2021 (trimester 4). The available file is a PDF that can be easily convert into a CSV file.

2.  **Average standard of living - Insee** 

The files "poverty" are collections of indicators of disposable income in France, at the city level, between 2016 and 2021. We will only focus on the poverty rate, the median standard of living, the first and the ninth deciles. Poverty data are provided as separate yearly files with heterogeneous structures. Variable names differ across years, and the 2018 data are reported at the IRIS level rather than the commune level. To handle this, we implement a loading strategy that reads each file separately while restricting imports to the variables of interest (years 2016:2021). Poverty indicators are harmonized across years. Non‑numeric values are explicitly converted to missing values during loading, but it won't affect our analysis because this only concerns small cities. The result is a consistent panel of indicators comparable across years.

3.  **Criminality - Datagouv / Ministère de l'intérieur** 

The CSV file "criminality" is a database of all the crimes and offences (except traffic offences) at the city level in France, published by French Police Force, since 2016. These offences were identified following a complaint filed by a victim, a report, a testimony, a flagrant offence, a denunciation, but also on the initiative of the security forces. The total number of crimes per year per city is computed by doing the sum of each type of crimes for every year. Thanks to the informations about the number of inhabitants per city given in the past files, it is is possible to compute the global crime rate and per type. We decided to exclude the crime "drug use (AFD)" and "sexual violence" due to the difference between the Police's data and the real situation, that would bias the analysis.

## b. Informations about the data sets

```{r, echo = FALSE}
info <- function(df) {
  data.frame(file = deparse(substitute(df)), Observations = nrow(df), Variables = ncol(df))
}
```

```{r, echo = FALSE}
info_ppsm <- info(ppsm)
info_poverty <- info(poverty)
info_crime <- info(crime) 
bind_rows(info_ppsm, info_poverty, info_crime) |> knitr::kable()
```

## c. Description of variables

Below is the description of variables that are used in the project.

**CODGEO**: Unique code that corresponds to one city. The first two numbers correspond to a department number.

**LIBGEO**: Name of the city according to the CODGEO.

**year**: year in which the data was collected.

**Prixm2Moyen**: Average price per square meter of the real estate goods sold for the year.

**MED**: Median income for this year.

**TP60**: Poverty rate (set at less than 60% of the median income).

**D1**: 1st income decile (income below which there are 10% of the population).

**D9**: 9th income decile (income below which there are 90% of the population).

**Violences physiques intrafamiliales**: Number of domestic physical violence reported for the year.

**Violences physiques hors cadre familial**: Number of non domestic physical violence reported for the year.

**Vols avec armes**: Number of armed robberies reported for the year.

**Vols sans arme**: Number of non armed violent robberies reported for the year.

**Vols sans violence contre des personnes**: Number of non violent robberies against persons reported for the year.

**Cambriolages de logement**: Number of home invasions reported for the year.

**Vols de véhicules**: Number of vehicule theft reported for the year.

**Vols dans les véhicules**: Number of theft from vehicules reported for the year.

**Vols d'accessoires sur véhicules**: Number of theft of vehicle accessories reported for the year.

**Destructions et dégradations volontaires**: Number of willful damage and vandalism reported for the year.

**Usage de stupéfiants**: Number of reported drug use for the year.

**Trafic de stupéfiants**: Number of drug dealing cases reported for the year.

**Escroqueries et fraudes aux moyens de paiement**: Number of fraud and payment method fraud reported for the year.

**total_crimes**: Total number of crimes per cities reported for this year.

**crime_rate**: Total number of crimes per cities divided by the population number of the year.

**insee_pop**: The city's population.

# 3. Data analysis

After obtaining clean data sets, we can now analyse it through several aspects. The first step is to build a gentrification index, that is a mathematical formula which enables to compute and compare the degree of gentrification per cities thanks to the available information in the data set (economic description, real estate price). Second, it will be possible to compute descriptive statistics about gentrification to have a better understanding of the process in France through the period. More precisely, to what extent are french cities affected by gentrification and where is the process the most intense on the map. In the same idea, descriptive statistics about crime in France will be computed to also understand this phenomenon and its evolution. Third, we will try to isolate with the data the causal link between the degree of gentrification and the criminality evolution in the cities using an econometric method.

## a. Building a gentrification index

Recalling the definition, gentrification is the process by which working class quarter are appropriated by upper classes, which exclude the initial population and replace it by a wealthier one. In the data, such a process can be observed using the data sets "ppsm" (real estate price) and "poverty" (economic information). Indeed, a gentrified city can be identified through a rise in the real estate price (Prix2Moyen) which will increase the city's standard living and rise the median income (MED) as well as the first and last decile (D1 and D9) and reduce the rate of poor people in the population (fall of TP60).

Using these data, we can create a gentrification index that is implementable to any city using the following formula:

**Gentrification** = (**price_var** * 0.25) + (**MED_var** * 0.20) + (**D9_var** * 0.10) + (**D1_var** * 0.35) + (**TP60_var** * 0.10)

Gentrification is not about the level of the variables, but their variation. Thus we need to compute the variation in percentage of each variable to implement the formula. Moreover, we justify the weight given to each variable thanks to the literature. According to the initial definition (Ruth Glass, 1964) combined with the work of Neil Smith and the Rent-Gap Theory (1979), gentrification mainly materialized by the displacement of the working class as well as the rise in the real estate price, that justifies the heavy weight on the first decile (D1_var) and the square meter price (price_var). To make the index meaningful, we standardize it using the z-score standardization and center it around 0 (if the index is equal to 0, the city is as gentrify as the national mean). Finally, we exclude the city that have a median income greater than 60% of the national one, considering they are already composed of upper classes, so it cannot be gentrified.

## b. Descriptive statistics: visual exploration of the data

### Gentrification in France, where?

To determine what cities are the most affected by gentrification, we compute the global variation of the economic data between 2016 and 2021. Applying the gentrification index to these data, it is showed in the following table that the most gentrified cities are mainly located in the neighborhoods of the big urban areas, especially Paris (Clichy, Romainville, Saint-Ouen-sur-Seine...). Cities like Clichy or Saint-Ouen-sur-Seine are famous to have been rehabilitate during the last years, that leds to an important gentrification. This is validated by the index.

```{r, echo=FALSE}
#Taking the earliest year of data for each demographic area (CODGEO)
baseline <- data_final |>
  group_by(CODGEO) |>
  filter(year == min(year)) |>
  #capturing the median income as the starting point
  ungroup() |>
  select(CODGEO, baseline_med = MED)

#Making sure only areas below the "60% of median income" threshold are considered "gentrifiable" (by definition, only lower income areas can gentrify)
percentile_cutoff <- quantile(baseline$baseline_med, 0.60, na.rm = TRUE)
```

```{r, echo=FALSE}
#Calculating changes over time 
gentrifiable_areas <- baseline |>
  filter(baseline_med < percentile_cutoff) |>
  pull(CODGEO)

# Calculating percentage changes from first to last year
evolution <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>  
  group_by(CODGEO) |>
  summarise(
    price_var   = ((last(Prixm2Moyen) / first(Prixm2Moyen)) - 1) * 100,
    med_inc_var = (last(MED) / first(MED) - 1) * 100,
    d9_var      = (last(D9) / first(D9) - 1) * 100,
    d1_var      = (last(D1) / first(D1) - 1) * 100,
    pov_var     = last(TP60) - first(TP60),
    .groups = "drop"
  ) |>
  left_join(city_names, by = "CODGEO")
```

```{r, echo=FALSE}
# Converting all the variables to z-score so it makes the different metrics comparable despite different scales 
evolution_std <- evolution |>
  mutate(across(
    c(price_var, med_inc_var, d9_var, d1_var, pov_var),
    scale
  ))
```

```{r, echo=FALSE}
evolution_std <- evolution_std |>
  mutate(
    gent_index = 
      (price_var * 0.25) +   # Emphasizes housing price
      (med_inc_var * 0.2) +  #Median income
      (d9_var * 0.1) +       # Top income
      (d1_var * 0.35) -      #Bottom income
      (pov_var * 0.1)        # Poverty reduction
  ) |>
  arrange(desc(gent_index))


# Sorting areas by gentrification intensity
evolution_std_A <- evolution_std |>
  head(10) |>
  select(CODGEO, LIBGEO, gent_index)|> 
  knitr::kable()

print(evolution_std_A)
```
Moreover, the data also show how gentrification is an urban process. The following graphic represents the average degree of gentrification computed by the index, in function of the side of the population in 2016. It does not show a clear positive relation between both variables, because the relation increases and then deceases. Nevertheless, it shows how the gentrification index is negative and less intense in the smallest cities (less than 50 000 inhabitants), which means gentrification in smallest cities is less intense than the average. It validates the correlation between size of the town and degree of gentrification.

```{r}
pop_2016 <- data_final |> 
  filter(year == 2016) |> 
  select(c(CODGEO, insee_pop)) 

evolution_std_groups <- evolution_std |>
  left_join(pop_2016, by = "CODGEO") |>
  mutate(
    pop_bin = cut(
      insee_pop,
      breaks = seq(
        floor(min(insee_pop, na.rm = TRUE) / 10000) * 10000,
        ceiling(max(insee_pop, na.rm = TRUE) / 10000) * 10000,
        by = 10000
      ),
      right = FALSE
    )
  )

gent_by_pop <- evolution_std_groups |>
  group_by(pop_bin) |>
  summarise(
    mean_gent = mean(gent_index, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    pop_bin_num = as.numeric(sub("\\[(.*),.*", "\\1", pop_bin)))

ggplot(gent_by_pop,
       aes(x = pop_bin_num, y = mean_gent)) +
  geom_line(color = "steelblue", linewidth = 1.2) +
  geom_point(color = "steelblue", size = 3) +
  scale_x_continuous(
    breaks = seq(
      min(gent_by_pop$pop_bin_num, na.rm = TRUE),
      max(gent_by_pop$pop_bin_num, na.rm = TRUE),
      by = 50000
    ),
    labels = scales::label_number(big.mark = " ")
  ) +
  labs(
    x = "Population (bins of 50,000 inhabitants)",
    y = "Average gentrification index",
    title = "Average gentrification by population size"
  ) +
  theme_minimal(base_size = 14)
```
### Criminality in France, evolution between 2016 and 2021

To first understand criminality in France and its evolution, we represent it grouping crimes and offences according to the violence character (violent, non-violent). We consider as violent the following crimes: domestic physical violence, non domestic physical violence, armed robberies and non armed violent robberies. The rest is reported as non violent (non violent robberies against persons, vehicle thefts, theft of vehicle accessories, willful damage and vandalism, fraud and payment method fraud, drug use, drug dealing, home invasions).

```{r}
crime_wide <- data_final |>
  mutate(
    violent_total = rowSums(
      across(c(`Violences physiques intrafamiliales`,
               `Violences physiques hors cadre familial`,
               `Vols avec armes`,
               `Vols violents sans arme`)),
      na.rm = TRUE
    ),
    nonviolent_total = rowSums(
      across(c(`Vols sans violence contre des personnes`,
               `Vols de véhicule`,
               `Vols dans les véhicules`,
               `Vols d'accessoires sur véhicules`,
               `Cambriolages de logement`,
               `Destructions et dégradations volontaires`,
               `Escroqueries et fraudes aux moyens de paiement`,
               `Usage de stupéfiants`,
               `Trafic de stupéfiants`)),
      na.rm = TRUE
    ),
    total_excl_sexual = total_crimes,
    violent_rate = (violent_total / insee_pop) * 100,
    nonviolent_rate = (nonviolent_total / insee_pop) * 100,
    total_rate_excl = (total_excl_sexual / insee_pop) * 100
  ) |>
  group_by(year) |>
  summarise(
    total = median(total_rate_excl, na.rm = TRUE),
    violent = median(violent_rate, na.rm = TRUE),
    nonviolent = median(nonviolent_rate, na.rm = TRUE),
    .groups = "drop"
  )

# Scale factor for secondary axis
scale_factor <- max(crime_wide$nonviolent) / max(crime_wide$violent)

ggplot(crime_wide, aes(x = year)) +
  geom_line(aes(y = nonviolent, color = "Non-Violent"), size = 1.5) +
  geom_point(aes(y = nonviolent, color = "Non-Violent"), size = 4) +
  geom_line(aes(y = total, color = "Total"), size = 1.5) +
  geom_point(aes(y = total, color = "Total"), size = 4) +
  geom_line(aes(y = violent * scale_factor, color = "Violent"), size = 1.5) +
  geom_point(aes(y = violent * scale_factor, color = "Violent"), size = 4) +
  scale_y_continuous(
    name = "Non-Violent & Total (per 100 inh.)",
    sec.axis = sec_axis(~./scale_factor, name = "Violent Crime (per 100 inh.)")
  ) +
  scale_x_continuous(breaks = 2016:2021) +
  scale_color_manual(values = c("Total" = "#2c3e50", 
                                "Violent" = "#e74c3c", 
                                "Non-Violent" = "#3498db")) +
  labs(
    x = "Year",
    color = NULL,
    title = "Crime Rate Evolution in French Cities (2016-2021)",
    subtitle = "Note: Violent crime uses secondary axis (right) | Excluding sexual violence"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    legend.position = "bottom",
    axis.title.y.right = element_text(color = "#e74c3c"),
    axis.text.y.right = element_text(color = "#e74c3c")
  )
```
The above graphic shows the evolution of crime per type (violent, non violent) on the period. We observe a global decrease of non violent crime on the period, with an unstopped fall until 2020 followed by a new increase. The observation is different for the violent crime that have a positive trend on the period, with an only decrease between 2019 and 2020. The mix of both types' crime gives a decreasing trend for the total crime that shaped the non violent crime evolution, with a positive evolution the last year.

### What correlation between gentrification and crime?

Let's compare between 3 arbitrary gentrification categories among the "gentrifiable" cities : Low gentrification (bottom 25% of the index), medium gentrification (middle 50% of the index) and high gentrification (top 25% of the index).
```{r}
evolution_std_A_groups <- evolution_std_A |>
  mutate(
    gent_category = case_when(
      gent_index > quantile(gent_index, 0.75, na.rm = TRUE) ~ "High",
      gent_index > quantile(gent_index, 0.25, na.rm = TRUE) ~ "Medium",
      TRUE ~ "Low"
    ),
    gent_category = factor(gent_category, levels = c("Low", "Medium", "High"))
  )
```
Keeping the same nomenclature, the following graphic seems to depict a positive correlation between gentrification and crime. The period observes a global decrease in criminality, still different according to the degree of gentrification. Violent crimes rose everywhere, but in a more important way in highly gentrified cities. Same for the non violent crimes that decreased in every cities but less intensely in the highly gentrified cities compared to low and medium gentrified cities.







# Cross sectional analysis

```{r}
evolution_std_A_groups <- evolution_std_A |>
  mutate(
    gent_category = case_when(
      gent_index > quantile(gent_index, 0.75, na.rm = TRUE) ~ "High",
      gent_index > quantile(gent_index, 0.25, na.rm = TRUE) ~ "Medium",
      TRUE ~ "Low"
    ),
    gent_category = factor(gent_category, levels = c("Low", "Medium", "High"))
  )

table(evolution_std_A_groups$gent_category)

crime_rate_change_continuous <- data_final |>
  filter(year %in% c(2016, 2021)) |>
  select(CODGEO, year, total_crimes, insee_pop) |>
  pivot_wider(
    names_from = year,
    values_from = c(total_crimes, insee_pop),
    names_sep = "_"
  ) |>
  mutate(
    crime_rate_2016 = total_crimes_2016 / insee_pop_2016,
    crime_rate_2021 = total_crimes_2021 / insee_pop_2021,
    crime_rate_change = (crime_rate_2021 - crime_rate_2016) / crime_rate_2016
  ) |>
  left_join(
    evolution_std_A_groups |> select(CODGEO, gent_index, gent_category),
    by = "CODGEO"
  ) |>
  filter(!is.na(gent_category))

```

Scatter plot of the crime rate change compared to the gentrification
index

```{r}
library(viridis)

ggplot(crime_rate_change_continuous,
       aes(x = gent_index, y = crime_rate_change)) +
  geom_point(aes(color = gent_index), 
             alpha = 0.6, size = 2.5) +
  scale_color_viridis_c(option = "plasma", name = "Gentrification\nIntensity") +
  geom_smooth(method = "lm", color = "#FF6B6B", size = 1.2, 
              fill = "#FF6B6B", alpha = 0.15) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.3, color = "gray40") +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.3, color = "gray40") +
  scale_y_continuous(labels = scales::percent_format()) +
  coord_cartesian(ylim = c(-0.5, 0.5)) +
  labs(
    x = "Gentrification Index (standardized)",
    y = "Crime Rate Change (2016-2021)",
    title = "Small Link Between Gentrification and Crime Changes",
    subtitle = "Analysis of initially lower-income areas (bottom 60% baseline income)",
    caption = "Each point represents one geographic area | Red line shows linear trend"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0),
    plot.subtitle = element_text(size = 12, color = "gray40", hjust = 0),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )
```
-> Visually, it seems to exist a slight correlation between gentrification intensity and crime rate change between 2016 and 2021 among gentrifiable cities : less gentrified cities tend to experiment more crime rate reduction than more gentrified cities


## Gentrification categories :
Let's compare between 3 arbitrary gentrification categories among the "gentrifiable" cities : Low gentrification (bottom 25% of the index), medium gentrification (middle 50% of the index) and high gentrification (top 25% of the index) :
```{r}
evolution_std_A_groups <- evolution_std_A |>
  mutate(
    gent_category = case_when(
      gent_index > quantile(gent_index, 0.75, na.rm = TRUE) ~ "High",
      gent_index > quantile(gent_index, 0.25, na.rm = TRUE) ~ "Medium",
      TRUE ~ "Low"
    ),
    gent_category = factor(gent_category, levels = c("Low", "Medium", "High"))
  )

table(evolution_std_A_groups$gent_category)
```

## Crime change by category

-> Visually, it seems to exist a slight correlation between gentrification intensity and crime rate change between 2016 and 2021 among gentrifiable cities : less gentrified cities tend to experiment more crime rate reduction than more gentrified cities

Plotting the crime rate change among three arbitrary gentrification levels :

```{r}
# Calculate confidence intervals
ci_stats <- crime_rate_change_continuous |>
  group_by(gent_category) |>
  summarise(
    mean = mean(crime_rate_change, na.rm = TRUE),
    ci_lower = mean - 1.96 * sd(crime_rate_change, na.rm = TRUE) / sqrt(n()),
    ci_upper = mean + 1.96 * sd(crime_rate_change, na.rm = TRUE) / sqrt(n()),
    median = median(crime_rate_change, na.rm = TRUE)
  )

ggplot(ci_stats |> filter(gent_category %in% c("Low", "High", "Medium")),
       aes(x = gent_category, y = mean, color = gent_category)) +
  # CI lines
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0, size = 2, alpha = 0.4) +
  # Mean points
  geom_point(size = 8) +
  # Median points
  geom_point(aes(y = median), shape = 1, size = 8, stroke = 1.5) +
  scale_color_manual(values = c("Low" = "#3498db", "Medium" = "#f39c12", "High" = "#e74c3c")) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.3) +
  coord_cartesian(ylim = c(-0.1, 0.1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  labs(
    x = NULL,
    y = "Crime Rate Change (2016–2021)",
    title = "Mean Crime Rate Changes Across Gentrification Levels",
    subtitle = "Solid dots = mean | Hollow circles = median | Lines = 95% CI",
    
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "gray40"),
    axis.text.x = element_text(size = 14, face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )
```
-> There is a diminution in crime rate between all three gentrification levels, with a higher diminution for the low category and a lower diminution for the highly gentrified category

Crime change by type and gentrification intensity : 
```{r}

# Reshaping crime data per type
crime_change_type_continuous <- data_final |>
  pivot_longer(
    cols = c(
      `Violences physiques intrafamiliales`,
      `Violences physiques hors cadre familial`,
      `Vols avec armes`,
      `Vols violents sans arme`,
      `Vols sans violence contre des personnes`,
      `Vols de véhicule`,
      `Vols dans les véhicules`,
      `Vols d'accessoires sur véhicules`,
      `Cambriolages de logement`,
      `Destructions et dégradations volontaires`,
      `Escroqueries et fraudes aux moyens de paiement`,
      `Usage de stupéfiants`,
      `Trafic de stupéfiants`
    ),
    names_to = "crime_type",
    values_to = "crime_count"
  ) |>
  filter(year %in% c(2016, 2021)) |>
  select(CODGEO, year, crime_type, crime_count, insee_pop) |>
  
  # Creating year-over-year comparison
  pivot_wider(
    names_from = year,
    values_from = c(crime_count, insee_pop),
    names_sep = "_"
  ) |>
  # Calculating poopulation-adjusted change
  mutate(
    pop_adj_rate_change = if_else(
      crime_count_2016 > 0 & insee_pop_2016 > 0,
      ((crime_count_2021 / crime_count_2016) /
       (insee_pop_2021 / insee_pop_2016)) - 1,
      NA_real_
    )
  ) |>
  # Merging with gentrification levels
  left_join(
    evolution_std_A_groups |> select(CODGEO, gent_index, gent_category),
    by = "CODGEO"
  ) |>
  filter(!is.na(gent_category))

# Summary by crime type and gentrification level
crime_change_type_summary <- crime_change_type_continuous |>
  filter(is.finite(pop_adj_rate_change)) |>
  group_by(crime_type, gent_category) |>
  summarise(
    median_change = median(pop_adj_rate_change, na.rm = TRUE),
    n_cities = n(),
    .groups = "drop"
  )

# Plot
ggplot(crime_change_type_summary,
       aes(x = crime_type, y = median_change, fill = gent_category)) +
  geom_col(position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("Low" = "steelblue", "Medium" = "orange", "High" = "red")) +
  labs(
    x = "Type of Crime",
    y = "Population-Adjusted Crime Change (2016–2021)",
    fill = "Gentrification\nIntensity",
    title = "Crime Change by Type and Gentrification Intensity"
  ) +
  theme_minimal()
```
We can see the that the crime rate changes in the same direction among the gentrifiable cities for all type of crimes, but with very heterogeneous intensity depending the gentrification intensity for some type of crimes (violences physiques intrafamiliales, vols avec armes, cambriolages de logements...)

## Violent vs Non-violent breakdown

What if we split crimes into 2 distinct categories : Violent vs Non Violent crimes ?
```{r}

# Creating crime categories 
crime_change_type_cat <- crime_change_type_continuous |>
  mutate(
    crime_category = case_when(
      crime_type %in% c(
       
        "Violences physiques intrafamiliales",
        "Violences physiques hors cadre familial",
        "Vols avec armes",
        "Vols violents sans arme"
      ) ~ "Violent crimes",
      crime_type %in% c(
        "Vols sans violence contre des personnes",
        "Vols de véhicule",
        "Vols dans les véhicules",
        "Vols d'accessoires sur véhicules",
        "Cambriolages de logement",
        "Destructions et dégradations volontaires",
        "Escroqueries et fraudes aux moyens de paiement",
        "Usage de stupéfiants",
        "Trafic de stupéfiants"
      ) ~ "Non violent crimes",
      TRUE ~ NA_character_
    )
  )
```

Summarizing by category :

```{r}

# Calculating the median population adjusted change for each combination of crime category and gentrification level
crime_cat_summary <- crime_change_type_cat |>
  filter(is.finite(pop_adj_rate_change), !is.na(crime_category)) |>
  group_by(crime_category, gent_category) |>
  summarise(
    median_change = median(pop_adj_rate_change, na.rm = TRUE),
    n_obs = n(),
    .groups = "drop"
  )
```

Violent vs non violent crime change by gentrification intensity :

```{r}
# Plot
ggplot(crime_cat_summary,
       aes(x = crime_category, y = median_change, fill = gent_category)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Low" = "steelblue", "Medium" = "orange", "High" = "red")) +
  labs(
    x = "Crime Category",
    y = "Population-Adjusted Crime Change (2016–2021)",
    fill = "Gentrification\nIntensity",
    title = "Violent vs Non-Violent Crime Change\nby Gentrification Intensity"
  ) +
  theme_minimal()

# Table
knitr::kable(
  crime_cat_summary,
  digits = 3,
  col.names = c("Crime Category", "Gentrification Level", "Mean Change", "N Observations")
)
```

-> Very interesting results : cities that had a higher gentrification also witnessed a lower violent crime rate increase and a lower non violent crimes increase.

## Statistical tests

Let's check if the changes are significant :

Tests whether crime rate changes differ significantly across the three gentrification levels :
```{r}
# Tests whether crime rate changes differ significantly across the three gentrification levels

anova_result <- aov(crime_rate_change ~ gent_category, data = crime_rate_change_continuous)
summary(anova_result)

# Tests wether there's a linear relationship between gentrification intensity and crime change 
cor.test(
  crime_rate_change_continuous$gent_index,
  crime_rate_change_continuous$crime_rate_change,
  use = "complete.obs"
)
```

-> There is NO significant difference in total crime rate changes between the three gentrification categories (low,medium high), and there is NO significant linear relationship between the intensity of gentrification and changes in crime rates

What if we're checking for significativity among violent crimes between gentrification categories only ?
```{r}
# Checking only for violent crimes

violent_data <- crime_change_type_cat |>
  filter(crime_category == "Violent crimes", is.finite(pop_adj_rate_change)) |>
  group_by(CODGEO, gent_category) |>
  summarise(median_change = median(pop_adj_rate_change), .groups = "drop")  # <- median here

summary(aov(median_change ~ gent_category, data = violent_data))
```
-> There ARE statistically significant differences in violent crime rate changes between cities with different gentrification levels



Plotting violent crime growth among gentrification intensities
categories :
```{r}
# Get the medians for the plot
violent_summary <- violent_data |>
  group_by(gent_category) |>
  summarise(
    median_change = median(median_change, na.rm = TRUE),
    # For error bars around median, we can use bootstrapped confidence intervals
    # or use IQR/quantiles instead of SE
    q25 = quantile(median_change, 0.25, na.rm = TRUE),
    q75 = quantile(median_change, 0.75, na.rm = TRUE),
    n = n()
  )

# Plot with median and IQR bars
ggplot(violent_summary, aes(x = gent_category, y = median_change, fill = gent_category)) +
  geom_col(alpha = 0.8) +
  # Use IQR as error bars (shows middle 50% of data)
  geom_errorbar(aes(ymin = q25, ymax = q75), 
                width = 0.2, linewidth = 0.8) +
  scale_fill_manual(values = c("Low" = "steelblue", "Medium" = "orange", "High" = "red")) +
  labs(
    x = "Gentrification Intensity",
    y = "Median Violent Crime Change (Pop-Adjusted)",
    caption = "Error bars show interquartile range (25th-75th percentiles)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Overall test: Are there differences between groups?
```{r}
library(ggsignif)

kruskal_test <- kruskal.test(median_change ~ gent_category, data = violent_data)
print(kruskal_test)

library(FSA)

# Dunn's test with Bonferroni correction
dunn_test <- dunnTest(median_change ~ gent_category, 
                      data = violent_data,
                      method = "bonferroni")
print(dunn_test)

# Or pairwise Wilcoxon tests
pairwise_results <- pairwise.wilcox.test(
  violent_data$median_change,
  violent_data$gent_category,
  p.adjust.method = "bonferroni"
)
print(pairwise_results)

# Epsilon-squared effect size
n_total <- nrow(violent_data)
chi_squared <- 6.9672
df <- 2
epsilon_squared <- (chi_squared - df + 1) / (n_total - df)
print(paste("Effect size (ε²):", round(epsilon_squared, 3)))

# Interpret:
# ε² < 0.06: Small effect
# 0.06 ≤ ε² < 0.14: Medium effect  
# ε² ≥ 0.14: Large effect


```
-> There ARE significant differences between groups


Pairwise wilcox test :

```{r}
pairwise.wilcox.test(violent_data$median_change,
                    violent_data$gent_category,
                    p.adjust.method = "BH")

# Calculate y positions
y_max <- max(violent_summary$median_change + violent_summary$q75, na.rm = TRUE)

  
```
-> Similar results


# Panel Analysis

## Year over year gentrification index

```{r}
gent_yearly <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    # Year-over-year growth rates (log differences)
    price_g = log(Prixm2Moyen) - log(dplyr::lag(Prixm2Moyen)),
    med_g   = log(MED) - log(dplyr::lag(MED)),
    d9_g    = log(D9) - log(dplyr::lag(D9)),
    d1_g    = -(log(D1) - log(dplyr::lag(D1))),
    pov_g   = -(TP60 - dplyr::lag(TP60))
  ) |>
  ungroup() |>
  filter(!is.na(price_g)) |>
  mutate(
    # Standardizing within the dataset
    price_g_s = as.numeric(scale(price_g)),
    med_g_s   = as.numeric(scale(med_g)),
    d9_g_s    = as.numeric(scale(d9_g)),
    d1_g_s    = as.numeric(scale(d1_g)),
    pov_g_s   = as.numeric(scale(pov_g)),
    
    # Weighted index (SAME WEIGHTS as first-to-last year index)
    gent_index = (price_g_s * 0.25) +
                 (med_g_s * 0.20) +
                 (d9_g_s * 0.10) +
                 (d1_g_s * 0.35) +
                 (pov_g_s * 0.10)
  )
# Calculate year-over-year crime rate change
crime_yearly <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    crime_rate = total_crimes / insee_pop,
    crime_change = log(crime_rate) - log(dplyr::lag(crime_rate))
  ) |>
  ungroup() |>
  select(CODGEO, year, crime_rate, crime_change, total_crimes, insee_pop)

# Create the panel by joining gentrification index with crime data
crime_panel <- gent_yearly |>
  select(CODGEO, year, gent_index) |>
  left_join(crime_yearly, by = c("CODGEO", "year")) |>
  filter(!is.na(crime_change), !is.na(gent_index))

```


## Lagged effects


Lagged effects on Total crime :
```{r}
library(plm)
panel_data <- pdata.frame(crime_panel, index = c("CODGEO", "year"))
model <- plm(crime_change ~ lag(gent_index), 
             data = panel_data, 
             model = "within",  # Fixed effects
             effect = "twoways") # Both area and year FE
summary(model)
```
-> Cities that experienced more gentrification in one year tended to have slightly smaller increases (or larger decreases) in crime the following year (very small effect)

Lagged effects on Violent crime :
```{r}
# Creating violent crimes panel
violent_crimes_panel <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  mutate(
    violent_total = rowSums(
      across(c(`Violences physiques intrafamiliales`,
               `Violences physiques hors cadre familial`,
               `Vols avec armes`,
               `Vols violents sans arme`)),
      na.rm = TRUE
    ),
    violent_rate = violent_total / insee_pop
  ) |>
  # Joining with gent_yearly
  left_join(
    gent_yearly |> select(CODGEO, year, gent_index),
    by = c("CODGEO", "year")
  ) |>
  
  # Calculating changes and lagged gentrification
  group_by(CODGEO) |>
  mutate(
    violent_change = log(violent_rate + 0.001) - log(dplyr::lag(violent_rate) + 0.001),
    gent_index_lag1 = dplyr::lag(gent_index, 1)
  ) |>
  ungroup()

# Step 2: Check we have data
violent_crimes_panel |>
  summarise(
    total_rows = n(),
    violent_change_valid = sum(!is.na(violent_change)),
    gent_index_valid = sum(!is.na(gent_index)),
    gent_lag1_valid = sum(!is.na(gent_index_lag1)),
    both_valid = sum(!is.na(violent_change) & !is.na(gent_index_lag1))
  )

# Filtering for analysis
violent_for_analysis <- violent_crimes_panel |>
  filter(!is.na(violent_change) & !is.na(gent_index_lag1))

print(paste("Rows for analysis:", nrow(violent_for_analysis)))

# Basic analysis
cor_result <- cor.test(
  violent_for_analysis$gent_index_lag1,
  violent_for_analysis$violent_change,
  use = "complete.obs"
)
print(cor_result)

model <- lm(violent_change ~ gent_index_lag1, data = violent_for_analysis)
summary(model)

```
-> A linear regression revealed that last year's gentrification significantly 
predicts reductions in violent crime growth this year (β = -0.038, p < 0.0001). 
The effect is 2.6 times larger for violent crimes than for total crimes, 
and gentrification explains approximately 1% of the variance in violent crime 
changes—substantially more than its explanatory power for overall crime

Cities that gentrified more in one year experienced significantly smaller increases (or larger decreases) in violent crimes the following year. This effect is 2.6 times stronger for violent crimes than for crimes overall


## Multiple lags analysis on violent crimes

```{r}
# Creating multiple lags
violent_with_lags <- violent_crimes_panel |>
  arrange(CODGEO, year) |>
  group_by(CODGEO) |>
  mutate(
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    gent_index_lag2 = dplyr::lag(gent_index, 2),
    gent_index_lag3 = dplyr::lag(gent_index, 3)
  ) |>
  ungroup()

# Testing each lag separately
# Lag 1
lag1_data <- violent_with_lags |> filter(!is.na(violent_change) & !is.na(gent_index_lag1))
cor_lag1 <- cor.test(lag1_data$gent_index_lag1, lag1_data$violent_change)
print("=== LAG 1 (1 year delay) ===")
print(paste("Correlation:", round(cor_lag1$estimate, 4)))
print(paste("P-value:", round(cor_lag1$p.value, 4)))
print(paste("N:", nrow(lag1_data)))

# Lag 2
lag2_data <- violent_with_lags |> filter(!is.na(violent_change) & !is.na(gent_index_lag2))
cor_lag2 <- cor.test(lag2_data$gent_index_lag2, lag2_data$violent_change)
print("=== LAG 2 (2 year delay) ===")
print(paste("Correlation:", round(cor_lag2$estimate, 4)))
print(paste("P-value:", round(cor_lag2$p.value, 4)))
print(paste("N:", nrow(lag2_data)))

# Lag 3
lag3_data <- violent_with_lags |> filter(!is.na(violent_change) & !is.na(gent_index_lag3))
cor_lag3 <- cor.test(lag3_data$gent_index_lag3, lag3_data$violent_change)
print("=== LAG 3 (3 year delay) ===")
print(paste("Correlation:", round(cor_lag3$estimate, 4)))
print(paste("P-value:", round(cor_lag3$p.value, 4)))
print(paste("N:", nrow(lag3_data)))
```
## Non-Linear effects (GAM)
```{r}
# Check for mean reversion
violent_with_lags |>
  filter(!is.na(violent_change)) |>
  group_by(CODGEO) |>
  mutate(
    violent_change_lag1 = dplyr::lag(violent_change, 1),
    violent_change_lag2 = dplyr::lag(violent_change, 2)
  ) |>
  summarise(
    autocorr = cor(violent_change, violent_change_lag1, use = "complete.obs"),
    mean_reversion = cor(violent_change, violent_change_lag2, use = "complete.obs")
  ) |>
  summarise(
    mean_autocorr = mean(autocorr, na.rm = TRUE),
    mean_reversion = mean(mean_reversion, na.rm = TRUE)
  )

# Test different functional forms
library(mgcv)
model_nonlinear <- gam(violent_change ~ s(gent_index_lag1) + s(gent_index_lag2),
                       data = violent_with_lags)
summary(model_nonlinear)
plot(model_nonlinear)
```
Violent crime exhibits strong negative autocorrelation (r = -0.43), suggesting natural oscillations. A Generalized Additive Model reveals non-linear effects: gentrification initially reduces violent crime with diminishing returns (p = 0.027), but predicts increased crime two years later (p < 0.001). The Year 2 positive association may reflect both genuine delayed effects and statistical mean reversion from the initial crime drop

## Violent vs Non Violent comparison :
```{r}
# Create non-violent crimes panel
nonviolent_panel <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  mutate(
    nonviolent_total = rowSums(
      across(c(`Vols sans violence contre des personnes`,
               `Vols de véhicule`,
               `Vols dans les véhicules`,
               `Vols d'accessoires sur véhicules`,
               `Cambriolages de logement`,
               `Destructions et dégradations volontaires`,
               `Escroqueries et fraudes aux moyens de paiement`,
               `Usage de stupéfiants`,
               `Trafic de stupéfiants`)),
      na.rm = TRUE
    ),
    nonviolent_rate = nonviolent_total / insee_pop
  ) |>
  left_join(
    gent_yearly |> select(CODGEO, year, gent_index),
    by = c("CODGEO", "year")
  ) |>
  group_by(CODGEO) |>
  mutate(
    nonviolent_change = log(nonviolent_rate + 0.001) - log(dplyr::lag(nonviolent_rate) + 0.001),
    gent_index_lag1 = dplyr::lag(gent_index, 1),
    gent_index_lag2 = dplyr::lag(gent_index, 2)
  ) |>
  ungroup()

# Test
nv_lag1 <- nonviolent_panel |> filter(!is.na(nonviolent_change) & !is.na(gent_index_lag1))
cor_nv1 <- cor.test(nv_lag1$gent_index_lag1, nv_lag1$nonviolent_change)
print("=== NON-VIOLENT CRIMES - LAG 1 ===")
print(paste("Correlation:", round(cor_nv1$estimate, 4)))
print(paste("P-value:", round(cor_nv1$p.value, 4)))

nv_lag2 <- nonviolent_panel |> filter(!is.na(nonviolent_change) & !is.na(gent_index_lag2))
cor_nv2 <- cor.test(nv_lag2$gent_index_lag2, nv_lag2$nonviolent_change)
print("=== NON-VIOLENT CRIMES - LAG 2 ===")
print(paste("Correlation:", round(cor_nv2$estimate, 4)))
print(paste("P-value:", round(cor_nv2$p.value, 4)))
```

Visualizing all the lag effects of gentrification on violent crime :
```{r}
lag_summary <- data.frame(
  lag = c(1, 2, 3),
  correlation = c(cor_lag1$estimate, cor_lag2$estimate, cor_lag3$estimate),
  p_value = c(cor_lag1$p.value, cor_lag2$p.value, cor_lag3$p.value),
  n = c(nrow(lag1_data), nrow(lag2_data), nrow(lag3_data))
) |>
  mutate(significant = ifelse(p_value < 0.05, "Yes", "No"))

# Plot
ggplot(lag_summary, aes(x = factor(lag), y = correlation, fill = significant)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("No" = "gray", "Yes" = "red")) +
  labs(
    x = "Lag (Years)",
    y = "Correlation with Violent Crime Change",
    title = "Effect of Gentrification on Violent Crime at Different Lags",
    subtitle = "Testing if gentrification effects take time to materialize"
  ) +
  theme_minimal()
```

Including multiple lags in one model (for violent crimes)
```{r}
library(plm)

panel_multi_lag <- violent_with_lags |>
  filter(!is.na(violent_change) & !is.na(gent_index_lag1) & !is.na(gent_index_lag2))

if(nrow(panel_multi_lag) > 100) {
  panel_data <- pdata.frame(panel_multi_lag, index = c("CODGEO", "year"))
  
  model_lags <- plm(
    violent_change ~ gent_index_lag1 + gent_index_lag2 + gent_index_lag3,
    data = panel_data,
    model = "within",
    effect = "individual"
  )
  
  print(summary(model_lags))
}
```

Visualizing the effects of lag 1 and 2 gentrification on non violent crime over time :
```{r}
# Create summary data
lag_effects_nv <- data.frame(
  Lag = c("Lag 1\n(1 year)", "Lag 2\n(2 years)"),
  Correlation = c(0.1411, -0.09),
  P_value = c(0, 0.0003),
  Effect = c("Increase", "Decrease")
)

ggplot(lag_effects_nv, aes(x = Lag, y = Correlation, fill = Effect)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_text(aes(label = paste0("r = ", Correlation, "\np < 0.001")), 
            vjust = ifelse(lag_effects_nv$Correlation > 0, -0.5, 1.5),
            size = 4) +
  scale_fill_manual(values = c("Increase" = "#e74c3c", "Decrease" = "#27ae60")) +
  labs(
    x = "Time After Gentrification",
    y = "Correlation with Non-Violent Crime Change",
    title = "Gentrification Effects on Non-Violent Crime Over Time",
    subtitle = "Short-term increase followed by longer-term decrease"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
Comparison table of the effects of gentrification at lag 1 and 2 on violent and non violent crime
```{r}
# Summary comparison table
comparison <- data.frame(
  Crime_Type = c("Violent", "Violent", "Non-Violent", "Non-Violent"),
  Lag = c("1 year", "2 years", "1 year", "2 years"),
  Correlation = c(cor_lag1$estimate, cor_lag2$estimate, 0.1411, -0.09),
  Significant = c(
    ifelse(cor_lag1$p.value < 0.05, "Yes", "No"),
    ifelse(cor_lag2$p.value < 0.05, "Yes", "No"),
    "Yes", "Yes"
  )
)

knitr::kable(comparison, digits = 3, 
             col.names = c("Crime Type", "Lag", "Correlation", "Significant (p<0.05)"))
```
Our analysis reveals divergent temporal patterns: gentrification immediately reduces violent crime (r = -0.102) but increases non-violent crime (r = +0.141). By the second year, these relationships reverse, with violent crime rebounding (r = +0.143) while non-violent crime declines (r = -0.090). This suggests gentrification affects different crime types through distinct mechanisms—perhaps immediately deterring interpersonal violence through increased policing while initially creating new targets for property crime.

## Effects by crime type 
Testing the effects of lag gentrification by crime type (non violents)
```{r}
# List of crime types to test
crime_types <- c(
  "Violences physiques intrafamiliales",
  "Violences physiques hors cadre familial",
  "Vols avec armes",
  "Vols violents sans arme",
  "Vols sans violence contre des personnes",
  "Vols de véhicule",
  "Vols dans les véhicules",
  "Vols d'accessoires sur véhicules",
  "Cambriolages de logement",
  "Destructions et dégradations volontaires",
  "Escroqueries et fraudes aux moyens de paiement",
  "Usage de stupéfiants",
  "Trafic de stupéfiants"
)


# Function to test each crime type
test_crime_type <- function(crime_col) {
  crime_panel <- data_final |>
    filter(CODGEO %in% gentrifiable_areas) |>
    arrange(CODGEO, year) |>
    mutate(
      crime_rate = .data[[crime_col]] / insee_pop
    ) |>
    left_join(
      gent_yearly |> select(CODGEO, year, gent_index),
      by = c("CODGEO", "year")
    ) |>
    group_by(CODGEO) |>
    mutate(
      crime_change = log(crime_rate + 0.0001) - log(dplyr::lag(crime_rate) + 0.0001),
      gent_index_lag1 = dplyr::lag(gent_index, 1)
    ) |>
    ungroup() |>
    filter(!is.na(crime_change) & !is.na(gent_index_lag1) & is.finite(crime_change))
  
  if(nrow(crime_panel) > 30) {
    cor_result <- cor.test(crime_panel$gent_index_lag1, crime_panel$crime_change)
    return(data.frame(
      crime_type = crime_col,
      correlation = cor_result$estimate,
      p_value = cor_result$p.value,
      n = nrow(crime_panel),
      significant = ifelse(cor_result$p.value < 0.05, "Yes", "No")
    ))
  }
}

# Run for all crime types
crime_results <- bind_rows(lapply(crime_types, test_crime_type))

# Display results
crime_results |>
  arrange(p_value) |>
  knitr::kable(digits = 4, col.names = c("Crime Type", "Correlation", "P-value", "N", "Significant"))

# Visualize
ggplot(crime_results, aes(x = reorder(crime_type, correlation), y = correlation, fill = significant)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("No" = "gray70", "Yes" = "#e74c3c")) +
  labs(
    x = NULL,
    y = "Correlation with Crime Change (Lag 1)",
    title = "Gentrification Effects by Crime Type",
    subtitle = "Which crimes respond most to gentrification?"
  ) +
  theme_minimal()
```
-> Gentrification has only negative significant effects on "Vols sans violence contre des personnes", "Destructions et dégradations volontaires", "vols dans les véhicules", "violences physiques hors cadre familial", "vols violents sans armes", "cambriolages de logement", "vols d'accessoires sur véhicules", "escroqueries et fraudes aux moyens de paiement"

# Robustness checks

## Granger Causality

But maybe the causality is in the other direction ? does gentrification cause violent crime change or does violent crime causes gentrification ? Let's use Granger Causality
```{r}
library(plm)

# Preparing balanced panel
granger_data <- violent_crimes_panel |>
  filter(!is.na(violent_change) & !is.na(gent_index) & !is.na(gent_index_lag1)) |>
  group_by(CODGEO) |>
  filter(n() >= 3) |>  # Need at least 3 observations per area
  ungroup()

granger_panel <- pdata.frame(granger_data, index = c("CODGEO", "year"))

# Model 1: Crime change predicted by lagged gentrification
model_gent_causes_crime <- plm(
  violent_change ~ gent_index_lag1,
  data = granger_panel,
  model = "within",
  effect = "individual"
)

# Model 2: Reverse - does crime cause gentrification?
granger_data2 <- granger_data |>
  group_by(CODGEO) |>
  mutate(violent_change_lag1 = dplyr::lag(violent_change, 1)) |>
  ungroup() |>
  filter(!is.na(violent_change_lag1))

granger_panel2 <- pdata.frame(granger_data2, index = c("CODGEO", "year"))

model_crime_causes_gent <- plm(
  gent_index ~ violent_change_lag1,
  data = granger_panel2,
  model = "within",
  effect = "individual"
)

print("=== Does Gentrification Cause Crime? ===")
summary(model_gent_causes_crime)

print("=== Does Crime Cause Gentrification? ===")
summary(model_crime_causes_gent)
```
-> Granger causality tests reveal bidirectional relationships: gentrification 
significantly reduces subsequent violent crime (β = -0.036, p < 0.001), while 
violent crime reductions also predict increased gentrification (β = -0.137, 
p = 0.041). This suggests a self-reinforcing cycle where safer neighborhoods 
attract investment that further reduces crime.

Visualization 1: Reverse causality
```{r}
ggplot(granger_data2 |> filter(!is.na(violent_change_lag1) & !is.na(gent_index)), 
       aes(x = violent_change_lag1, y = gent_index)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Previous Year's Violent Crime Change",
    y = "This Year's Gentrification Index",
    title = "Crime Drives Gentrification Patterns",
    subtitle = "Areas with declining crime attract more gentrification (r = -0.19, p < 0.001)"
  ) +
  theme_minimal()
```

## Heterogeneity by city size

Comparing the effects of gentrification on  richer vs poorer areas :
```{r}
# Separating areas by income
interaction_data <- violent_crimes_panel |>
  left_join(baseline |> select(CODGEO, baseline_med), by = "CODGEO") |>
  mutate(
    income_tercile = ntile(baseline_med, 3),
    income_group = case_when(
      income_tercile == 1 ~ "Low Income",
      income_tercile == 2 ~ "Medium Income",
      income_tercile == 3 ~ "High Income"
    )
  ) |>
  filter(!is.na(violent_change) & !is.na(gent_index_lag1))

# Test by income group
interaction_results <- interaction_data |>
  group_by(income_group) |>
  summarise(
    correlation = cor(gent_index_lag1, violent_change, use = "complete.obs"),
    p_value = cor.test(gent_index_lag1, violent_change)$p.value,
    n = n()
  )

print(interaction_results)

# Regression with interaction
interaction_model <- lm(violent_change ~ gent_index_lag1 * income_group, 
                        data = interaction_data)
summary(interaction_model)
```
-> There is NO statistically significant difference in how gentrification affects violent crime across income groups.



Visualization 1: Interaction effect
```{r}
ggplot(interaction_data |> filter(!is.na(income_group)), 
       aes(x = gent_index_lag1, y = violent_change, color = income_group)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("Low Income" = "#e74c3c", 
                                 "Medium Income" = "#f39c12", 
                                 "High Income" = "#27ae60")) +
  labs(
    x = "Previous Year's Gentrification Index",
    y = "Violent Crime Change",
    color = "Initial Income Level",
    title = "Gentrification's Effect on Crime Varies by Initial Income",
    subtitle = "Gentrification REDUCES crime in low-income areas, but INCREASES it in high-income areas"
  ) +
  theme_minimal()
```
## Spatial spillover effects

Do neighboring areas' gentrification affect local crime?
```{r}
#Identifying department from CODGEO
spatial_data <- violent_crimes_panel |>
  mutate(department = substr(CODGEO, 1, 2)) |>
  group_by(department, year) |>
  mutate(
    # Average gentrification in the same department (excluding self)
    n_in_dept = n(),
    sum_gent_dept = sum(gent_index, na.rm = TRUE),
    neighbor_gent = (sum_gent_dept - gent_index) / (n_in_dept - 1)
  ) |>
  ungroup() |>
  group_by(CODGEO) |>
  mutate(
    neighbor_gent_lag1 = dplyr::lag(neighbor_gent, 1)
  ) |>
  ungroup() |>
  filter(!is.na(violent_change) & !is.na(neighbor_gent_lag1))

# Does neighboring gentrification affect local crime?
cor_neighbor <- cor.test(spatial_data$neighbor_gent_lag1, spatial_data$violent_change)
print("=== Neighboring Gentrification Effect ===")
print(paste("Correlation:", round(cor_neighbor$estimate, 4)))
print(paste("P-value:", round(cor_neighbor$p.value, 4)))

# Combined model
spatial_model <- lm(violent_change ~ gent_index_lag1 + neighbor_gent_lag1, 
                    data = spatial_data)
summary(spatial_model)
```

```{r}
# First, create total crime change variable
total_crime_panel <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  mutate(
    total_crime_rate = total_crimes / insee_pop
  ) |>
  left_join(
    gent_yearly |> select(CODGEO, year, gent_index),
    by = c("CODGEO", "year")
  ) |>
  group_by(CODGEO) |>
  mutate(
    total_crime_change = log(total_crime_rate + 0.0001) - 
                         log(dplyr::lag(total_crime_rate) + 0.0001),
    gent_index_lag1 = dplyr::lag(gent_index, 1)
  ) |>
  ungroup()

# Now run the same spatial analysis for total crimes
spatial_total <- total_crime_panel |>
  mutate(department = substr(CODGEO, 1, 2)) |>
  group_by(department, year) |>
  mutate(
    n_in_dept = n(),
    sum_gent_dept = sum(gent_index, na.rm = TRUE),
    neighbor_gent = (sum_gent_dept - gent_index) / (n_in_dept - 1)
  ) |>
  ungroup() |>
  group_by(CODGEO) |>
  mutate(neighbor_gent_lag1 = dplyr::lag(neighbor_gent, 1)) |>
  ungroup() |>
  filter(!is.na(total_crime_change) & !is.na(neighbor_gent_lag1))

# Correlation
cor_total_neighbor <- cor.test(spatial_total$neighbor_gent_lag1, 
                               spatial_total$total_crime_change)
print("=== TOTAL CRIMES: Neighboring Gentrification Effect ===")
print(paste("Correlation:", round(cor_total_neighbor$estimate, 4)))
print(paste("P-value:", round(cor_total_neighbor$p.value, 4)))

# Combined model
spatial_model_total <- lm(total_crime_change ~ gent_index_lag1 + neighbor_gent_lag1, 
                          data = spatial_total)
summary(spatial_model_total)
```
```{r}
# First calculate non-violent crimes
nonviolent_panel <- data_final |>
  filter(CODGEO %in% gentrifiable_areas) |>
  arrange(CODGEO, year) |>
  mutate(
    nonviolent_total = rowSums(
      across(c(
        `Vols sans violence contre des personnes`,
        `Vols de véhicule`,
        `Vols dans les véhicules`,
        `Vols d'accessoires sur véhicules`,
        `Cambriolages de logement`,
        `Destructions et dégradations volontaires`,
        `Escroqueries et fraudes aux moyens de paiement`,
        `Usage de stupéfiants`,
        `Trafic de stupéfiants`
      )),
      na.rm = TRUE
    ),
    nonviolent_rate = nonviolent_total / insee_pop
  ) |>
  left_join(
    gent_yearly |> select(CODGEO, year, gent_index),
    by = c("CODGEO", "year")
  ) |>
  group_by(CODGEO) |>
  mutate(
    nonviolent_change = log(nonviolent_rate + 0.0001) - 
                       log(dplyr::lag(nonviolent_rate) + 0.0001),
    gent_index_lag1 = dplyr::lag(gent_index, 1)
  ) |>
  ungroup()

# Spatial analysis for non-violent crimes
spatial_nonviolent <- nonviolent_panel |>
  mutate(department = substr(CODGEO, 1, 2)) |>
  group_by(department, year) |>
  mutate(
    n_in_dept = n(),
    sum_gent_dept = sum(gent_index, na.rm = TRUE),
    neighbor_gent = (sum_gent_dept - gent_index) / (n_in_dept - 1)
  ) |>
  ungroup() |>
  group_by(CODGEO) |>
  mutate(neighbor_gent_lag1 = dplyr::lag(neighbor_gent, 1)) |>
  ungroup() |>
  filter(!is.na(nonviolent_change) & !is.na(neighbor_gent_lag1))

# Correlation
cor_nonviolent_neighbor <- cor.test(spatial_nonviolent$neighbor_gent_lag1, 
                                    spatial_nonviolent$nonviolent_change)
print("=== NON-VIOLENT CRIMES: Neighboring Gentrification Effect ===")
print(paste("Correlation:", round(cor_nonviolent_neighbor$estimate, 4)))
print(paste("P-value:", round(cor_nonviolent_neighbor$p.value, 4)))

# Combined model
spatial_model_nonviolent <- lm(nonviolent_change ~ gent_index_lag1 + neighbor_gent_lag1, 
                               data = spatial_nonviolent)
summary(spatial_model_nonviolent)
```


-> Spillover effect for total crimes, not violent crimes and violent crimes : Cities surrounded by slowly-gentrifying (or non-gentrifying) neighbors tend to experience MORE crime growth.

## Heterogeneity by city size

Do we observe different effects depending on the size of the city ? (urban/rural comparison based on population)
```{r}
# Creating urban/rural indicator based on population
pop_data <- data_final |>
  group_by(CODGEO) |>
  summarise(mean_pop = mean(insee_pop, na.rm = TRUE))

urban_rural_data <- violent_crimes_panel |>
  left_join(pop_data, by = "CODGEO") |>
  mutate(
    area_type = case_when(
      mean_pop > 50000 ~ "Large City",
      mean_pop > 10000 ~ "Medium City",
      TRUE ~ "Small City"
    )
  ) |>
  filter(!is.na(violent_change) & !is.na(gent_index_lag1))

# Testing by area type
urban_rural_results <- urban_rural_data |>
  group_by(area_type) |>
  summarise(
    correlation = cor(gent_index_lag1, violent_change, use = "complete.obs"),
    p_value = cor.test(gent_index_lag1, violent_change)$p.value,
    n = n()
  )

print(urban_rural_results)

# Visualization
ggplot(urban_rural_data, aes(x = gent_index_lag1, y = violent_change, color = area_type)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~area_type) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x = "Previous Year's Gentrification Index",
    y = "Violent Crime Change",
    title = "Gentrification-Crime Relationship by City Size"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
In larger and medium cities there is a significantly negative correlation between the lagged gentrification and violent crime change rate, but not in small cities.

# Conclusion
## Summary of key findings
## Limitations

The main limitation of the analyse is the lack of data both in terms of year and city. Whole areas are missing, like Alsace or Moselle, as well as important cities at the arrondissement level (Marseille, Lyon). Moreover, using the Police data to study criminality can be biased. It can reflect the police activity or the number of police officers rather than the true number of crimes and offences. 

## Policy implications

## References

Ruth Glass (1964): https://www.tandfonline.com/doi/full/10.1080/03058034.2024.2425593 Neil Smith (1979): https://miguelangelmartinez.net/IMG/pdf/1979_smith_gentrification_japa.pdf