---
title: "Phase 2 - Data Loading and Cleaning"
author: "Noé MANSUY and Léo-Paul GARELLI"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

Github project: https://github.com/LEOPAUL55/Noe_Leo_final.git

```{r}
#| label: setup
#| include: false

here::i_am("Noe_Leo_Final.Rproj")
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(vroom)
library(data.table)
library(readr)
library(stringr)

# Suppress all messages globally
options(dplyr.summarise.inform = FALSE)
```

# Data Loading and Cleaning

## population 2018

Source : https://www.insee.fr/fr/statistiques/5650720

This is the 2018 population by IRIS code. Made by the INSEE.

Because the 2018 income data is available by IRIS code rather than by
GEO code, we need to weight each IRIS-level mean by its 2018 population
in order to compute the correct city-level values. Since the 2018
population data at the IRIS level is available for only 994 of the
cities included in the other years, we will have to restrict our
analysis to these 994 cities.

```{r}
#| message: false
#| warning: false

pop2018_IRIS <- read_delim(here("data","pop2018.csv"),
    na = "s") |>
  select(IRIS, P18_POP)
```

## Real estate prices

Source :
https://www.data.gouv.fr/datasets/indicateurs-immobiliers-par-commune-et-par-annee-prix-et-volumes-sur-la-periode-2014-2024

This is a collection of various variables on residential housing market
at the city level between 2014 and 2024. What is interesting for us is
the mean price by square meter of the goods that have been sold each
year. This was published by Boris Mericksay

Real‑estate price data are provided as separate yearly CSV files. To
improve efficiency and avoid redundant code, we use vroom’s multi‑file
loading functionality to read all files simultaneously. The year is
extracted from the file name, and city identifiers are harmonized to
allow consistent merging with other datasets.

```{r}
#| message: false
#| warning: false

ppsm <- vroom(
  list.files(
    here("data", "price per square meter"),
    pattern = "\\.csv$",
    full.names = TRUE
  ),
  col_select = c(INSEE_COM, Prixm2Moyen),
  id = "source",
  show_col_types = FALSE
) |>
  transmute(
    CODGEO = as.character(INSEE_COM),
    year   = as.integer(str_extract(source, "\\d{4}")),
    Prixm2Moyen
  )
```

We need to add the information for Paris by arrondissements that are
missing in the dataset above. To do so, we will use the data available
at https://paris.notaires.fr/fr/carte-des-prix?utm_source=chatgpt.com.

For simplicity, we only focus on the selling price of appartements in
the city from 2016 to 2021 (trimester 4). The available file is a PDF
that can be easily convert into a CSV file.

```{r}
#| label: load-paris
#| message: false
#| warning: false

ppsm_paris <- vroom(
  here("data", "dvfparis.csv"),
  delim = ",",
  quote = "\"",
  show_col_types = FALSE
) |>
  mutate(
    year    = as.integer(str_extract(Trimestre, "\\d{4}")),
    quarter = as.integer(str_extract(Trimestre, "(?<=T)\\d"))
  ) |>
  filter(quarter == 4, year >= 2016, year <= 2021) |>
  pivot_longer(
    cols = -c(Trimestre, quarter, year, Centre),
    names_to = "arrondissement",
    values_to = "Prixm2Moyen"
  ) |>
  transmute(
    CODGEO = as.character(75100 + as.integer(str_remove(arrondissement,
                                                        "er|e"))),
    year,
    Prixm2Moyen = as.numeric(str_remove_all(Prixm2Moyen, "[ €]"))
  ) |>
  arrange(year, CODGEO)
```

After that, we just need to merge the two datasets to obtain clean data
about the real estate price in France.

```{r}
ppsm <- bind_rows(ppsm, ppsm_paris) |>
  arrange(year, CODGEO)
```

## Poverty

sources :

-   2016, 2017, 2018 :
    https://www.data.gouv.fr/datasets/revenus-et-pauvrete-des-menages-aux-niveaux-national-et-local-revenus-localises-sociaux-et-fiscaux/

-   2019 : https://www.insee.fr/fr/statistiques/6036902?sommaire=6036904

-   2020 : https://www.insee.fr/fr/statistiques/6692392?sommaire=6692394

-   2021 : https://www.insee.fr/fr/statistiques/7756729?sommaire=7756859

These are collections of indicators of disposable income in France, at
the city level, between 2016 and 2021. These are made by the Insee.

We will only focus on the poverty rate, the median standard of living,
the first and the ninth deciles.

Poverty data are provided as separate yearly files with heterogeneous
structures. Variable names differ across years, and the 2018 data are
reported at the IRIS level rather than the commune level. To handle
this, we implement a loading strategy that reads each file separately
while restricting imports to the variables of interest (years
2016:2021). Poverty indicators are harmonized across years. Non‑numeric
values are explicitly converted to missing values during loading, but it
won't affect our analysis because this only concerns small cities. The
result is a consistent panel of indicators comparable across years.

```{r}
#| label: load-poverty
#| message: false
#| warning: false

poverty_files <- list.files(
  here("data", "poverty"),
  pattern = "\\.csv$",
  full.names = TRUE
)

load_poverty <- function(file) {
  year <- as.integer(str_extract(basename(file), "\\d{4}"))

  if (year %in% c(2016, 2017, 2019, 2020, 2021)) {
    out <- vroom(
      file,
      delim = ";",
      col_select = any_of(c(
        "CODGEO",
        "MED16", "MED17", "MED19", "MED20", "MED21",
        "TP6016", "TP6017", "TP6019", "TP6020", "TP6021",
        "D116", "D117", "D119", "D120", "D121",
        "D916", "D917", "D919", "D920", "D921"
      )),
      show_col_types = FALSE
    ) |>
      transmute(
        CODGEO,
        MED  = Reduce(coalesce, pick(any_of(c("MED16", "MED17", "MED19",
                                              "MED20", "MED21")))),
        TP60 = Reduce(coalesce, pick(any_of(c("TP6016", "TP6017", "TP6019",
                                              "TP6020", "TP6021")))),
        D1   = Reduce(coalesce, pick(any_of(c("D116", "D117", "D119", "D120", 
                                              "D121")))),
        D9   = Reduce(coalesce, pick(any_of(c("D916", "D917", "D919", "D920", 
                                              "D921")))),
        year
      )
  } else if (year == 2018) {
    out <- vroom(
      file,
      delim = ";",
      col_select = any_of(c(
        "IRIS",
        "DISP_MED18",
        "DISP_TP6018",
        "DISP_D118",
        "DISP_D918"
      )),
      show_col_types = FALSE
    ) |>
      transmute(
        IRIS,
        MED  = DISP_MED18,
        TP60 = DISP_TP6018,
        D1   = DISP_D118,
        D9   = DISP_D918,
        year
      )
  }

  out |>
    mutate(across(
      any_of(c("MED", "TP60", "D1", "D9")),
      ~ suppressWarnings(as.numeric(gsub(",", ".", .x)))
    ))
}

poverty <- bind_rows(lapply(poverty_files, load_poverty))

# Aggregate 2018 IRIS Data to City Level
poverty_2018 <- poverty |>
  filter(year == 2018) |>
  left_join(pop2018_IRIS, by = "IRIS") |>
  mutate(CODGEO = substr(IRIS, 1, 5)) |>
  group_by(CODGEO) |>
  summarise(
    MED = weighted.mean(MED, P18_POP, na.rm = TRUE),
    TP60 = weighted.mean(TP60, P18_POP, na.rm = TRUE),
    D1 = weighted.mean(D1, P18_POP, na.rm = TRUE),
    D9 = weighted.mean(D9, P18_POP, na.rm = TRUE),
    year = 2018,
    .groups = "drop"
  )

codgeo_2018 <- poverty_2018 |> pull(CODGEO)

poverty <- bind_rows(
  poverty |> filter(year != 2018),
  poverty_2018
) |>
  filter(CODGEO %in% codgeo_2018)
```



## Delinquency

Source:
https://www.data.gouv.fr/datasets/bases-statistiques-communale-departementale-et-regionale-de-la-delinquance-enregistree-par-la-police-et-la-gendarmerie-nationales

Here is the loading of the database for the delinquency. This is a
database of all the crimes and offences (except traffic offences) at the
city level in France, published by French Police Force, since 2016.

The CSV format of the file creates problems of fromat that must be
solved (types of column, delimiotation...).

The type of crimes must be reshape to long format (in rows) to wide
format (in columns).

The total number of crimes per year per city is computed by doing the
sum of each type of crimes.

```{r}
#| label: load-crime
#| message: false
#| warning: false

crime <- vroom(
  here("data_criminality.csv"),
  delim = ";",
  locale = locale(decimal_mark = ","),
  col_select = c(annee, CODGEO_2025, indicateur, nombre, insee_pop),
  col_types = cols(
    annee = col_integer(),
    nombre = col_integer(),
    insee_pop = col_integer()
  ),
  show_col_types = FALSE
) |>
  filter(!indicateur %in% c("Usage de stupéfiants (AFD)", 
                            "Violences sexuelles")) |>
  pivot_wider(
    names_from = indicateur,
    values_from = nombre,
    values_fill = 0
  ) |>
  rename(CODGEO = CODGEO_2025, year = annee) |>
  mutate(
    total_crimes = rowSums(across(-c(CODGEO, year, insee_pop)), na.rm = TRUE)
  )
```

## Selecting cities present in the 2018 data

The analysis will only focus on the cities where all the informations
are available on 2018, because of the IRIS nomenclature.

The period that we will study is from 2016 to 2021. We are adding in the
dataset the cities' name as well as removing the ones for which we don't
have the information of the real estate price in at least one year.

We also compute the crime rate.

```{r}
#| label: create-final-dataset
#| message: false
#| warning: false

city_names <- vroom(
  here("data", "poverty", "pov2016.csv"),
  delim = ";",
  col_select = c(CODGEO, LIBGEO),
  show_col_types = FALSE
) |>
  distinct()

data_final <- poverty |>
  left_join(crime, by = c("CODGEO", "year")) |>
  left_join(ppsm, by = c("CODGEO", "year")) |>
  select(-IRIS) |>
  left_join(city_names, by = "CODGEO") |>
  distinct(CODGEO, year, .keep_all = TRUE) |>
  # Calculate crime_rate first
  mutate(
    crime_rate = (total_crimes / insee_pop) * 100
  ) |>
  # Group and filter (OLD ORDER)
  group_by(CODGEO) |>
  filter(!is.na(Prixm2Moyen)) |>         # Filter NAs FIRST
  filter(!is.na(crime_rate)) |>           # Filter NA crime rates
  filter(n_distinct(year) == 6) |>        # THEN check for 6 years
  ungroup() |>
  # Population filter (NEW)
  filter(insee_pop >= 5000)
```

We obtained a clean dataset for `r n_distinct(data_final$CODGEO)` cities between 2016 and 2021.

# Cleaned data tables : summary and most important variables

After loading the datasets, we needed to convert the IRIS code into the
GEOCODE by weighting the poverty rate of each geocode by the population
of the corresponding IRIS code. By doing that, we obtained clean data
for `r n_distinct(data_final$CODGEO)` cities. We only focus on these cities in the analysis from 2016
to 2021. We also computed the crime rates for each city.

We only kept the variables of interest by loading the datasets, that are
the poverty rate, the first and last decile, the median income, the
number of crimes per crime type and the real estate price (average price
per square meter).Then we merged them into a single dataset that combines
all the data per cities per year.

```{r}
#| label: summary-table
#| echo: false

descriptions <- c(

"Unique city code (first 2 digits = department)",
"Median income",
"Poverty rate (< 60% of median income)",
"1st income decile",
"9th income decile",
"Year",
"City population",
"Domestic physical violence",
"Non-domestic physical violence",
"Armed robberies",
"Violent robberies (no weapon)",
"Non-violent robberies",
"Home invasions",
"Vehicle thefts",
"Thefts from vehicles",
"Thefts of vehicle accessories",
"Vandalism",
"Drug use",
"Drug dealing",
"Fraud",
"Total crimes",
"Price per m²",
"City name",
"Crime rate per 100 inhabitants"
)

summary_table <- data.table(
  Variable = names(data_final),
  Type = sapply(data_final, function(x) class(x)[1]),
  Description = descriptions
)

knitr::kable(summary_table, caption = "Variables in the final dataset")
```

```{r}
#| label: dataset-info
#| echo: false

dataset_info <- data.table(
  Metric = c("Number of observations", "Number of variables", 
             "Number of cities", "Time period"),
  Value = c(
    format(nrow(data_final), big.mark = ","),
    ncol(data_final),
    n_distinct(data_final$CODGEO),
    paste(min(data_final$year), "-", max(data_final$year))
  )
)

knitr::kable(dataset_info, caption = "Dataset overview")

```

# Research question :

Does the acceleration of the gentrification process in certain French
cities with more than 5000 inhabitants (2016-2021) correlate with a
decrease in recorded crime in these same municipalities, and is there a
displacement effect towards non-gentrified cities?

The first goal will be to make a gentrification index per city with a
few variables: square meter price growth, median wage growth, poverty
rate growth.

Then we will select the cities with the highest gentrification index
(say top 25%) and compare it to the rest of the cities : do we observe a
declination in terms of crime  between the top 25% and the others? Do we observe any linear relationship between crime rate growth and gentrification ?
If yes, do we observe a spatial displacement of crime to non-gentrifying
areas?

```{r}
#| label: filter-and-save-datasets
#| include: false 

# Get the exact CODGEO-year combinations from data_final
valid_keys <- data_final |> 
  select(CODGEO, year) |>
  distinct()

# Get unique cities for city_names
valid_codgeo <- unique(valid_keys$CODGEO)

# Filter all datasets to match data_final exactly
poverty_clean <- poverty |>
  semi_join(valid_keys, by = c("CODGEO", "year")) |>
  distinct(CODGEO, year, .keep_all = TRUE)

crime_clean <- crime |>
  semi_join(valid_keys, by = c("CODGEO", "year")) |>
  distinct(CODGEO, year, .keep_all = TRUE)

ppsm_clean <- ppsm |>
  semi_join(valid_keys, by = c("CODGEO", "year")) |>
  distinct(CODGEO, year, .keep_all = TRUE)

city_names_clean <- city_names |>
  filter(CODGEO %in% valid_codgeo) |>
  distinct(CODGEO, .keep_all = TRUE)

# Verify all have same number of rows
cat("Verification:\n")
cat("  - data_final:", nrow(data_final), "rows\n")
cat("  - poverty:", nrow(poverty_clean), "rows\n")
cat("  - crime:", nrow(crime_clean), "rows\n")
cat("  - ppsm:", nrow(ppsm_clean), "rows\n")
cat("  - city_names:", nrow(city_names_clean), "cities\n")

# Create output folder
if (!dir.exists(here("data", "cleaned"))) {
  dir.create(here("data", "cleaned"), recursive = TRUE)
}

# Save all filtered datasets
saveRDS(data_final, here("data", "cleaned", "data_final.rds"))
saveRDS(city_names_clean, here("data", "cleaned", "city_names.rds"))
saveRDS(poverty_clean, here("data", "cleaned", "poverty.rds"))
saveRDS(crime_clean, here("data", "cleaned", "crime.rds"))
saveRDS(ppsm_clean, here("data", "cleaned", "ppsm.rds"))
```






